---
title: "Bayesian Adaptive Design"
subtitle: >-
  Contextual Multi-Armed Bandits with Multivariate Bayesian
  Causal Forests
share:
  permalink: "https://book.martinez.fyi/bad.html"
  description: "Business Data Science: What Does it Mean to Be Data-Driven?"
  linkedin: true
  email: true
  mastodon: true
---

## The Business Dilemma: The Cancellation Desk

Imagine you are the Principal Data Scientist at a subscription-based
platform—whether a streaming service, an enterprise SaaS tool, or a
cloud developer product. Every week, roughly **500 paying
subscribers** contact customer support or enter the cancellation
flow intending to terminate their subscription.

Customer churn is the silent killer of recurring-revenue business
models. To combat this, your Customer Success and Growth teams have
brainstormed **four candidate actions (a control and three active
treatments)**:

1. **Arm 0: Control (Standard Exit)**: Standard cancellation survey
   and polite exit interview. No financial incentive or special
   intervention. Cost = **$0**.
2. **Arm 1: Bill Credit / Discount**: A $5 billing credit on their
   upcoming invoice. Highly effective for price-sensitive,
   long-tenure subscribers who have lower recent platform usage.
   Cost = **$5**.
3. **Arm 2: Premium Feature Upgrade**: A complimentary 2-month
   upgrade to the advanced feature tier. Highly effective for
   heavy-usage subscribers early in their customer lifecycle who are
   bumping against product limits. Cost = **$8**.
4. **Arm 3: Proactive Onboarding Assist**: A 1-on-1 technical setup
   and onboarding session. Highly effective for brand-new
   subscribers with low initial engagement who are struggling to
   adopt the product. Cost = **$2**.

Here is the central challenge: **The optimal retention offer is
completely different for different customers, but on Week 1, you
have zero empirical information about what works for whom.**

---

## The Economic Objective: Net ROI, Not Just Raw Retention

In academic causal inference, researchers often focus exclusively on
maximizing conversion or retention rates. But in business data
science, **interventions have real marginal costs**.

Suppose each successfully retained subscriber represents **$25 in
Net Customer Lifetime Value (LTV)**. If an expensive $8 feature
upgrade increases a customer's retention probability by only 10%,
its expected gross uplift value is:

$$
\mathbb{E}[\text{Uplift Value}] = 0.10 \times \$25 = \$2.50
$$

Because the treatment costs $8.00, deploying it to that customer
produces a **net loss of -$5.50 per customer treated**!

Therefore, the rational business objective is not merely to maximize
retention probability, but to maximize **Incremental Net Economic
Value**:

$$
\text{Net Value}_k(X) = \tau_k(X) \times \text{LTV} - \text{Cost}_k
$$

Where:

- $\tau_k(X) = \mathbb{E}[Y(k) - Y(0) \mid X]$ is the Conditional
  Average Treatment Effect (CATE) uplift in retention probability
  for arm $k$ relative to Control.
- $\text{LTV}$ is the monetary value of saving a customer ($25).
- $\text{Cost}_k$ is the direct operational cost of treatment $k$
  ($0, $5, $8, and $2, respectively).

Notice that for Arm 0 (Control), $\tau_0(X) \equiv 0$ and
$\text{Cost}_0 = \$0$, yielding a baseline net value of $\$0$. If
every active treatment produces a negative net value for a customer,
the cost-effective decision is to do nothing (Control).

---

## The Three-Way Strategic Showdown

When designing an experimentation strategy for this problem, data
science teams typically choose between three approaches:

```{mermaid}
flowchart TD
    Start["Candidate Experimentation Strategies"]
    Start --> RCT["1. Static RCT<br/>Uniform 1/K Allocation"]
    Start --> MAB["2. Global MAB<br/>Non-Contextual"]
    Start --> BCF["3. BCF Thompson Sampling<br/>Personalized, Contextual"]

    RCT --> RCT1["Equal 25% split"]
    RCT --> RCT2["Unbiased CATE estimates"]
    RCT --> RCT3["High regret, slow to react"]
    RCT --> RCT4["Massive revenue loss during the run"]

    MAB --> MAB1["Pools all users together"]
    MAB --> MAB2["Converges to one population winner"]
    MAB --> MAB3["Starves minority customer segments"]

    BCF --> BCF1["Models the CATE surface tau_k(X)"]
    BCF --> BCF2["Routes each user by posterior net ROI"]
    BCF --> BCF3["Balances exploration and exploitation"]
```

### 1. Static RCT Baseline (Uniform 25% Allocation)
In a traditional A/B/C/D test, you divide incoming traffic equally
(125 callers per arm each week) for a fixed period (e.g., 10 weeks).
While statistically straightforward, it comes at a crushing business
cost: you knowingly assign 75% of callers to suboptimal or
cost-ineffective offers for months, losing hundreds of salvageable
customers and tens of thousands in revenue.

### 2. Global (Non-Contextual) Multi-Armed Bandit
Product teams often ask: *"Why not just run an off-the-shelf
multi-armed bandit (e.g., Beta-Bernoulli Thompson Sampling)?"* When
treatment effects are heterogeneous, **a global bandit suffers from
a catastrophic winner-takes-all failure mode**. It pools all
customers together and converges to a single *population-average*
winner (e.g., Arm 1). In doing so, it completely starves minority
customer segments of the treatments they uniquely need: new users
never get Onboarding (Arm 3) and power users never get Feature
Upgrades (Arm 2), destroying value for over 60% of the customer
base.

### 3. Personalized Contextual Bandit via Bayesian Causal Forests (BCF)
By pairing response-adaptive randomization [@finucane2018works]
with **Multivariate Bayesian Causal Forests**
[@hahn2020bayesian; @stochastictree], we model the multi-treatment
CATE landscape
$\boldsymbol{\tau}(X) = (\tau_1(X), \tau_2(X), \tau_3(X))^\top$ over
continuous customer features $X$. For every incoming caller, we draw
from the BCF posterior distribution and route them to the treatment
with the highest sampled net economic return:

$$
A_i = \arg\max_{k \in \{0, 1, 2, 3\}} \left( \tilde{\tau}_k^{(s)}(X_i)
  \times \text{LTV} - \text{Cost}_k \right)
$$

---

## Why Multivariate BCF is Built for Adaptive Experimentation

As introduced by @hahn2020bayesian, Bayesian Causal Forests
decompose potential outcomes into:

$$
Y_i = \mu\left(X_i, \hat{\boldsymbol{\pi}}(X_i)\right) +
  \sum_{k=1}^K \tau_k(X_i) Z_{i, k} + \epsilon_i
$$

In a response-adaptive trial, assignment probabilities
$\boldsymbol{\pi}(X)$ change dynamically as the model learns. In
naive machine learning models (like regularized regression or
pooled random forests), this creates **Regularization-Induced
Confounding (RIC)**: shrinkage priors pull the treatment effect
toward the prognostic baseline because treatment assignment is
correlated with prognosis.

BCF solves this natively:

1. **Explicit Propensity Conditioning**: The prognostic forest
   $\mu(X, \hat{\boldsymbol{\pi}})$ conditions directly on the known
   design assignment probabilities $\hat{\boldsymbol{\pi}}$,
   insulating the treatment forest $\boldsymbol{\tau}(X)$ from
   confounding.
2. **Shared Tree Partitioning**: In StochTree's multivariate
   implementation (`stochtree`), the treatment forest outputs a
   vector of leaf effects $\boldsymbol{\tau}(X) \in \mathbb{R}^K$.
   All active arms share tree split structures, allowing treatments
   to "borrow strength" across one another while maintaining
   distinct leaf values.
3. **Exact Bayesian Thompson Sampling**: Instead of relying on
   artificial $\epsilon$-greedy heuristics, drawing samples from the
   BCF posterior ensemble naturally balances exploration and
   exploitation without artificial tuning — though in production we
   still blend in a small $\epsilon$-floor as a safety net (see
   Practical Lesson #2 below).

---

## Simulating the Continuous Customer Environment in R

Let's simulate a continuous customer feature space:

- $X_1 \in [0, 10]$: **Customer Tenure** (in months).
- $X_2 \in [0, 10]$: **Monthly Platform Usage** (in tens of hours).

We will run a 10-week simulation with 500 callers per week (5,000
total customers), comparing **BCF Thompson Sampling**, a **Global
Non-Contextual Bandit**, and a **Static RCT**.

### Step 1: Defining the Ground Truth Environment

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(stochtree)

set.seed(42)

# Economic constants
SAVED_CUSTOMER_LTV <- 25.0
ARM_COSTS <- c(0.0, 5.0, 8.0, 2.0)
ARM_NAMES <- c(
  "Arm 0: Control", "Arm 1: $5 Credit",
  "Arm 2: 2-Mo Upgrade", "Arm 3: Onboarding Assist"
)
ARM_COLORS <- c("#70757a", "#4285f4", "#ea4335", "#34a853")

# Minimum exploration probability enforced on every arm once the
# BCF model starts routing customers (Practical Lesson #2 below).
EPSILON_FLOOR <- 0.05

# Customer generator: continuous tenure X1 and usage X2
generate_customers <- function(n, week = 1) {
  tibble(
    week = week,
    customer_id = paste0("W", week, "_", seq_len(n)),
    tenure = runif(n, 0, 10),  # Months (0 to 10)
    usage = runif(n, 0, 10)    # Active hours / 10 (0 to 10)
  )
}

# Baseline retention probability under Control (Arm 0). Rises with
# tenure and usage; clipped to a plausible 15%-90% band for
# cancellation-desk callers (the raw logistic curve alone would
# reach as high as ~98% for the most engaged, longest-tenured
# callers).
true_baseline_retention <- function(df) {
  x1 <- df$tenure
  x2 <- df$usage
  logits <- -1.6 + 0.20 * x1 + 0.22 * x2 + 0.015 * (x1 * x2)
  pmin(pmax(1 / (1 + exp(-logits)), 0.15), 0.90)
}

# True CATE uplift in retention probability for Arms 1, 2, 3
true_cate_uplift <- function(df) {
  x1 <- df$tenure
  x2 <- df$usage
  n <- nrow(df)
  tau <- matrix(0, nrow = n, ncol = 4)

  # Arm 1 ($5 Credit): Peak at Tenure=7.5, Usage=2.5 (up to +36%)
  tau[, 2] <- 0.36 * exp(-0.06 * (x1 - 7.5)^2 - 0.08 * (x2 - 2.5)^2)

  # Arm 2 (Feature Upgrade, $8 Cost): Peak at Tenure=2.5, Usage=7.5
  # (up to +48%)
  tau[, 3] <- 0.48 * exp(-0.08 * (x1 - 2.5)^2 - 0.06 * (x2 - 7.5)^2)

  # Arm 3 (Onboarding Assist, $2 Cost): Peak at Tenure=2.0, Usage=2.0
  # (up to +32%)
  tau[, 4] <- 0.32 * exp(-0.08 * (x1 - 2.0)^2 - 0.08 * (x2 - 2.0)^2)

  return(tau)
}

# True net economic values: tau_k(X) * LTV - Cost_k
compute_true_net_values <- function(df) {
  tau <- true_cate_uplift(df)
  sweep(tau * SAVED_CUSTOMER_LTV, 2, ARM_COSTS, "-")
}

# True optimal arm per user
get_true_optimal_arms <- function(df) {
  net_vals <- compute_true_net_values(df)
  max.col(net_vals, ties.method = "random") - 1
}
```

### Visualizing the Ground Truth Optimal Decision Landscape

Let's visualize the ground truth optimal action across the
continuous 2D customer space:

```{r true_landscape_plot, fig.width=8, fig.height=5.5}
# Generate a dense 2D evaluation grid
grid_df <- expand.grid(
  tenure = seq(0, 10, length.out = 120),
  usage = seq(0, 10, length.out = 120)
)
grid_df$optimal_arm <- factor(
  get_true_optimal_arms(grid_df), levels = 0:3, labels = ARM_NAMES
)

ggplot(grid_df, aes(x = tenure, y = usage, fill = optimal_arm)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_manual(values = setNames(ARM_COLORS, ARM_NAMES)) +
  labs(
    title = "Ground Truth Optimal Action Landscape (Net ROI)",
    subtitle = paste(
      "Distinct customer regions require different retention",
      "interventions"
    ),
    x = "Customer Tenure (Months)",
    y = "Monthly Usage (x10 Hours)",
    fill = "Optimal Action"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

Notice that every single arm—including **Arm 0 (Control)**—has a
distinct territory:

- **Arm 0 (Control)** dominates in regions where no treatment
  produces enough uplift to overcome its cost.
- **Arm 1 ($5 Credit)** dominates for veteran, lower-usage users.
- **Arm 2 (Feature Upgrade)** dominates for newer power users.
- **Arm 3 (Onboarding Assist)** dominates for brand-new, low-usage
  users.

---

### Step 2: The Multi-Arm Adaptive Simulation Engine

```{r simulation_suite, message=FALSE, warning=FALSE}
simulate_experiment_suite <- function(n_weeks = 10, n_per_week = 500,
                                       warm_start_weeks = 1) {

  bcf_records <- list()
  global_records <- list()
  rct_records <- list()

  # Global bandit state (Beta-Bernoulli priors)
  global_alpha <- rep(1, 4)
  global_beta <- rep(1, 4)

  for (w in seq_len(n_weeks)) {
    cohort <- generate_customers(n_per_week, week = w)
    cohort$true_opt_arm <- get_true_optimal_arms(cohort)

    p_base <- true_baseline_retention(cohort)
    tau_mat <- true_cate_uplift(cohort)

    # -----------------------------------------------------------
    # Strategy 1: Static RCT Baseline (Uniform 25% Allocation)
    # -----------------------------------------------------------
    rct_cohort <- cohort
    rct_cohort$assigned_arm <- sample(0:3, n_per_week, replace = TRUE)
    rct_uplift <- tau_mat[
      cbind(seq_len(n_per_week), rct_cohort$assigned_arm + 1)
    ]
    rct_prob <- pmin(pmax(p_base + rct_uplift, 0), 1)
    rct_cohort$retained <- rbinom(n_per_week, 1, rct_prob)
    rct_cohort$cost <- ARM_COSTS[rct_cohort$assigned_arm + 1]
    rct_cohort$net_value <-
      rct_cohort$retained * SAVED_CUSTOMER_LTV - rct_cohort$cost
    rct_records[[w]] <- rct_cohort

    # -----------------------------------------------------------
    # Strategy 2: Global Non-Contextual Multi-Armed Bandit
    # -----------------------------------------------------------
    glb_cohort <- cohort
    # Draw retention rates from Beta posterior
    sampled_rates <- matrix(
      rbeta(
        4 * n_per_week,
        rep(global_alpha, each = n_per_week),
        rep(global_beta, each = n_per_week)
      ),
      nrow = n_per_week, ncol = 4
    )
    sampled_net <- sweep(
      sampled_rates * SAVED_CUSTOMER_LTV, 2, ARM_COSTS, "-"
    )
    glb_cohort$assigned_arm <-
      max.col(sampled_net, ties.method = "random") - 1

    glb_uplift <- tau_mat[
      cbind(seq_len(n_per_week), glb_cohort$assigned_arm + 1)
    ]
    glb_prob <- pmin(pmax(p_base + glb_uplift, 0), 1)
    glb_cohort$retained <- rbinom(n_per_week, 1, glb_prob)
    glb_cohort$cost <- ARM_COSTS[glb_cohort$assigned_arm + 1]
    glb_cohort$net_value <-
      glb_cohort$retained * SAVED_CUSTOMER_LTV - glb_cohort$cost

    # Update global Beta posteriors
    for (k in 0:3) {
      wins <- sum(glb_cohort$retained[glb_cohort$assigned_arm == k])
      trials <- sum(glb_cohort$assigned_arm == k)
      global_alpha[k + 1] <- global_alpha[k + 1] + wins
      global_beta[k + 1] <- global_beta[k + 1] + (trials - wins)
    }
    global_records[[w]] <- glb_cohort

    # -----------------------------------------------------------
    # Strategy 3: BCF Thompson Sampling (Personalized Contextual)
    # -----------------------------------------------------------
    bcf_cohort <- cohort

    if (w <= warm_start_weeks) {
      # Initial warm-start exploration phase: pure uniform draw
      bcf_cohort$assigned_arm <- sample(0:3, n_per_week, replace = TRUE)
      bcf_cohort$alloc_prob <- 0.25
    } else {
      # Ingest historical accumulated data
      past_bcf <- bind_rows(bcf_records)
      X_train <- as.matrix(past_bcf[, c("tenure", "usage")])

      # 3D treatment indicator basis Z in {0, 1}^3 (Arm 0 is the
      # reference level [0, 0, 0])
      Z_train <- matrix(0, nrow = nrow(past_bcf), ncol = 3)
      for (k in 1:3) {
        Z_train[past_bcf$assigned_arm == k, k] <- 1.0
      }

      y_train <- past_bcf$retained
      pi_train <- past_bcf$alloc_prob

      # Fit Multivariate BCF
      bcf_fit <- bcf(
        X_train = X_train,
        Z_train = Z_train,
        y_train = y_train,
        propensity_train = pi_train,
        num_gfr = 5,
        num_burnin = 80,
        num_mcmc = 50,
        general_params = list(num_threads = 2, random_seed = 42 + w)
      )

      # Predict posterior CATE draws for the incoming cohort
      X_new <- as.matrix(bcf_cohort[, c("tenure", "usage")])
      tau_draws <- predict(
        bcf_fit,
        X = X_new,
        Z = matrix(0, nrow = n_per_week, ncol = 3),
        propensity = rep(0.25, n_per_week),
        terms = "cate",
        type = "posterior"
      ) # Shape: [n_per_week, num_mcmc, 3]

      num_mcmc <- dim(tau_draws)[2]

      # Full 4-arm net-value draws (Arm 0 uplift is fixed at 0)
      tau_draws_all <- array(0, dim = c(n_per_week, num_mcmc, 4))
      tau_draws_all[, , 2:4] <- tau_draws
      net_draws <- sweep(
        tau_draws_all * SAVED_CUSTOMER_LTV, 3, ARM_COSTS, "-"
      )

      # Which arm wins the net value in each posterior draw, for
      # every customer: an [n_per_week, num_mcmc] matrix of arms.
      winner_by_draw <- apply(net_draws, c(1, 2), which.max) - 1

      # Exact per-arm Thompson-sampling win probability: the share
      # of the full posterior ensemble in which that arm is the
      # argmax, for each customer. There is no closed form for this
      # under Thompson sampling, so we compute it directly from the
      # posterior draws we already have.
      ts_win_prob <- t(apply(winner_by_draw, 1, function(draws) {
        tabulate(draws + 1, nbins = 4)
      })) / num_mcmc

      # Blend with an exploration floor so every arm keeps positive
      # support everywhere in covariate space (Practical Lesson #2).
      design_prob <-
        EPSILON_FLOOR / 4 + (1 - EPSILON_FLOOR) * ts_win_prob

      # Exact Thompson Sampling: draw one posterior sample per user,
      # with an epsilon chance of a uniform exploration draw instead.
      sampled_draw_idx <- sample(
        seq_len(num_mcmc), n_per_week, replace = TRUE
      )
      ts_choice <- winner_by_draw[
        cbind(seq_len(n_per_week), sampled_draw_idx)
      ]
      explore <- runif(n_per_week) < EPSILON_FLOOR
      random_choice <- sample(0:3, n_per_week, replace = TRUE)
      bcf_cohort$assigned_arm <-
        ifelse(explore, random_choice, ts_choice)

      # The known design propensity of the arm actually assigned -
      # exact by construction, so it eliminates Regularization-
      # Induced Confounding rather than merely approximating it.
      bcf_cohort$alloc_prob <- design_prob[
        cbind(seq_len(n_per_week), bcf_cohort$assigned_arm + 1)
      ]
    }

    bcf_uplift <- tau_mat[
      cbind(seq_len(n_per_week), bcf_cohort$assigned_arm + 1)
    ]
    bcf_prob <- pmin(pmax(p_base + bcf_uplift, 0), 1)
    bcf_cohort$retained <- rbinom(n_per_week, 1, bcf_prob)
    bcf_cohort$cost <- ARM_COSTS[bcf_cohort$assigned_arm + 1]
    bcf_cohort$net_value <-
      bcf_cohort$retained * SAVED_CUSTOMER_LTV - bcf_cohort$cost

    bcf_records[[w]] <- bcf_cohort
  }

  list(
    bcf = bind_rows(bcf_records),
    global = bind_rows(global_records),
    rct = bind_rows(rct_records)
  )
}

sim_results <- simulate_experiment_suite(n_weeks = 10, n_per_week = 500)
```

---

## Evaluating the Results

### 1. Learned Decision Boundaries vs. Ground Truth

How accurately did Multivariate BCF recover the optimal action
boundaries across the 2D customer space?

```{r decision_boundary_plot, fig.width=12, fig.height=5.5}
# Evaluate BCF predictions on the dense grid using the final fitted
# model
past_bcf <- sim_results$bcf
X_train_final <- as.matrix(past_bcf[, c("tenure", "usage")])
Z_train_final <- matrix(0, nrow = nrow(past_bcf), ncol = 3)
for (k in 1:3) Z_train_final[past_bcf$assigned_arm == k, k] <- 1.0

final_bcf <- bcf(
  X_train = X_train_final,
  Z_train = Z_train_final,
  y_train = past_bcf$retained,
  propensity_train = past_bcf$alloc_prob,
  num_gfr = 5,
  num_burnin = 100,
  num_mcmc = 50,
  general_params = list(num_threads = 2, random_seed = 999)
)

grid_X <- as.matrix(grid_df[, c("tenure", "usage")])
grid_pred <- predict(
  final_bcf,
  X = grid_X,
  Z = matrix(0, nrow = nrow(grid_X), ncol = 3),
  propensity = rep(0.25, nrow(grid_X)),
  terms = "cate",
  type = "mean"
)

# Predicted net values on grid
grid_pred_all <- cbind(0, grid_pred)
grid_pred_net <- sweep(
  grid_pred_all * SAVED_CUSTOMER_LTV, 2, ARM_COSTS, "-"
)
grid_df$bcf_optimal_arm <- factor(
  max.col(grid_pred_net, ties.method = "random") - 1,
  levels = 0:3, labels = ARM_NAMES
)

p1 <- ggplot(grid_df, aes(x = tenure, y = usage, fill = optimal_arm)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_manual(values = setNames(ARM_COLORS, ARM_NAMES)) +
  labs(
    title = "Ground Truth Optimal Landscape",
    x = "Tenure (Months)", y = "Usage (x10 Hours)", fill = "Arm"
  ) +
  theme_minimal() + theme(legend.position = "none")

p2 <- ggplot(
  grid_df, aes(x = tenure, y = usage, fill = bcf_optimal_arm)
) +
  geom_raster(interpolate = TRUE) +
  scale_fill_manual(values = setNames(ARM_COLORS, ARM_NAMES)) +
  labs(
    title = "BCF Learned Decision Landscape",
    x = "Tenure (Months)", y = "Usage (x10 Hours)", fill = "Arm"
  ) +
  theme_minimal() + theme(legend.position = "right")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

The side-by-side comparison demonstrates that **Multivariate BCF
successfully reconstructs the non-linear decision regions** without
any prespecified feature engineering or manual subgroup binning.

---

### 2. Dynamic Routing Accuracy Over Time

Did the algorithms learn to route customers to their personal
optimal intervention?

```{r routing_accuracy_plot, fig.width=9, fig.height=4.8}
# Calculate rolling accuracy to the true optimal arm. Each input
# data frame already holds one strategy's customers in chronological
# order, so row_number() gives a valid per-strategy customer index -
# essential so the three strategies overlay on the same x-axis below
# instead of being plotted as disjoint, back-to-back segments.
calculate_rolling_acc <- function(df, window = 250) {
  df %>%
    mutate(
      is_optimal = as.numeric(assigned_arm == true_opt_arm),
      customer_idx = row_number(),
      rolling_acc = zoo::rollmean(
        is_optimal, k = window, fill = NA, align = "right"
      )
    )
}

acc_df <- bind_rows(
  calculate_rolling_acc(sim_results$bcf) %>%
    mutate(strategy = "BCF Thompson Sampling (Personalized)"),
  calculate_rolling_acc(sim_results$global) %>%
    mutate(strategy = "Global MAB (Non-Contextual)"),
  calculate_rolling_acc(sim_results$rct) %>%
    mutate(strategy = "Static RCT Baseline (Uniform 25%)")
) %>%
  filter(!is.na(rolling_acc))

ggplot(
  acc_df, aes(x = customer_idx, y = rolling_acc, color = strategy)
) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.15, 0.95)
  ) +
  scale_color_manual(values = c(
    "BCF Thompson Sampling (Personalized)" = "#1a73e8",
    "Global MAB (Non-Contextual)" = "#f9ab00",
    "Static RCT Baseline (Uniform 25%)" = "#d93025"
  )) +
  labs(
    title = "Dynamic Routing Accuracy to Personal Optimal Arm Over Time",
    subtitle = paste(
      "BCF learns the multi-treatment landscape, climbing to",
      "~85% personalized accuracy"
    ),
    x = "Customer Sequence Within Strategy",
    y = "Proportion Assigned to True Optimal Arm",
    color = "Strategy"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

**Key Takeaways:**

- **Static RCT (Red)**: Stays flat at exactly 25% (random guessing).
- **Global Bandit (Orange)**: Converges to whichever single arm has
  the highest population-average performance, plateauing at only
  ~35–40% accuracy. It permanently fails the 60%+ of customers who
  belong to other segments.
- **BCF Thompson Sampling (Blue)**: Starts at 25% during warm-start
  exploration, then rapidly discovers the non-linear boundaries
  across Tenure and Usage, climbing to **~80–85%+ personalized
  accuracy**.

---

### 3. Posterior Uncertainty & 95% Credible Intervals Along Slices

To understand how well BCF quantifies treatment effect uncertainty,
let's examine posterior credible intervals along characteristic
slices through the continuous feature space:

```{r slices_plot, fig.width=11, fig.height=4.2}
N_test <- 30
eval_range <- seq(0.2, 9.8, length.out = N_test)

# Slice 1: Sweep Tenure at Usage = 2.5 (through peak of Arm 1)
X_slice1 <- data.frame(tenure = eval_range, usage = rep(2.5, N_test))
# Slice 2: Sweep Usage at Tenure = 2.5 (through peak of Arm 2)
X_slice2 <- data.frame(tenure = rep(2.5, N_test), usage = eval_range)
# Slice 3: Diagonal Sweep Tenure = Usage (through peak of Arm 3)
X_slice3 <- data.frame(tenure = eval_range, usage = eval_range)

# Extract posterior CATE draws from the final fitted BCF
predict_slice <- function(X_slice) {
  predict(
    final_bcf,
    X = as.matrix(X_slice),
    Z = matrix(0, N_test, 3),
    propensity = rep(0.25, N_test),
    terms = "cate",
    type = "posterior"
  )
}
draws_s1 <- predict_slice(X_slice1)
draws_s2 <- predict_slice(X_slice2)
draws_s3 <- predict_slice(X_slice3)

slice_summary <- bind_rows(
  tibble(
    slice = "Arm 1: $5 Credit (Usage = 2.5)",
    x_val = eval_range,
    true_tau = true_cate_uplift(X_slice1)[, 2],
    post_mean = apply(draws_s1[, , 1], 1, mean),
    ci_lower = apply(draws_s1[, , 1], 1, quantile, 0.025),
    ci_upper = apply(draws_s1[, , 1], 1, quantile, 0.975),
    arm_color = ARM_COLORS[2]
  ),
  tibble(
    slice = "Arm 2: 2-Mo Upgrade (Tenure = 2.5)",
    x_val = eval_range,
    true_tau = true_cate_uplift(X_slice2)[, 3],
    post_mean = apply(draws_s2[, , 2], 1, mean),
    ci_lower = apply(draws_s2[, , 2], 1, quantile, 0.025),
    ci_upper = apply(draws_s2[, , 2], 1, quantile, 0.975),
    arm_color = ARM_COLORS[3]
  ),
  tibble(
    slice = "Arm 3: Onboarding Assist (Diagonal)",
    x_val = eval_range,
    true_tau = true_cate_uplift(X_slice3)[, 4],
    post_mean = apply(draws_s3[, , 3], 1, mean),
    ci_lower = apply(draws_s3[, , 3], 1, quantile, 0.025),
    ci_upper = apply(draws_s3[, , 3], 1, quantile, 0.975),
    arm_color = ARM_COLORS[4]
  )
)

ggplot(slice_summary, aes(x = x_val)) +
  geom_line(
    aes(y = true_tau), linetype = "dashed", color = "#d93025",
    linewidth = 1
  ) +
  geom_point(aes(y = post_mean, color = slice), size = 2) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper, color = slice),
    width = 0.25, alpha = 0.7
  ) +
  facet_wrap(~ slice, scales = "free_x") +
  labs(
    title = paste(
      "BCF Posterior Credible Intervals vs. Ground Truth CATE",
      "Along Characteristic Slices"
    ),
    subtitle = paste(
      "Red dashed line: True Uplift. Points and bars: BCF",
      "Posterior Mean and 95% Credible Interval."
    ),
    x = "Feature Value",
    y = "Treatment Uplift in Retention (tau)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

The credible intervals accurately envelope the non-linear treatment
effect peaks. Where data density is high, the intervals shrink
tightly around the truth; where data is sparse, the credible
intervals naturally expand, providing the uncertainty signals
necessary for calibrated exploration.

---

### 4. Design Propensities Stay Bounded Away From Zero

Practical Lesson #3 below claims the design propensities are known
without error, and Practical Lesson #2 claims an exploration floor
keeps every arm's support positive. Both claims are only as good as
the numbers the simulation actually produced, so let's look at them
directly: the exact `alloc_prob` computed for every BCF-routed
customer once the model started making assignments (Week 2 onward).

```{r propensity_plot, fig.width=8, fig.height=4.5}
propensity_df <- sim_results$bcf %>%
  filter(week > 1) %>%
  mutate(
    assigned_arm_name = factor(
      assigned_arm, levels = 0:3, labels = ARM_NAMES
    )
  )

ggplot(propensity_df, aes(x = alloc_prob, fill = assigned_arm_name)) +
  geom_histogram(bins = 40, alpha = 0.85) +
  geom_vline(
    xintercept = EPSILON_FLOOR / 4, linetype = "dashed",
    color = "grey30"
  ) +
  scale_fill_manual(values = setNames(ARM_COLORS, ARM_NAMES)) +
  labs(
    title = "Exact Design Propensity of the Assigned Arm",
    subtitle = paste(
      "Dashed line marks the epsilon/4 exploration floor;",
      "no customer ever hits exactly zero"
    ),
    x = "P(this arm | X, adaptive design)",
    y = "Count",
    fill = "Assigned Arm"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Every bar sits to the right of the dashed exploration floor
($\epsilon / 4$): no customer-arm pair ever collapses to exactly
zero probability, so BCF's causal trees can always find
counterfactual observations to learn from. The spread above the
floor also shows the design at work — propensities climb toward 1
exactly where the posterior is confident an arm dominates, and stay
near the floor where the arms are still close.

---

### 5. Cumulative Net Business Value ($) Generated

```{r financial_plot, fig.width=9, fig.height=4.8}
financial_summary <- bind_rows(
  sim_results$bcf %>% mutate(strategy = "BCF Thompson Sampling"),
  sim_results$global %>% mutate(strategy = "Global Bandit"),
  sim_results$rct %>% mutate(strategy = "Static RCT Baseline")
) %>%
  group_by(strategy) %>%
  mutate(
    customer_idx = row_number(),
    cum_net = cumsum(net_value)
  )

total_bcf_val <- sum(sim_results$bcf$net_value)
total_glb_val <- sum(sim_results$global$net_value)
total_rct_val <- sum(sim_results$rct$net_value)

ggplot(
  financial_summary,
  aes(x = customer_idx, y = cum_net, color = strategy)
) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_color_manual(values = c(
    "BCF Thompson Sampling" = "#1a73e8",
    "Global Bandit" = "#f9ab00",
    "Static RCT Baseline" = "#d93025"
  )) +
  labs(
    title = paste(
      "Cumulative Net Economic Value ($) Generated During",
      "Experiment"
    ),
    subtitle = sprintf(
      paste(
        "BCF Thompson Sampling generates +$%.0f over Global",
        "Bandit and +$%.0f over Static RCT"
      ),
      total_bcf_val - total_glb_val, total_bcf_val - total_rct_val
    ),
    x = "Customer Timeline",
    y = "Cumulative Net Value ($)",
    color = "Strategy"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The economic lift of personalization is undeniable: by routing each
user to their cost-effective sweet spot, **BCF Thompson Sampling
generates tens of thousands in incremental profit during the trial
itself**.

---

## Practical Lessons for the Business Data Scientist

### 1. The Trap of the Non-Contextual Bandit
Product teams often ask: *"Why build a causal forest when an
off-the-shelf bandit is so simple?"* As shown above, when treatment
effects are heterogeneous, **a non-contextual bandit suffers from a
winner-takes-all failure mode**. It crowns a single majority arm and
starves everyone else. In customer retention, that means driving
high-churn minority segments straight out the door.

### 2. Always Include an Exploration Floor ($\epsilon$)
Pure greedy Thompson sampling can prematurely collapse allocation to
zero for an arm that experiences an unlucky run of cancellations
early in Week 1. Enforcing an exploration floor — the simulation
above uses `EPSILON_FLOOR <- 0.05` — guarantees **positivity
(overlap)** across covariate space, ensuring your causal trees can
always observe counterfactuals if underlying customer behavior
shifts. The propensity histogram above shows this floor holding in
practice: no customer-arm pair ever hits exactly zero probability.

### 3. Propensity Scores are Known Without Error (If You Compute Them Right)
In observational studies, propensity scores $\hat{\pi}(X)$ must be
estimated with noise. In our adaptive trial, **you designed the
assignment mechanism**, so the propensities are knowable exactly —
but for Thompson sampling that doesn't make them *trivial*. There is
no closed form for "the probability arm $k$ wins the posterior
argmax for this customer," so the simulation computes it directly:
for every posterior draw, tally which arm has the highest net value,
then average across the full MCMC ensemble (blended with the
$\epsilon$-floor). Passing this exact per-customer, per-arm design
propensity into StochTree's `propensity_train` — rather than a rough
placeholder — is what actually eliminates Regularization-Induced
Confounding by construction.

### 4. Handling Delayed Feedback Loops
In our case study, retention was observable immediately at the
conclusion of the call. If churn takes 30 or 60 days to confirm, you
can:

- Adapt every 30 days instead of weekly.
- Impute full retention using short-term **surrogate metrics**
  (e.g., product logins in the 7 days following the cancellation
  call) using the techniques covered in Chapter 7 (*Surrogates*).

### 5. Transitioning to Production
At the end of the trial, you don't just produce a static PowerPoint
report declaring a single "winner." You extract the fitted
multivariate BCF model, save it (`stochtree` JSON serialization),
and deploy the learned decision policy:

$$
z^*(X) = \arg\max_{k \in \{0, 1, 2, 3\}} \left( \hat{\tau}_k(X)
  \times \text{LTV} - \text{Cost}_k \right)
$$

into your CRM or customer support software, delivering personalized
retention offers to every future customer in real time.

---

## Conclusion

Bayesian Adaptive Design powered by Multivariate BCF bridges the gap
between **rigorous causal inference** and **modern algorithmic
decision-making**. It allows business data scientists to replace
static, high-regret A/B tests with dynamic learning systems that
protect customer relationships and maximize company revenue from the
very first week.

::: {.callout-tip}
## Learn more
- @finucane2018works: *What works for whom? A Bayesian approach to
  channeling big data streams for public program evaluation.*
  American Journal of Evaluation.
- @hahn2020bayesian: *Bayesian regression tree models for causal
  inference: Regularization, confounding, and heterogeneous
  effects.* Bayesian Analysis.
- @stochastictree: *StochTree: High-performance stochastic tree
  ensembles for causal inference and machine learning.*
:::
