---
title: "Bayesian Synthetic Control"
description: "Synthetic control builds a weighted combination of untreated units to stand in for the treated one. What the Bayesian formulation adds in honest uncertainty quantification."
share:
  permalink: "https://book.martinez.fyi/bsynth.html"
  description: "Bayesian synthetic control: building a counterfactual from a weighted donor pool"
  linkedin: true
  email: true
  mastodon: true
---

Synthetic control methods have become a widely applied tool for empirical
researchers to estimate the effect of interventions or treatments, especially
when traditional randomized controlled trials aren't feasible. In a recent
Journal of Economic Perspectives survey on the econometrics of policy
evaluation, Susan Athey and Guido Imbens describe synthetic controls as
"arguably the most important innovation in the policy evaluation literature in
the last 15 years" [@athey2017state]. The technique involves creating a
"synthetic" version of the treated unit by weighting untreated units from a
donor pool. This essentially allows us to estimate what would have happened if
the treatment had never occurred.

## Key Concepts and Principles


Before we dive into the Bayesian approach, let's review some fundamental
concepts:

**Pre-treatment Fit:** The credibility of a synthetic control estimator hinges
on how well it can track the trajectory of the outcome variable for the treated
unit before the intervention. A close pre-treatment fit makes for more reliable
post-treatment estimates.

**Convex Hull Condition:** The synthetic control method works best when the
characteristics of the treated unit fall within the convex hull of the donor
pool units' characteristics. This ensures that the treated unit can be
approximated by a weighted average of donor units.

**Sparse Solutions:** Synthetic control estimates typically involve only a few
donor pool units with non-zero weights. This sparsity aids in interpretability
and helps reduce overfitting.

**No Anticipation:** The method assumes that there are no anticipation effects
before the intervention. If such effects exist, it's advisable to backdate the
intervention in the dataset.

**Sufficient Pre- and Post-intervention Information:** The credibility of the
estimates depends on having enough pre-intervention periods to establish a good
fit and enough post-intervention periods to observe the full effect of the
intervention.

**No Interference:** The method assumes that the intervention does not affect
the outcomes of the untreated units. This assumption should be carefully
considered in the study design.

## The Bayesian Advantage

**Prior Information:** Bayesian methods allow us to incorporate prior knowledge
or beliefs about the data. This can be particularly useful when we have relevant
information from past studies or expert opinions.

**Posterior Distribution:** By combining the prior distribution with the
likelihood of the observed data, we get a posterior distribution. This
distribution represents our updated beliefs about the parameters after taking
into account the new data.

**Uncertainty Quantification:** One of the key strengths of Bayesian methods is
their ability to quantify uncertainty. The posterior distribution gives us a
range of plausible values for the treatment effects, along with associated
probabilities.

**Hierarchical Models:** Bayesian synthetic control models can be built with
hierarchical structures. This allows for more complex relationships and
dependencies within the data.

### Mathematical Formulation

In the Bayesian approach, we typically use a Dirichlet distribution as the prior
for the weights, ensuring they are positive and sum to 1. We can also introduce
a scaling matrix, often denoted as Γ, to control the importance of different
predictors.

Let's formalize this with some notation:

- $X_1$: A $k \times 1$ matrix of predictors for the treated unit.
- $X_0$: A $k \times J$ matrix of predictors for the donor units.
- $w$: A $J\times 1$ vector of weights for the synthetic control.
- $\sigma$: A scaling parameter.
- $\Gamma$ A $k \times k$ scaling matrix.

A simple Bayesian synthetic control model can be formulated as:


$$ 
\begin{aligned}
X_1 | w, \sigma &\sim N(X_0w , \text{diag}(\Gamma)^{-2}\sigma^2) \\
w &\sim \text{Dir}(1)\\
\sigma &\sim N^+(0,1)\\
\Gamma &\sim Dir((v_1, \dots, v_k)') \quad \text{s.t. } 1'v = 1 \\
\end{aligned}
$$


### Practical Implementation: The German Re-unification Example

In 1989, a monumental event occurred: the reunification of East and West
Germany. A natural question for policymakers was: "What impact did reunification
have on West Germany's GDP?"

This very question was addressed in one of the seminal papers on synthetic
control [see @abadie2015comparative]. Using a Bayesian approach, we can not only
estimate the effect of reunification but also quantify the uncertainty around
that estimate.

The {bsynth} package in R provides a convenient way to apply Bayesian synthetic
control methods. Let's see how we can analyze the German reunification data:

```{r germany}
library("bsynth")
load("germany.rda")
germany_synth <- bayesianSynth$new(data = germany,
                                   time = year,
                                   id = country,
                                   treated = D,
                                   outcome = gdp,
                                   ci_width = 0.95,
                                   predictor_match = FALSE)

germany_synth$timeTiles + ggplot2::xlab("Year") + ggplot2::ylab("Country")
```
In this example, we're starting with a simple model that doesn't include predictor matching. We'll fit the model and visualize the results:

```{r fit, message=FALSE, results = "hide"}
germany_synth$fit(cores = 4)

# Vizualize the Bayesian Synthetic Control
germany_synth$synthetic + 
  ggplot2::xlab("Year") +
  ggplot2::ylab("Per Capita GDP (PPP, 2002 USD)") +
  ggplot2::scale_y_continuous(labels=scales::dollar_format())

```
::: {.content-visible when-format="html"}
We can also examine the estimated lift (the cumulative effect of the treatment) over a specific time period:

```{r liftDraws}
#| eval: !expr knitr::is_html_output()
germany_synth$liftDraws(from = lubridate::as_date("1990-01-01"), 
                        to = lubridate::as_date("2002-01-01"))
```
:::

### When Things Go Wrong: The Pitfalls of Synthetic Controls

It's crucial to remember that synthetic control isn't a magic bullet. Things can
go awry, and you could end up with estimates that are entirely off the mark.
Here are some common pitfalls to watch out for:

  - **Poor Pre-treatment Fit:** If your synthetic control doesn't accurately
    replicate the treated unit's pre-treatment behavior, don't use it. It's as
    simple as that.

  - **Overfitting:** Even with a perfect pre-treatment fit, there's the danger
    of overfitting. This is more likely to happen if you have a short
    pre-treatment period, a large donor pool, noisy data, or if you relax the
    weight constraints and allow for extrapolation.


**Be careful** when using synthetic controls, things co go bad and you could end
up with an estimate that is the wrong sign!! The weight restriction allows us
to cleanly characterize an upper bound for the bias:


\begin{align*}
E[|\hat{\tau}_{1t} - \tau_{1t}|] \lesssim \underbrace{C_1\mathbb{E}\text{MAD}\left(Y_1^P, \hat{Y}_j^P\right) + k C_2 \mathbb{E}\text{MAD}\left(Z_1^1,\hat{Z}_j^1\right)}_{\text{First Order}} + \underbrace{C_3 J^{1/3} \frac{\bar{\sigma}}{T_0^{1/2}}}_{\text{Second Order}}
\end{align*}

1.  **Fit matters most**: If the synthetic control can not replicate the treated
    unit over time, you should **not** use it.

2.  **Don't chase noise**: Even with perfect pre-treatment fit there is the
    danger that you are **over-fitting** to the pre-treatment period.

Over-fitting is more likely in the following situations:

  - You have a short pre-treatment period (small $T_0$).
  - You have a large donor pool (large $J$) or the units are not similar to your
    treated unit.
  - You have very noisy data.
  - You allow for extrapolation by relaxing the weight constraints. In this
    case, you might have perfect pre-treatment fit but you will likely have
    significant bias from over-fitting.

### Check the Bias of your Bayesian Synthetic Controls

The 'bsynth' package offers you a nice and easy way to check how likely it is
that your estimate is badly biased! By computing an upper bound on the relative
bias we get an estimate of the probability that your effect could change signs
because of the bias.


In the case of the German re-unification this is unlikely when we consider the
full post-treatment period of 12 years.

::: {.content-visible when-format="html"}
```{r bias1, warning=FALSE}
#| eval: !expr knitr::is_html_output()
germany_synth$biasDraws(small_bias = 0.2, 
                        firstT = lubridate::as_date("1990-01-01"), 
                        lastT = lubridate::as_date("2002-01-01"))

```
:::

However, for a smaller time frame of just 5 years after the re-unification, the
bias could overturn the effect! Be careful when you choose a time period to
measure cumulative effects as it will change the relative bias too.

::: {.content-visible when-format="html"}
```{r bias2, warning=FALSE}
#| eval: !expr knitr::is_html_output()
germany_synth$biasDraws(small_bias = 0.2, 
                        firstT = lubridate::as_date("1990-01-01"), 
                        lastT = lubridate::as_date("1994-01-01"))
```
:::

::: {.callout-tip}
## Learn more
  - @abadie2021using Using Synthetic Controls: Feasibility, Data Requirements,
    and Methodological Aspects.
  - @abadie2022synthetic Synthetic Controls in Action.
  - @martinez2023bayesian Bayesian and Frequentist Inference for Synthetic
    Controls.
:::
