---
title: "LongBet: Time-Varying Treatment Effects in Panel Data"
description: "LongBet extends Bayesian causal forests to panel data. Applied to a randomized staggered rollout, it recovers how a treatment effect evolves week by week, for whom, and where it is heading after the experiment ends."
share:
  permalink: "https://book.martinez.fyi/longbet.html"
  description: "LongBet: heterogeneous, time-varying treatment effects in panel data"
  linkedin: true
  email: true
  mastodon: true
---

## Introduction to LongBet

The models in the previous two chapters answer a question with no time
dimension: what is the effect of the treatment on the outcome? In a business,
that is rarely the whole question. A feature that lifts revenue by 10% in its
first week and fades to nothing by its tenth averages about the same over a
ten-week experiment as one that lifts revenue by 5% and holds there. They
report the same "average treatment effect." They are worth completely different
amounts a year later, and they lead to opposite decisions.

LongBet, introduced by @wang2024longbet, extends the Bayesian Causal Forest
(BCF) model of @hahn2020bayesian to panel data so that the treatment effect
becomes a function of two things at once: **who** the unit is, and **how long
it has been treated**. It was designed for short panels with large
cross-sections and observed confounders --- the shape of data that
e-commerce, subscription, and marketplace businesses generate by default.

LongBet models the data generating process as follows:

$$
Y_{it} = \alpha\mu(X_i, T=t) + \beta_S\nu(X_i, S_{it}, T=t) + \epsilon_{it}
$$

where:

  - $Y_{it}$ is the outcome for unit $i$ at time $t$
  - $X_i$ are time-invariant, pre-treatment covariates
  - $T = t$ is calendar time, shared by every unit
  - $S_{it} = \sum_{s \le t} Z_{is}$ is the time elapsed since unit $i$ adopted
    the treatment ($S_{it} = 0$ for a unit that has not adopted yet)
  - $\mu(\cdot)$ is the prognostic function: what the outcome would have done
    anyway
  - $\nu(\cdot)$ is the treatment effect function
  - $\beta_S$ is a Gaussian process factor over time-since-adoption, capturing
    the trend in treatment effect shared by all units
  - $\epsilon_{it}$ is an independent Gaussian error term

Both $\mu(\cdot)$ and $\nu(\cdot)$ are XBART forests [@he2019xbart] that are
allowed to split on the time dimensions as well as on $X$. The treatment effect
estimator is the contrast between being $S$ periods into treatment and not
being treated at all:

$$
\tau_t(X_i, S) = \beta_S\,\nu(X_i, S_{it}, T=t) - \beta_0\,\nu(X_i, 0, T=t).
$$

Two pieces of that expression do different jobs, and it is worth separating
them. $\beta_S$ is **one** curve over time-since-adoption, shared by the whole
population: it is where the model pools information about the average shape of
the response. $\nu(X_i, S_{it}, t)$ is a forest, so it can make that shape
differ across units --- and because it splits on $S$ too, different kinds of
units can have genuinely different *shapes*, not just different magnitudes of
the same shape. The shared $\beta_S$ is also what makes forecasting possible:
because it is a Gaussian process over $S$, it can be extrapolated past the
largest $S$ in the data.

## Key Features of LongBet

1.  **Time-varying effects.** The effect is estimated as a trajectory over
    time-since-adoption, not a single number.
2.  **Heterogeneity in the trajectory.** Because $\nu$ splits on $X$ and $S$
    jointly, LongBet can find subpopulations whose effects decay while others'
    compound.
3.  **Staggered adoption.** Units can be treated at different times; once
    treated they stay treated. Effects are aligned on $S$, not on calendar
    time.
4.  **Separate regularization.** As in BCF, prognostic and treatment surfaces
    get separate forests and separate priors, so the model can be flexible
    about the baseline while staying conservative about the effect.
5.  **No parallel trends requirement.** Identification comes from conditioning
    on $X_i$, not from assuming treated and untreated units would have moved
    in parallel.
6.  **Forecasting.** The Gaussian process on $\beta_S$ projects the effect
    beyond the end of the observation window.
7.  **Bayesian uncertainty.** Every quantity above --- including the effect for
    a single unit at a single time --- comes with a posterior.

## What a Panel Model Buys You in a Randomized Experiment {#sec-longbet-why}

LongBet's headline claim in @wang2024longbet is about observational panels: it
does not need the parallel trends assumption that difference-in-differences
rests on [@callaway2021did; @roth2023trending]. This chapter deliberately
applies it somewhere else --- to data from a **randomized experiment** --- and
it is worth being precise about why, because the argument changes.

In a randomized rollout you already have identification. Random assignment
makes treated and untreated units exchangeable, so parallel trends holds in
expectation and a simple difference-in-differences against a holdout is
unbiased for the average effect. You do not need a model to believe the number.

What you do not get for free is everything else:

  - **Precision.** A marketplace's weekly revenue per seller varies by orders
    of magnitude across sellers and swings seasonally within them. A model of
    $\mu(X_i, t)$ explains that variation away before the effect is read off.
    This is the non-parametric cousin of ANCOVA and of CUPED
    [@deng2013cuped], and it is the difference between an experiment that
    answers your question and one that runs for another quarter.
  - **Resolution.** A difference-in-differences gives you one number per event
    time. It cannot tell you which sellers those numbers came from unless you
    pre-specify the subgroups, and every subgroup you add costs you power.
    LongBet returns a posterior for $\tau_i(S)$ for *every* unit at *every*
    time-since-adoption, from a single fit.
  - **Reach.** Experiments end; business cases run for a year. The Gaussian
    process on $\beta_S$ extrapolates the trajectory past the last week you
    observed, with intervals that widen as it goes.

Keep the division of labour clear. **Randomization is what makes the estimate
credible; the model is what makes it useful.** If LongBet's average effect and
a simple difference-in-differences against the holdout disagree materially, the
simple estimator is the one to trust for the average, and the disagreement is a
signal to go debug the model. What the simple estimator *cannot* do is tell you
which sellers to ship to.

::: {.callout-note title="Assumptions still required"}
Randomization buys unconfoundedness and overlap, but not SUTVA. If sellers
compete for the same buyers, one seller's better listings can cannibalize
another's sales, and the experiment measures the *relative* advantage of being
treated rather than the effect of treating everyone. That is a design problem,
not a modelling one --- see @sec-longbet-limits.
:::

## The Business Problem: A Capacity-Constrained Rollout

You are the data scientist at an online marketplace. The product team has
built an **AI listing optimizer**: it rewrites seller titles and descriptions,
suggests attributes, and re-crops photos. Early qualitative feedback is
enthusiastic. Leadership wants to know whether to make it permanent, and
finance wants a number to put in next year's plan.

There are 3,000 sellers in the pilot population and the optimizer cannot simply
be switched on for all of them: each seller needs a supervised migration of
their catalog, and the onboarding team can absorb about 450 sellers a week.
The rollout is going to be phased whether you like it or not.

That constraint is an opportunity. The field team's instinct is to start with
the sellers most likely to benefit. Doing so would make the rollout order a
function of expected outcomes --- exactly the targeted selection that makes
observational panel data hard to analyze. Instead, you randomize the order:
sellers are assigned at random (stratified on vertical, catalog size, and
fulfillment status) to one of four weekly launch waves or to a **holdout** that
stays on the old experience for the duration of the study. The result is a
**randomized staggered rollout** --- a real experiment that happens to produce
panel data with four independent replications of the launch.

Three questions have to be answered before the decision can be made:

1.  **Does the lift last?** A rewritten catalog could produce a one-off
    ranking bump that fades, or a compounding advantage as better listings earn
    better placement.
2.  **Who does it help?** If the effect is concentrated, the right decision is
    to ship to a segment, not to everyone.
3.  **What is it worth over a year?** The study runs 14 weeks after the first
    launch. The business case runs for 52.

The economics are simple enough to state up front. The marketplace keeps a
**12% take rate** on gross merchandise value (GMV), and running the optimizer
costs about **\$500 per seller per year** in inference, image processing, and
support. A seller only earns their keep if the lift in their GMV, valued at the
take rate, clears that cost --- and because the cost is flat per seller while
the benefit scales with the seller's GMV, the bar is much easier for a large
seller to clear than for a small one. Hold on to that asymmetry; it turns out
to matter more than the treatment effect itself.

## Simulating the Rollout

As elsewhere in this book we simulate the data so the truth is known and every
claim can be checked against it. The outcome is **log weekly GMV**, which
keeps a heavy-tailed dollar quantity roughly Gaussian and makes the treatment
effect read as an approximate percentage lift.

The seller population carries five pre-treatment characteristics: the vertical
they sell in, whether they use marketplace fulfillment, how many active
listings they carry, and two summaries of their own recent history. The
untreated trajectory is deliberately **non-parallel**: apparel sellers swing
seasonally, electronics sellers ramp into the holidays, and larger sellers
drift upward faster.

```{r setup}
#| message: false
#| warning: false
library(longbet)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(purrr)
library(knitr)

set.seed(1982)

PAL <- c(LongBet = "#0072B2", `DiD vs holdout` = "#D55E00",
         Truth = "#000000", `Broad catalog` = "#009E73",
         `Narrow catalog` = "#CC79A7")
```

The calendar has three parts. Weeks 1--6 are a **lookback window** used only
to build covariates. Weeks 7--24 are the **study panel** handed to the model:
everyone is untreated through week 10, and the four waves launch in weeks
11--14. Weeks 25--30 are simulated but **withheld**, and exist only so we can
score the forecast later.

```{r calendar}
n           <- 3000
week_base   <- 1:6      # lookback: builds covariates, never modelled
week_study  <- 7:24     # the panel LongBet sees
week_future <- 25:30    # withheld; used only to score the forecast
weeks_all   <- c(week_base, week_study, week_future)

LAUNCH <- c(W1 = 11, W2 = 12, W3 = 13, W4 = 14, Holdout = Inf)
```

### Sellers and the randomized rollout

```{r sellers}
vertical  <- sample(c("Apparel", "Home", "Electronics"), n,
                    replace = TRUE, prob = c(0.40, 0.35, 0.25))
fulfilled <- rbinom(n, 1, 0.45)
listings  <- round(exp(rnorm(n, log(35), 0.9)))   # active listings per seller
seller_sd <- rnorm(n)                             # latent seller scale

# Blocked randomization: within each stratum, deal out the five arms in
# fixed proportions and shuffle. This is stratified randomization applied to
# launch *order* rather than to treatment status.
strata <- interaction(
  vertical,
  cut(listings, quantile(listings, 0:3 / 3), include.lowest = TRUE),
  fulfilled, drop = TRUE
)
deal_waves <- function(k) {
  arms  <- c("W1", "W2", "W3", "W4", "Holdout")
  props <- c(0.15, 0.15, 0.15, 0.15, 0.40)
  sample(rep(arms, diff(round(k * cumsum(c(0, props))))))
}
wave <- character(n)
for (s in levels(strata)) {
  idx <- which(strata == s)
  wave[idx] <- deal_waves(length(idx))
}
launch <- LAUNCH[wave]

table(wave)
```

### The untreated trajectory

```{r potential-outcomes}
# Seller level: size, catalog breadth, fulfillment, plus idiosyncratic scale.
level_i <- 6.4 + 0.55 * seller_sd + 0.45 * log(listings / 35) +
           0.25 * fulfilled + rnorm(n, 0, 0.35)

# Trends differ by vertical and by size: parallel trends is false unit by unit.
drift_i <- 0.004 * seller_sd +
           c(Apparel = 0.000, Home = 0.002, Electronics = 0.008)[vertical]

season <- function(v, t) {
  ifelse(v == "Apparel",     0.22 * sin(2 * pi * (t - 3) / 26),
  ifelse(v == "Home",        0.12 * cos(2 * pi * (t - 1) / 26),
                             0.35 * (t / 26)^2 - 0.05))
}

Tn   <- length(weeks_all)
tmat <- matrix(rep(weeks_all, each = n), n, Tn)
y0   <- matrix(level_i, n, Tn) +
        outer(drift_i, weeks_all, "*") +
        season(matrix(vertical, n, Tn), tmat) +
        matrix(rnorm(n * Tn, 0, 0.28), n, Tn)
```

### The treatment effect

This is the heart of the simulation. Each seller's effect is a blend of two
shapes, and the weight on each is a smooth function of catalog breadth:

$$
\tau_i(S) = w_i\underbrace{\left[0.30\left(1 - e^{-S/6}\right)\right]}_{\text{compounding}}
          + (1 - w_i)\underbrace{\left[0.20\,\tfrac{S}{1.5}e^{1 - S/1.5}\right]}_{\text{novelty}},
\qquad
w_i = \text{logit}^{-1}\!\left(\frac{\log L_i - \log 50}{0.35}\right)
$$

A seller with a **broad catalog** ($L_i$ well above 50 listings) has a lot of
inventory for the optimizer to keep improving, and better listings earn better
placement, which earns more sales, which earns better placement: their lift
climbs toward 30% and stays. A seller with a **narrow catalog** optimizes their
handful of listings once, gets a 20% bump, and reverts. Nothing in the model
will be told about the threshold at 50 listings; it has to find it.

```{r treatment-effect}
S <- tmat - matrix(launch, n, Tn) + 1
S[!is.finite(S) | S < 0] <- 0
S <- matrix(as.numeric(S), n, Tn)          # weeks since launch (0 = untreated)
Z <- matrix(as.integer(S > 0), n, Tn)      # treatment indicator panel

w_i    <- plogis((log(listings) - log(50)) / 0.35)
h_grow <- function(s) 0.30 * (1 - exp(-s / 6))
h_fade <- function(s) 0.20 * (s / 1.5) * exp(1 - s / 1.5)

tau_true <- ifelse(S > 0, w_i * h_grow(S) + (1 - w_i) * h_fade(S), 0)
y        <- y0 + tau_true                  # observed log weekly GMV

broad <- listings >= 50
cat("Share of sellers with a broad catalog (>= 50 listings):",
    scales::percent(mean(broad), 0.1), "\n")
```

### Covariates handed to the model

LongBet takes **time-invariant** covariates, and its C++ backend requires the
continuous columns to come first, followed by the categorical ones, whose count
is passed as `pcat`. Two of our five columns are summaries of the seller's own
lookback window --- their average level and their trend --- and they matter
more than they look. This is discussed in @sec-longbet-baseline.

```{r covariates}
wb         <- week_base - mean(week_base)
base_level <- rowMeans(y[, week_base])
base_slope <- as.vector((y[, week_base] %*% wb) / sum(wb^2))

x <- cbind(base_level, base_slope, log(listings),      # continuous
           fulfilled, as.integer(factor(vertical)))    # categorical (pcat = 2)
colnames(x) <- c("base_level", "base_slope", "log_listings",
                 "fulfilled", "vertical")

y_train <- y[, week_study]
z_train <- Z[, week_study]
z_ext   <- Z[, c(week_study, week_future)]   # extended horizon for forecasting
```

## Looking at the Data Before Modelling It

Two pictures are worth drawing before any model is fit. The first, @fig-rollout,
is the design itself: who is treated when.

```{r}
#| label: fig-rollout
#| fig-cap: >-
#|   The randomized staggered rollout. Each row is a launch wave, each column a
#|   week. Wave membership was assigned at random within strata, so the four
#|   waves and the holdout are exchangeable by construction.
#| fig-width: 8
#| fig-height: 2.6
#| message: false

expand_grid(wave = c("W1", "W2", "W3", "W4", "Holdout"), week = week_study) %>%
  mutate(
    status = ifelse(week >= LAUNCH[wave], "Treated", "Not yet treated"),
    wave   = factor(wave, levels = c("Holdout", "W4", "W3", "W2", "W1"))
  ) %>%
  ggplot(aes(week, wave, fill = status)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  scale_fill_manual(values = c(`Not yet treated` = "grey85",
                               Treated = PAL[["LongBet"]])) +
  scale_x_continuous(breaks = week_study) +
  labs(x = "Week", y = NULL, fill = NULL,
       title = "Four randomized launch waves and a holdout") +
  theme_minimal() +
  theme(legend.position = "bottom", panel.grid = element_blank())
```

The second, @fig-trends, is the thing that would sink a naive analysis: the
untreated trajectories are not parallel. Restricting attention to the holdout --- whose
outcomes are never touched by the treatment --- apparel, home, and electronics
sellers move in visibly different directions over the study window.

```{r}
#| label: fig-trends
#| fig-cap: >-
#|   Mean log weekly GMV among holdout sellers only. The three verticals do not
#|   move in parallel, and the gaps between them change sign inside the study
#|   window. Nothing here is caused by the treatment.
#| fig-width: 8
#| fig-height: 4
#| message: false

tibble(
  vertical = rep(vertical, times = Tn),
  week     = rep(weeks_all, each = n),
  gmv_log  = as.vector(y),
  wave     = rep(wave, times = Tn)
) %>%
  filter(wave == "Holdout") %>%
  group_by(vertical, week) %>%
  summarise(gmv_log = mean(gmv_log), .groups = "drop") %>%
  ggplot(aes(week, gmv_log, colour = vertical)) +
  geom_line(linewidth = 0.9) +
  annotate("rect", xmin = min(week_study) - 0.5, xmax = max(week_study) + 0.5,
           ymin = -Inf, ymax = Inf, alpha = 0.06) +
  geom_vline(xintercept = c(10.5, 24.5), linetype = "dashed",
             colour = "grey40", linewidth = 0.3) +
  scale_colour_manual(values = c(Apparel = "#0072B2", Home = "#E69F00",
                                 Electronics = "#009E73")) +
  labs(x = "Week", y = "Mean log weekly GMV", colour = NULL,
       title = "Untreated trajectories are not parallel",
       subtitle = "Shaded region: study panel. Dashed lines: first launch and end of study.") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Because the rollout was randomized, this non-parallelism is not a threat to
identification --- it is a threat to *precision*, and a large one. Any
estimator that does not explain it has to carry it as noise.

## Fitting LongBet

```{r fit}
#| cache: true
#| message: false
#| output: false

fit_time <- system.time(
  lb_fit <- longbet(
    y = y_train, x = x, x_trt = x, z = z_train, t = week_study,
    pcat = 2,
    num_sweeps = 200, num_burnin = 60,
    num_trees_pr = 60, num_trees_trt = 60,
    sig_knl = 1, lambda_knl = 2
  )
)
```

A few arguments deserve comment:

  - `x` and `x_trt` are the covariate matrices for the prognostic and treatment
    forests. They can differ --- you might want a rich `x` for the baseline and
    a short, interpretable `x_trt` for the effect --- but here we give both the
    same five columns.
  - `t` is the calendar time index of the panel columns. LongBet infers each
    unit's time-since-adoption $S_{it}$ from `z` and `t`; you never construct
    it yourself.
  - `pcat = 2` says the last two columns of `x` are categorical. The ordering
    requirement is real and silent: get it wrong and continuous variables are
    treated as unordered levels.
  - `sig_knl` and `lambda_knl` are the variance and lengthscale of the squared
    exponential kernel on $\beta_S$. A longer lengthscale forces a smoother
    effect trajectory.

Prediction takes the fitted object, the covariates, and a treatment panel. Note
that the panel we pass for prediction is *longer* than the one we trained on:
this is what triggers the Gaussian process extrapolation of $\beta_S$ past the
largest observed time-since-adoption.

```{r predict}
#| cache: true
#| message: false
#| output: false

lb_pred <- predict.longbet(
  lb_fit, x, x, z_ext,
  t = c(week_study, week_future)
)
lb_att <- get_att(lb_pred, alpha = 0.05)
```

::: {.callout-warning title="Always pass `t` explicitly"}
`predict.longbet()` accepts `t = NULL`, but in the installed version that path
reads a variable named `longbet.fit` from the global environment and fails with
`object 'longbet.fit' not found` unless one happens to exist --- which it does
in the package's own examples, because they name the fitted object exactly
that. Even when it does exist, `t` then falls back to the *training* time grid,
which silently mismatches a longer prediction panel. Passing `t` explicitly
avoids both problems.
:::

## What the Average Effect Says

The natural summary of a staggered rollout is the average effect on the treated
as a function of **time since launch**, with every wave aligned on its own
launch week. `get_att()` does that alignment; the helper below does the same
thing for arbitrary subsets and for the known truth, so that everything we plot
is aligned identically.

```{r alignment-helpers}
# Align an [n x T x draws] array of unit-level effects on time-since-adoption
# and average over a subset of units, returning a [S x draws] matrix.
event_time_draws <- function(tauhat, z, keep = rep(TRUE, nrow(z))) {
  P    <- max(rowSums(z))
  D    <- dim(tauhat)[3]
  num  <- matrix(0, P, D)
  den  <- numeric(P)
  patt <- apply(z, 1, paste0, collapse = "")   # one pattern per launch wave
  for (p in unique(patt[keep & rowSums(z) > 0])) {
    ii   <- which(patt == p & keep)
    cols <- which(z[ii[1], ] == 1)
    k    <- length(cols)
    num[1:k, ] <- num[1:k, ] + apply(tauhat[ii, cols, , drop = FALSE], c(2, 3), sum)
    den[1:k]   <- den[1:k] + length(ii)
  }
  num / den
}

# Same alignment for a known [n x T] matrix of true effects.
event_time_truth <- function(m, z, keep = rep(TRUE, nrow(z))) {
  as.vector(event_time_draws(array(m, c(dim(m), 1)), z, keep))
}

tau_future  <- tau_true[, c(week_study, week_future)]
truth_att   <- event_time_truth(tau_future, z_ext)
S_observed  <- max(rowSums(z_train))   # longest time-since-launch in the study
S_max       <- max(rowSums(z_ext))     # longest available for forecasting
```

The benchmark is the estimator a careful analyst would reach for first:
for each wave and each week since its launch, take the change in log GMV
relative to the common pre-launch window (weeks 7--10) and subtract the same
change among holdout sellers, then pool the waves. Random assignment makes it
unbiased; the within-seller differencing removes each seller's level entirely,
which is a stronger adjustment than any covariate.

```{r benchmark}
pre_weeks <- 7:10                # common pre-launch window for every wave
is_hold   <- wave == "Holdout"
y_pre     <- rowMeans(y[, pre_weeks])
dY        <- y - y_pre           # change from each seller's own baseline
                                 # (weeks_all starts at 1, so column == week)

did_holdout <- map_dfr(1:S_observed, function(s) {
  # Waves that have reached event time s before the study window closes
  ws <- c("W1", "W2", "W3", "W4")
  ws <- ws[LAUNCH[ws] + s - 1 <= max(week_study)]
  cw <- as.integer(LAUNCH[ws] + s - 1)
  nt <- vapply(ws, function(w) sum(wave == w), integer(1))
  a  <- nt / sum(nt)             # sample-size weights across waves

  treated_mean <- vapply(seq_along(ws),
                         function(j) mean(dY[wave == ws[j], cw[j]]), numeric(1))
  treated_var  <- vapply(seq_along(ws),
                         function(j) var(dY[wave == ws[j], cw[j]]) / nt[j],
                         numeric(1))

  # Every wave is compared against the *same* holdout sellers, so the holdout
  # contributes one weighted combination per seller. Treating the waves'
  # holdout means as independent would understate the variance.
  h_combo <- as.vector(dY[is_hold, cw, drop = FALSE] %*% a)

  est <- sum(a * treated_mean) - mean(h_combo)
  se  <- sqrt(sum(a^2 * treated_var) + var(h_combo) / sum(is_hold))
  tibble(s = s, estimate = est, lower = est - 1.96 * se, upper = est + 1.96 * se)
})
```

```{r}
#| label: fig-att
#| fig-cap: >-
#|   Average effect on the treated by weeks since launch, on the log GMV scale.
#|   Both estimators track the truth. LongBet's credible bands are far narrower
#|   because the prognostic forest has already explained away seller level and
#|   vertical seasonality before the effect is read off.
#| fig-width: 8.5
#| fig-height: 4.6
#| message: false

att_df <- tibble(
  s        = seq_len(S_max),
  estimate = lb_att$att,
  lower    = lb_att$intervals[1, ],
  upper    = lb_att$intervals[2, ],
  truth    = truth_att
) %>% filter(s <= S_observed)

ggplot(att_df, aes(s)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = "LongBet"), alpha = 0.22) +
  geom_line(aes(y = estimate, colour = "LongBet"), linewidth = 0.9) +
  geom_line(aes(y = truth, colour = "Truth"), linewidth = 0.9) +
  geom_pointrange(
    data = did_holdout,
    aes(x = s, y = estimate, ymin = lower, ymax = upper,
        colour = "DiD vs holdout"),
    size = 0.28, position = position_nudge(x = 0.14)
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  scale_colour_manual(values = PAL) +
  scale_fill_manual(values = PAL, guide = "none") +
  scale_x_continuous(breaks = 1:S_observed) +
  labs(
    x = "Weeks since launch", y = "Effect on log weekly GMV", colour = NULL,
    title = "The average lift fades over the first three months",
    subtitle = "Effect on the treated, aligned on each wave's own launch week"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r att-scorecard}
ins_ <- seq_len(S_observed)

score <- function(est, lo, hi, truth) {
  tibble(
    RMSE          = sqrt(mean((est - truth)^2)),
    `Mean bias`   = mean(est - truth),
    `Interval width` = mean(hi - lo),
    `Contained truth` = sprintf("%d of %d", sum(truth >= lo & truth <= hi), length(truth))
  )
}

bind_rows(
  score(att_df$estimate, att_df$lower, att_df$upper, att_df$truth) %>%
    mutate(Estimator = "LongBet", .before = 1),
  score(did_holdout$estimate, did_holdout$lower, did_holdout$upper,
        truth_att[ins_]) %>%
    mutate(Estimator = "DiD vs holdout", .before = 1)
) %>%
  kable(digits = 4, caption = paste(
    "Accuracy and precision of the two estimators across the",
    S_observed, "observed event times. Both are aimed at the same estimand."
  ))
```

Both estimators recover the shape, which is the reassurance we wanted: the
model is not inventing anything the randomization does not support. The
difference is width. LongBet's intervals are
`r sprintf("%.1f", mean(did_holdout$upper - did_holdout$lower) / mean(lb_att$intervals[2, ins_] - lb_att$intervals[1, ins_]))`
times narrower on average, and that is not a free lunch --- it is the payoff
for having modelled $\mu(X_i, t)$ explicitly rather than differencing it away.
The same precision would otherwise have to be bought with a bigger experiment
or a longer one.

::: {.callout-important title="One realization is not a coverage study"}
The "contained truth" column above is *not* an estimate of coverage. The
14 event times share the same treated sellers and the same holdout, so their
errors are strongly correlated: a single unlucky draw can push the whole curve
off at once. Interval *width* is stable enough to compare from one run;
coverage is not. @wang2024longbet report 94%--98% coverage for LongBet's ATT
across 100 Monte Carlo replications of a similar design, which is the kind of
evidence that claim requires. That said, LongBet's intervals rest on an
assumption the difference-in-differences intervals do not need, and
@sec-longbet-limits shows the assumption failing in exactly this dataset.
:::

Read literally, @fig-att is discouraging. The lift peaks at about
`r scales::percent(max(truth_att[ins_]), 0.1)` two weeks after launch, slides
for the next month and a half, and is still only
`r scales::percent(truth_att[S_observed], 0.1)` at week `r S_observed` --- about
a third of the peak surrendered. That is the signature of a novelty effect, and
the standard conclusion is that the optimizer buys a temporary bump not worth a
permanent cost. Every number on that chart is correct. The conclusion drawn
from it is wrong.

## What the Average Hides

The pooled curve in @fig-att is an average over two populations that are doing
opposite things. LongBet gives us the posterior for every seller at every
event time, so splitting it costs nothing but a subset.

```{r}
#| label: fig-catt-segments
#| fig-cap: >-
#|   The same effect, split by catalog breadth. Sellers with broad catalogs are
#|   on a compounding trajectory that is at its maximum when the study ends;
#|   sellers with narrow catalogs get a two-week bump and give almost all of it
#|   back. The pooled average in @fig-att is the weighted mixture of these two,
#|   and describes neither.
#| fig-width: 8.5
#| fig-height: 4.6
#| message: false

segment_att <- function(keep, label) {
  d <- event_time_draws(lb_pred$tauhats, z_ext, keep)
  tibble(
    s        = seq_len(nrow(d)),
    estimate = rowMeans(d),
    lower    = apply(d, 1, quantile, 0.025),
    upper    = apply(d, 1, quantile, 0.975),
    truth    = event_time_truth(tau_future, z_ext, keep),
    segment  = label
  )
}

segments_df <- bind_rows(
  segment_att(broad,  "Broad catalog"),
  segment_att(!broad, "Narrow catalog")
) %>% filter(s <= S_observed)

ggplot(segments_df, aes(s, colour = segment, fill = segment)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18, colour = NA) +
  geom_line(aes(y = estimate), linewidth = 0.9) +
  geom_line(aes(y = truth), linewidth = 0.7, linetype = "22") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  scale_colour_manual(values = PAL) +
  scale_fill_manual(values = PAL, guide = "none") +
  scale_x_continuous(breaks = 1:S_observed) +
  labs(
    x = "Weeks since launch", y = "Effect on log weekly GMV", colour = NULL,
    title = "Two opposite trajectories inside one average",
    subtitle = "Solid: posterior mean with 95% band. Dashed: truth."
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Nothing told the model where to split. It was handed `log_listings` as one of
five covariates and found, on its own, that the *shape* of the response over
time-since-launch depends on it. To see how sharply, we can ask the model for
the effect of a full quarter of exposure for every seller in the population,
whether or not they were actually treated: build a treatment panel in which
everyone launches in week 11 and read off the last column.

```{r counterfactual}
#| cache: true
#| message: false
#| output: false

# Counterfactual panel: every seller launches in week 11.
z_all <- cbind(matrix(0L, n, sum(week_study < 11)),
               matrix(1L, n, sum(week_study >= 11)))
pred_all <- predict.longbet(lb_fit, x, x, z_all, t = week_study)

S_target      <- ncol(z_all) - sum(week_study < 11)     # = 14 weeks of exposure
tau_hat_draws <- pred_all$tauhats[, ncol(z_all), ]      # [n x draws] at S_target
tau_hat       <- rowMeans(tau_hat_draws)
tau_truth_S   <- w_i * h_grow(S_target) + (1 - w_i) * h_fade(S_target)
```

```{r}
#| label: fig-catt-listings
#| fig-cap: >-
#|   Estimated versus true effect after 14 weeks of exposure, for all 3,000
#|   sellers. The model recovers a smooth transition centred near the 50-listing
#|   threshold that generated the data, without being told the threshold exists.
#| fig-width: 8.5
#| fig-height: 4.4
#| message: false

tibble(listings = listings, estimate = tau_hat, truth = tau_truth_S) %>%
  ggplot(aes(listings)) +
  geom_point(aes(y = estimate, colour = "LongBet"), alpha = 0.16, size = 0.8) +
  geom_line(aes(y = truth, colour = "Truth"), linewidth = 1) +
  geom_vline(xintercept = 50, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 52, y = 0.02, label = "50 listings",
           hjust = 0, size = 3.2, colour = "grey35") +
  scale_x_log10(breaks = c(5, 10, 25, 50, 100, 250, 500)) +
  scale_colour_manual(values = PAL) +
  labs(
    x = "Active listings (log scale)", y = "Effect on log weekly GMV",
    colour = NULL,
    title = "Effect after 14 weeks of exposure, seller by seller"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r catt-quality}
tibble(
  Quantity = c("Correlation with truth", "RMSE", "Mean absolute error"),
  Value = c(cor(tau_hat, tau_truth_S),
            sqrt(mean((tau_hat - tau_truth_S)^2)),
            mean(abs(tau_hat - tau_truth_S)))
) %>%
  kable(digits = 3, caption = paste(
    "Quality of the seller-level effect estimates after", S_target,
    "weeks of exposure."
  ))
```

This is the resolution a difference-in-differences cannot give you. It is one
posterior per seller, obtained without pre-registering a single subgroup.

## The Decision

The question was never "what is the average treatment effect." It was whether
to make the optimizer permanent, and for whom. The marketplace keeps 12% of
GMV and the optimizer costs \$500 per seller per year, so a seller is worth
enabling only if the extra GMV they generate, valued at the take rate, clears
that cost.

We value one year of exposure per seller. For the first 14 weeks we use the
model's own trajectory, $\hat\tau_i(1), \dots, \hat\tau_i(14)$; for the
remaining 38 weeks we hold the effect at its week-14 level. That second
assumption is the one worth arguing about, and @sec-longbet-forecast is where
we test how far it can be defended. Because the model estimates a shift in
*log* GMV and the treatment does not change the shape of the residual
distribution, $e^{\hat\tau} - 1$ is the proportional effect on mean GMV.

```{r economics}
TAKE_RATE   <- 0.12
ANNUAL_COST <- 500
HOLD_WEEKS  <- 52 - S_target      # weeks held at the week-14 effect

gmv_week <- rowMeans(exp(y[, week_base]))   # observed baseline $/week per seller

# Effect trajectory for every seller over 14 weeks of exposure: [n x 14 x draws]
tau_path_draws <- pred_all$tauhats[, (sum(week_study < 11) + 1):ncol(z_all), ]

# Annual GMV lift, expressed in "weeks of baseline GMV": [n x draws]
annual_lift <- apply(exp(tau_path_draws) - 1, c(1, 3), sum) +
               HOLD_WEEKS * (exp(tau_hat_draws) - 1)

net_draws <- TAKE_RATE * gmv_week * annual_lift - ANNUAL_COST
p_worth   <- rowMeans(net_draws > 0)

# The same quantity computed from the known truth, for scoring the decision.
annual_lift_true <- rowSums(exp(outer(w_i, 1:S_target, function(a, s)
                       a * h_grow(s) + (1 - a) * h_fade(s))) - 1) +
                    HOLD_WEEKS * (exp(tau_truth_S) - 1)
net_true <- TAKE_RATE * gmv_week * annual_lift_true - ANNUAL_COST

breakeven <- ANNUAL_COST / (TAKE_RATE * 52 * gmv_week)

cat("Break-even annual GMV lift -- median seller: ",
    scales::percent(median(breakeven), 0.1),
    "; the smallest quarter of sellers need at least ",
    scales::percent(quantile(breakeven, 0.75), 0.1), "\n", sep = "")
```

```{r policy-table}
policy_value <- function(keep, label) {
  if (!any(keep)) {
    return(tibble(Policy = label, Sellers = 0L,
                  `Estimated (k/yr)` = 0, `95% CrI` = "--",
                  `Realized (k/yr)` = 0))
  }
  total <- colSums(net_draws[keep, , drop = FALSE])
  tibble(
    Policy             = label,
    Sellers            = sum(keep),
    `Estimated (k/yr)` = mean(total) / 1000,
    `95% CrI`          = sprintf("[%.0f, %.0f]", quantile(total, 0.025) / 1000,
                                                 quantile(total, 0.975) / 1000),
    `Realized (k/yr)`  = sum(net_true[keep]) / 1000
  )
}

bind_rows(
  policy_value(rep(FALSE, n),      "Kill the feature"),
  policy_value(rep(TRUE, n),       "Ship to everyone"),
  policy_value(broad,              "Ship to broad catalogs (>= 50 listings)"),
  policy_value(p_worth >= 0.90,    "Ship where P(net value > 0) >= 0.90"),
  policy_value(net_true > 0,       "Oracle (knows the truth)")
) %>%
  kable(digits = 0, caption = paste0(
    "Annual net marketplace revenue under five rollout policies, in thousands ",
    "of US dollars per year. 'Estimated' is the posterior from the fitted ",
    "model; 'Realized' scores the same policy against the true effects."
  ))
```

```{r effect-vs-value}
tibble(
  Segment = c("Narrow catalog (< 50 listings)", "Broad catalog (>= 50 listings)"),
  Sellers = c(sum(!broad), sum(broad)),
  `Median true lift after 14 weeks` = scales::percent(
    c(median(exp(tau_truth_S[!broad]) - 1), median(exp(tau_truth_S[broad]) - 1)), 0.1),
  `Share actually worth enabling` = scales::percent(
    c(mean(net_true[!broad] > 0), mean(net_true[broad] > 0)), 0.1)
) %>%
  kable(caption = paste(
    "Effect size and economic value do not select the same sellers.",
    "'Worth enabling' means the true annual net value exceeds zero."
  ))
```

Three things to take from these tables.

First, **blanket rollout is profitable and still leaves a great deal on the
table.** Shipping to everyone beats killing the feature, which is the
conclusion a well-run experiment reporting only an average effect would reach
--- and it is worth roughly a third less than the best available policy,
because a large share of the population is being charged \$500 a year for an
effect that has gone to zero long before the bill arrives.

Second, and less obviously, **effect heterogeneity is not value
heterogeneity.** The "broad catalog" rule follows the treatment effect, which
is what @fig-catt-segments and @fig-catt-listings are about. The money follows
effect *times* seller size, because the \$500 is flat. That is why a
non-trivial share of narrow-catalog sellers are still worth enabling --- they
are large enough that even a small residual lift pays the bill --- while some
broad-catalog sellers are not. A targeting rule read straight off the
heterogeneity plot gets both groups wrong, and the table shows it: the
posterior net-value rule enables *fewer* sellers than the catalog rule and
earns more. The rule you want is written on the posterior of *net value*, not
on the posterior of the effect.

Third, **the model is conservative about the money.** The estimated column sits
a few percent below the realized column for every policy. That is the treatment
prior doing its job: BCF-style models shrink $\tau$ toward homogeneity, so the
business case they produce is slightly understated. A second, opposing force is
present too --- selecting sellers on `P(net value > 0) >= 0.90` and then valuing
that same selection with the same posterior double-counts the noise that pushed
marginal sellers over the threshold. Here shrinkage is the larger of the two,
but on a smaller experiment, or with a looser threshold, the winner's curse can
dominate. Either way, the realized column is the one the truth is measured
against, and only a simulation has it.

## Projecting Past the End of the Study {#sec-longbet-forecast}

The business case runs for a year; the study ran for 14 weeks after the first
launch. LongBet is the only tool in this chapter that will say anything about
week 20, and it is worth being precise about what it is actually doing when it
does.

Prediction on a treatment panel longer than the training panel triggers
Gaussian process extrapolation of $\beta_S$: the shared factor is projected
forward under a squared-exponential kernel whose variance and lengthscale you
supply as `sigma` and `lambda`. Two properties of that machinery decide how
much the projection is worth:

  - **Only $\beta_S$ is extrapolated.** The forest $\nu(X_i, S, t)$ has no
    splits beyond the largest $S$ it was trained on, so it is frozen at its
    last value. Whatever *shape* heterogeneity the forest discovered inside the
    window is rescaled by the projection, not extended by it.
  - **The Gaussian process has mean zero.** Far enough past the data, the
    projection says the effect goes away --- because that is what the prior
    says, not because the data said so. `lambda` controls how fast "far enough"
    arrives.

Weeks 25--30 of our simulation were never shown to the model, so we can score
the projection instead of admiring it. (The projected draws are not seeded
reproducibly --- see @sec-longbet-limits --- so re-running this code moves the
numbers to the right of week `r S_observed` a little, while everything to the
left of it is fixed.)

```{r forecast-grid}
#| cache: true
#| message: false
#| output: false

gp_grid <- expand_grid(sigma = c(0.3, 1.0), lambda = c(2, 5, 8, 12, 20))

gp_runs <- pmap_dfr(gp_grid, function(sigma, lambda) {
  pr <- predict.longbet(lb_fit, x, x, z_ext, t = c(week_study, week_future),
                        sigma = sigma, lambda = lambda)
  a  <- get_att(pr, alpha = 0.05)
  tibble(sigma = sigma, lambda = lambda, s = seq_len(S_max),
         estimate = a$att, lower = a$intervals[1, ], upper = a$intervals[2, ])
})
```

```{r forecast-scores}
gp_runs %>%
  filter(s > S_observed) %>%
  left_join(tibble(s = seq_len(S_max), truth = truth_att), by = "s") %>%
  group_by(sigma, lambda) %>%
  summarise(
    RMSE             = sqrt(mean((estimate - truth)^2)),
    `Mean width`     = mean(upper - lower),
    `Contained truth` = sprintf("%d of %d", sum(truth >= lower & truth <= upper), n()),
    .groups = "drop"
  ) %>%
  kable(digits = 4, caption = paste0(
    "Projection quality over the ", S_max - S_observed,
    " withheld weeks, across Gaussian process settings. The truth over this ",
    "window is essentially flat near ",
    scales::percent(mean(truth_att[(S_observed + 1):S_max]), 0.1), "."
  ))
```

```{r}
#| label: fig-forecast
#| fig-cap: >-
#|   Projecting the average effect past the end of the study. Everything left of
#|   the dashed line is estimated from data the model saw; everything right of
#|   it is a projection of the shared factor under the stated Gaussian process
#|   prior. The lengthscale, not the data, decides how quickly the projected
#|   effect decays.
#| fig-width: 8.5
#| fig-height: 4.8
#| message: false

fc_show <- gp_runs %>%
  filter(sigma == 1.0, lambda %in% c(2, 8, 20)) %>%
  mutate(setting = factor(paste0("lambda = ", lambda),
                          levels = paste0("lambda = ", c(2, 8, 20))))

ggplot(fc_show, aes(s)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = PAL[["LongBet"]], alpha = 0.18) +
  geom_line(aes(y = estimate), colour = PAL[["LongBet"]], linewidth = 0.9) +
  geom_line(data = tibble(s = seq_len(S_max), truth = truth_att),
            aes(y = truth), colour = PAL[["Truth"]], linewidth = 0.8) +
  geom_vline(xintercept = S_observed + 0.5, linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60") +
  facet_wrap(~setting) +
  labs(x = "Weeks since launch", y = "Effect on log weekly GMV",
       title = "The same fit, three Gaussian process lengthscales",
       subtitle = "Black: truth. Blue: LongBet posterior mean and 95% band. sigma = 1 throughout.") +
  theme_minimal()
```

Two lessons come out of that table and figure, and neither is about LongBet
specifically.

**The lengthscale, not the data, decides the shape of the projection.** A short
`lambda` pulls the effect to zero within a couple of weeks --- not because
anything in the study suggested it would collapse, but because a
squared-exponential Gaussian process with mean zero forgets the data that fast.
A `lambda` comfortably longer than the projection horizon carries the level
forward. The observed window cannot tell you which is right; you are choosing
between priors, and you should choose the one that matches what you actually
believe about how product effects decay in your business.

**Do not shrink `sigma` to make the bands look good.** Tightening the prior
variance narrows the projected interval, and it narrows it around whatever the
lengthscale happened to imply. The wide bands under `sigma = 1` are the honest
ones: they say the projection is a prior-driven scenario, which it is.

There is also a structural reason a projection can understate a
still-compounding effect, and it is a property of the model rather than a
tuning failure. The shared factor and the population-average effect do not
decline at the same rate:

```{r beta-decline}
post_sweeps <- (lb_fit$model_params$burnin + 1):ncol(lb_fit$beta_values)
beta_post   <- rowMeans(lb_fit$beta_values[, post_sweeps])   # beta_S, S = 0..14

cat("Shared factor beta_S falls ",
    scales::percent(1 - beta_post[length(beta_post)] / max(beta_post), 1),
    " from its peak.\n",
    "True average effect falls ",
    scales::percent(1 - truth_att[S_observed] / max(truth_att[ins_]), 1),
    " from its peak.\n", sep = "")
```

$\beta_S$ sheds more than half of its peak; the average effect gives back only
about a third of its. The forest absorbs the difference by making $\nu$ increase
in $S$. That split is a perfectly good description of the data inside the
window --- but it parks a large share of the late-window dynamics inside the
component that does *not* extrapolate. It is why the projections in the table
that miss, miss low.

::: {.callout-tip title="`split_time_trt`: shape heterogeneity or a clean forecast"}
`longbet()` takes `split_time_trt`, which controls whether the treatment forest
may split on time-since-adoption. It is the switch that governs this trade-off
directly:

  - `split_time_trt = TRUE` (the default, used here) lets different kinds of
    units have different effect *shapes* --- which is the only reason
    @fig-catt-segments works --- at the cost of parking dynamics in a component
    that cannot be projected.
  - `split_time_trt = FALSE` forces $\nu(X_i, t)$ to be constant in $S$, so
    $\tau_i(S) = (\beta_S - \beta_0)\,\nu(X_i, t)$: one shared shape, scaled per
    unit. Every unit's trajectory is then a rescaling of $\beta_S$, which is
    exactly the thing the Gaussian process extrapolates, so projections behave
    much better --- and the compounding-versus-fading split this chapter is
    about becomes inexpressible, because every unit's path is now the same
    curve up to a scale factor.

Fit both when the forecast matters. If they disagree about the trajectory, the
disagreement is telling you that your effect's shape is heterogeneous, and that
the forecast from either one deserves a wide error bar.
:::

The practical guidance that follows is deliberately modest. Use the projection
to ask *how sensitive is my business case to the effect decaying*, not to
produce the number that goes in the plan. Set `lambda` at least as long as the
horizon you intend to project. And sanity check it against what you can already
see: in @fig-catt-segments the broad-catalog trajectory is at its maximum at
week 14 with no sign of turning down, so any projection in which it collapses
is contradicting evidence you already have.

## Practical Notes {#sec-longbet-limits}

### Give the model each unit's own history {#sec-longbet-baseline}

This is the single most consequential thing to get right, and it is easy to
miss. **LongBet has no unit fixed effects.** Everything it knows about a seller
has to arrive through the time-invariant covariate vector $X_i$. A within-unit
difference-in-differences removes each seller's level exactly and for free;
LongBet has to *reconstruct* it from covariates.

That is why `base_level` and `base_slope` --- the seller's mean and trend over
a lookback window that is deliberately excluded from the modelled panel --- are
the first two columns of `x`. Refitting without them shows what they are worth:

```{r fit-thin}
#| cache: true
#| message: false
#| output: false

x_thin <- x[, c("log_listings", "fulfilled", "vertical")]   # pcat still 2

lb_thin      <- longbet(y = y_train, x = x_thin, x_trt = x_thin, z = z_train,
                        t = week_study, pcat = 2,
                        num_sweeps = 200, num_burnin = 60,
                        num_trees_pr = 60, num_trees_trt = 60,
                        sig_knl = 1, lambda_knl = 2)
lb_thin_pred <- predict.longbet(lb_thin, x_thin, x_thin, z_ext,
                                t = c(week_study, week_future))
lb_thin_att  <- get_att(lb_thin_pred, alpha = 0.05)
```

```{r thin-comparison}
resid_sd <- function(fit) {
  s <- fit$sigma0_draws[, ncol(fit$sigma0_draws)]
  mean(s) * fit$sdy                     # sigma is on the standardized scale
}
ins <- seq_len(S_observed)

tibble(
  Covariates = c("With lookback summaries", "Structural covariates only"),
  `Residual SD` = c(resid_sd(lb_fit), resid_sd(lb_thin)),
  `Mean interval width` = c(
    mean(lb_att$intervals[2, ins] - lb_att$intervals[1, ins]),
    mean(lb_thin_att$intervals[2, ins] - lb_thin_att$intervals[1, ins])
  ),
  RMSE = c(
    sqrt(mean((lb_att$att[ins] - truth_att[ins])^2)),
    sqrt(mean((lb_thin_att$att[ins] - truth_att[ins])^2))
  )
) %>%
  kable(digits = 4, caption = paste(
    "Dropping the two lookback summaries from x. The true residual standard",
    "deviation in this simulation is 0.28."
  ))
```

Compute those summaries on a window that is **not** part of the panel you
model. Using weeks that also appear as outcomes makes the prognostic forest
partly a function of its own target. The point estimate of the treatment
contrast survives that, but the residual variance does not: it is deflated by
the leakage, and every credible interval downstream of it inherits the
deflation.

### The error term is independent by assumption

LongBet's $\epsilon_{it}$ is independent across units *and across time*. Real
panels usually are not: a seller who has a good week tends to have a good next
week for reasons no covariate captures. When that is true, the effective sample
size is smaller than $n \times T$ and every credible interval in this chapter
is too narrow. It costs nothing to check, using the holdout sellers whose
outcomes the treatment never touched:

```{r residual-acf}
mu_hat  <- apply(lb_pred$muhats[, seq_along(week_study), ], c(1, 2), mean)
resid_h <- (y[, week_study] - mu_hat)[is_hold, ]

lag1 <- mean(mapply(function(a, b) cor(resid_h[, a], resid_h[, b]),
                    1:(ncol(resid_h) - 1), 2:ncol(resid_h)))

cat("Mean lag-1 residual correlation among holdout sellers:",
    round(lag1, 3), "\n")
```

It is not zero --- and the simulation generated independent errors, so this is
not autocorrelated noise. It is the part of each seller's persistent level that
the prognostic forest did not manage to reconstruct from $X_i$, showing up as
persistence in what is left over. The consequence is the same either way: the
effective sample size is smaller than $n \times T$, and the credible intervals
are correspondingly optimistic. It is the most likely reason LongBet's much
narrower intervals contained the truth at fewer event times than the
difference-in-differences intervals did.

On real data, treat a materially positive value as a signal to widen the
intervals --- by a block bootstrap over units, for instance --- before anyone
quotes them, and as a prompt to go find better unit-level covariates.

### Column order, `pcat`, and other sharp edges

  - **Continuous columns first, categorical last**, and `pcat` is the count of
    the categorical ones. Getting this wrong does not raise an error; it
    silently treats a continuous variable as an unordered factor.
  - **`longbet()` ignores `set.seed()`.** The sampler seeds itself internally
    (`random_seed = 0`), so fits are reproducible but insensitive to R's RNG
    state. To probe sensitivity, vary `num_sweeps` and the tree counts, or
    resample the data --- refitting alone will return the identical answer.
  - **The extrapolation is not deterministic.** `predict.longbet()` draws the
    projected $\beta_S$ from a system-entropy-seeded generator. Two calls with
    identical arguments return identical estimates inside the observed window
    and different ones outside it. Re-running the projection a few times is a
    free, and revealing, robustness check.
  - **Covariates must be time-invariant.** Anything that changes over the panel
    has to be summarized into a pre-treatment number, which is a real
    restriction and the most obvious direction for future work on the model.

### Design notes for the experiment itself

  - **The event-time tail is thin.** In this rollout the last event time is
    reached by one wave only. Estimates and intervals at the far right of
    @fig-att rest on a quarter of the treated sample, and the Gaussian process
    extrapolation starts from exactly there. If the long-run effect is the
    decision-relevant quantity, launch the waves close together and run the
    study longer, rather than spreading the launches out.
  - **Randomize the launch order, and stratify it.** Letting the field team
    choose who goes first turns a clean experiment into an observational study
    with targeted selection --- the setting BCF and LongBet were built to
    survive, but survival is not the same as free precision.
  - **SUTVA does not come with randomization.** If sellers compete for the same
    buyers, better listings for the treated group can come partly at the
    expense of the holdout, and the rollout measures a relative advantage
    rather than the effect of treating everyone. Randomizing at a level with
    less interference --- category, geography, marketplace --- is a design
    answer to a problem no model can fix.
  - **Used observationally, everything rests on $X$.** Outside an experiment,
    LongBet trades the parallel trends assumption for the assumption that
    $X_i$ contains every confounder of adoption timing and outcome. That is a
    different assumption, not a weaker one; it is only an improvement when you
    actually observe the drivers of adoption.

## Conclusion

LongBet earns its place in a business setting by answering a question that
neither a cross-sectional causal forest nor a classical panel estimator can:
not *what is the effect*, but *what does the effect do over time, and to whom*.

In this chapter the pooled average told a clean and completely misleading
story --- a lift that gave back a third of its peak inside three months, the
textbook signature of a novelty effect. The same model, asked one level down,
found two populations moving in opposite directions: broad-catalog sellers on a
compounding trajectory that showed no sign of turning down when the study
ended, and narrow-catalog sellers whose bump was largely spent by the second
month. Acting on that distinction was worth roughly half a million dollars a
year in a 3,000-seller pilot, and no amount of extra sample would have
surfaced it from an average.

Three cautions are worth carrying forward. The randomization, not the model, is
what makes the estimates credible: keep a simple difference-in-differences
against the holdout in view as a check, and treat disagreement as a bug report.
The credible intervals assume errors are independent across units and time, and
they will be too narrow when that fails. And the forecast extrapolates one
shared factor under a prior you choose, so it belongs in a sensitivity analysis
rather than in the plan.

::: {.callout-tip}
## Learn more
  - @wang2024longbet LongBet: Heterogeneous Treatment Effect Estimation in
    Panel Data --- the method, its simulation study, and the forecasting
    extension.
  - @longbetpkg The `longbet` R package.
  - @hahn2020bayesian Bayesian regression tree models for causal inference ---
    the cross-sectional model LongBet extends.
  - @wang2024localgp Local Gaussian process extrapolation for BART models ---
    the alternative extrapolation strategy referenced by @wang2024longbet.
  - @callaway2021did and @roth2023trending for the econometric literature on
    staggered adoption that LongBet is positioned against.
  - @deng2013cuped for the variance-reduction idea that $\mu(X_i, t)$
    generalizes.
:::
