---
title: "LongBet: Forecasting and Observational Panels"
description: "Four practical extensions: projecting the trajectory past the end of the study, panels where adoption was chosen rather than randomized, where a unit's own level comes from, and covariates that move within a unit over time."
format:
  html:
    css: longbet.css
share:
  permalink: "https://book.martinez.fyi/longbet_extensions.html"
  description: "LongBet: forecasting, observational panels, and practical modelling"
  linkedin: true
  email: true
  mastodon: true
---

::: {.callout-note title="Continues the marketplace rollout"}
This chapter reuses the sellers, calendar and fitted model of **[LongBet: Dynamic Treatment Effects in Staggered Rollouts](longbet.qmd)** and the valuation of **[LongBet: From Effects to Decisions](longbet_decisions.qmd)**. The estimator comparison sits in **[LongBet and Modern Difference-in-Differences](longbet_did.qmd)**.
:::

```{r setup}
#| message: false
#| warning: false
library(longbet)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(purrr)
library(knitr)

source("R/longbet-common.R")
source("R/longbet-sim.R")
source("R/longbet-artifacts.R")
verify_longbet_environment()
theme_set(theme_minimal(base_size = 12))

ro <- simulate_rollout()
list2env(ro[c("n", "week_base", "week_study", "week_future", "weeks_all", "LAUNCH", "Tn",
              "listings", "level_i", "a_i", "S", "Z", "tau_true", "y", "x", "base_level",
              "sd_eps", "y_train", "z_train", "z_ext", "is_hold", "strata", "launch")],
         environment())
lb_fit <- lb_core_fit(ro)
S_observed <- max(rowSums(z_train)); S_max <- max(rowSums(z_ext))
truth_att <- event_time_truth(tau_true[, match(week_study, weeks_all)], z_train)
truth_att_ext <- event_time_truth(tau_true[, match(c(week_study, week_future), weeks_all)], z_ext)
```

## Projecting Past the End of the Study {#sec-longbet-forecast}

The business case runs for a year and the study ran for sixteen weeks. Weeks 21 to 24 were simulated but never shown to the model, which makes them a fair test of what it can say about periods it has not seen.

The Gaussian process on $\beta_S$ is what makes the projection possible. Because the trajectory is a smooth function of exposure rather than a set of unrelated period effects, it can be extended past the largest exposure in the data, with uncertainty governed by the kernel. The forests do not extrapolate: they evaluate the partitions they already have. Passing `sig_knl` and `lambda_knl` to `predict()` changes **only** the projection, holding the fitted in-window draws fixed, so a sweep over them is a sensitivity analysis of the extrapolation assumption rather than a refit.

```{r forecast}
#| message: false
s_obs <- t(apply(z_train, 1, cumsum)); s_ext <- t(apply(z_ext, 1, cumsum))
n_obs <- tabulate(s_obs[z_train == 1], nbins = S_max); n_ext <- tabulate(s_ext[z_ext == 1], nbins = S_max)
n_future <- n_ext - n_obs; keep_s <- which(n_future > 0)
lb_att <- get_att(predict(lb_fit, x = x, z = z_train, t = week_study, summary_only = TRUE, random_seed = 1))
obs_draws <- matrix(0, S_max, ncol(lb_att$att_full)); obs_draws[seq_len(S_observed), ] <- lb_att$att_full
truth_obs <- numeric(S_max); truth_obs[seq_len(S_observed)] <- truth_att
forecast_truth <- ((truth_att_ext * n_ext - truth_obs * n_obs) / pmax(n_future, 1))[keep_s]

gp_grid <- expand_grid(sigma = c(0.3, 1.0), lambda = c(2, 12, 20))
gp_future <- vector("list", nrow(gp_grid))
gp_runs <- map_dfr(seq_len(nrow(gp_grid)), function(j) {
  pr <- predict(lb_fit, x = x, z = z_ext, t = c(week_study, week_future),
                sig_knl = gp_grid$sigma[j], lambda_knl = gp_grid$lambda[j],
                summary_only = TRUE, random_seed = 1)
  a <- get_att(pr)
  fut <- ((a$att_full * n_ext - obs_draws * n_obs) / pmax(n_future, 1))[keep_s, , drop = FALSE]
  gp_future[[j]] <<- fut
  tibble(sigma = gp_grid$sigma[j], lambda = gp_grid$lambda[j], s = keep_s,
         estimate = rowMeans(fut), lower = apply(fut, 1, quantile, .025),
         upper = apply(fut, 1, quantile, .975), truth = forecast_truth)
})
forecast_diag <- bind_rows(lapply(seq_len(nrow(gp_grid)), function(j)
  diagnose_draws(gp_future[[j]], lb_fit,
                 sprintf("Withheld weeks, kernel SD %.1f, lengthscale %d",
                         gp_grid$sigma[j], gp_grid$lambda[j]))$summary))
```

The extended average mixes cells the model saw with cells it did not, so the future-only average is recovered by subtracting the observed part draw by draw. Without that step a forecast score is flattered by the in-sample cells inside it.

```{r fig-forecast}
#| fig-cap: "Effects in the four withheld weeks, averaged within exposure time, under three projection lengthscales at kernel SD 1. Black is the matching simulated truth; the dotted line marks the last exposure seen in training."
#| fig-alt: "Three panels. In each, a blue curve with a band tracks a black truth curve across exposures twelve to sixteen, with the band widening past the dotted line."
#| fig-width: 8.5
#| fig-height: 4
gp_runs %>% filter(sigma == 1) %>%
  mutate(setting = factor(paste0("lengthscale = ", lambda), levels = paste0("lengthscale = ", c(2, 12, 20)))) %>%
  ggplot(aes(s)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = PAL[["LongBet"]], alpha = .18) +
  geom_line(aes(y = estimate), colour = PAL[["LongBet"]], linewidth = .9) +
  geom_line(aes(y = truth), colour = PAL[["Truth"]], linewidth = .8) +
  geom_vline(xintercept = S_observed + 0.5, linetype = "dotted", colour = "grey40") +
  facet_wrap(~ setting) +
  labs(x = "Exposure time among withheld-week cells", y = "Effect on log weekly GMV")
```

```{r forecast-scores}
kable(forecast_diag %>% select(-ok), digits = 3,
      caption = "Sampling checks on every reported future-cell summary. Fresh projection noise can make forecast draws look better mixed than the fit behind them, so these are read alongside the training checks.")
gp_runs %>% group_by(sigma, lambda) %>%
  summarise(RMSE = sqrt(mean((estimate - truth)^2)), `Mean width` = mean(upper - lower),
            `Contained truth` = sprintf("%d of %d", sum(truth >= lower & truth <= upper), n()),
            .groups = "drop") %>%
  kable(digits = 4, caption = "Withheld-week effects by exposure. The fitted lengthscale is 2; the smoother settings are sensitivity analyses of the projection assumption.")
```

The projection is accurate here and its band widens with the horizon, which is the behaviour to want. Two practical notes. A narrower band is a stronger assumption, not better evidence, so report the sweep rather than the single setting that looks tightest. And four withheld weeks validate the mechanics, not a forty-week tail: when a decision rests on a long horizon, hold out a long window from an earlier launch and score the projection there before trusting it.

## When Adoption Was Not Randomized {#sec-longbet-observational}

LongBet was built for observational panels, where the argument is that conditioning on covariates and a unit intercept can stand in for a lottery. Here are two panels with the same sellers and the same effect curve, and no lottery at all.

In the first, the marketplace rolls out to its **largest, best-performing sellers first**. The rule uses catalog size, which the model sees, and the seller's own level, which the intercept carries. In the second, the field team picks on something the data never record, and that same trait also lifts the untreated path from the first launch onward.

```{r obs-fit}
#| message: false
obs_data <- list(observed = simulate_observational(ro, "observed"),
                 hidden = simulate_observational(ro, "hidden"))
obs_truth <- function(d) vapply(seq_len(S_observed), function(s) {
  m <- d$S[, match(week_study, weeks_all)] == s
  if (any(m)) mean(d$tau[, match(week_study, weeks_all)][m]) else NA_real_
}, numeric(1))
obs_results <- lapply(setNames(nm = names(obs_data)), function(sc) {
  d <- obs_data[[sc]]
  lb_artifact(paste0("obs_", sc),
              lb_key("obs", sc, lb_budget(4), LB_MODEL, digest::digest(d[c("y", "z", "x")])),
              export = c("week_study", "weeks_all", "S_observed"),
              builder = function() {
                f <- do.call(longbet::longbet,
                             c(list(y = d$y[, match(week_study, weeks_all)], x = d$x,
                                    z = d$z[, match(week_study, weeks_all)], t = week_study,
                                    random_seed = 1L), LB_MODEL, lb_budget(4)))
                a <- get_att(predict(f, x = d$x, z = d$z[, match(week_study, weeks_all)],
                                     t = week_study, summary_only = TRUE, random_seed = 1))
                list(att = a$att[seq_len(S_observed)],
                     layout = chain_layout(f$model_params$num_chains, f$model_params$num_sweeps),
                     att_full = a$att_full)
              })
})
kable(bind_rows(lapply(names(obs_results), function(sc)
  diagnose_draws(obs_results[[sc]]$att_full, obs_results[[sc]]$layout,
                 paste("Average effect:", sc))$summary)) %>% select(-ok),
  digits = 3, caption = "Sampling checks for each observational fit.")
```

```{r obs-table}
imap_dfr(obs_data, function(d, sc) {
  tr <- obs_truth(d)
  gt <- group_time_did(d$y[, match(week_study, weeks_all)], d$launch, week_study, pre_periods = 5:8)
  e_gt <- gt$estimate - tr; e_lb <- obs_results[[sc]]$att - tr
  tibble(`How launch dates were set` = c(observed = "Largest, best-performing sellers first (visible in x)",
                                         hidden = "Chosen on something x never records")[[sc]],
         `Group-time mean error` = mean(e_gt), `Group-time RMSE` = sqrt(mean(e_gt^2)),
         `LongBet mean error` = mean(e_lb), `LongBet RMSE` = sqrt(mean(e_lb^2)),
         `True effect` = mean(tr))
}) %>% kable(digits = 4,
             caption = "The same sellers and effects with the lottery removed, scored against the effect experienced by the sellers who actually adopted.")
```

The first row is the case the model is built for: adoption is driven by things it can condition on, and it recovers the effect. The second row is the boundary of what any covariate-based method can do. A flexible regression cannot condition on a variable it never sees, and neither can a differencing estimator when the unrecorded trait moves the untreated path after the pre-window. Both estimators are off, and no amount of model flexibility fixes it.

That is the practical rule for an observational panel. Write down what drove adoption, ask whether it is in the data, and if it is not, ask whether it moved the untreated trajectory. Randomizing the launch order, when you can, removes the question entirely, which is why the [first chapter](longbet.qmd) recommends it.

::: {.callout-note title="Panels with calendar structure"}
This panel's untreated path is flat over the calendar, so both forests are told not to split on calendar time. When the business has seasonality, holidays or a secular trend, turn those splits back on with `split_time_ps = TRUE` and `split_calendar_trt = TRUE`, give the chains room, and check the reported effects as usual.
:::

## Where a Unit's Own Level Comes From {#sec-longbet-baseline}

Everything the model knows about a seller arrives through $X_i$ and through the unit intercept $\gamma_i$. A within-unit difference removes each seller's level for free; LongBet reconstructs it. There are two ways to help it, and they are not equivalent.

```{r fit-variants}
#| message: false
x_lookback <- cbind(base_level = base_level, x)
v_lookback <- lb_artifact(
  "lookback_fit",
  lb_key("lookback", lb_budget(8), LB_MODEL, digest::digest(list(y_train, x_lookback, z_train))),
  export = c("y_train", "x_lookback", "z_train", "week_study"),
  builder = function() {
    f <- do.call(longbet::longbet, c(list(y = y_train, x = x_lookback, z = z_train, t = week_study,
                                          random_seed = 42L), LB_MODEL, lb_budget(8)))
    p <- predict(f, x = x_lookback, z = z_train, t = week_study, summary_only = TRUE, random_seed = 1)
    list(att = get_att(p, alpha = 0.05), resid_sd = mean(f$sigma0_draws) * f$sdy,
         unit_sd = mean(sqrt(f$sigma_gamma_draws)) * f$sdy,
         layout = chain_layout(f$model_params$num_chains, f$model_params$num_sweeps))
  })
lb_pred_sum <- predict(lb_fit, x = x, z = z_train, t = week_study, summary_only = TRUE, random_seed = 1)
lb_att_main <- get_att(lb_pred_sum, alpha = 0.05)
variants <- list(
  list(label = "Structural covariates; intercept (main fit)", att = lb_att_main,
       resid_sd = mean(lb_fit$sigma0_draws) * lb_fit$sdy,
       unit_sd = mean(sqrt(lb_fit$sigma_gamma_draws)) * lb_fit$sdy,
       layout = chain_layout(lb_fit$model_params$num_chains, lb_fit$model_params$num_sweeps)),
  list(label = "+ lookback level; intercept", att = v_lookback$att, resid_sd = v_lookback$resid_sd,
       unit_sd = v_lookback$unit_sd, layout = v_lookback$layout))
kable(bind_rows(lapply(variants, function(v)
  diagnose_draws(v$att$att_full, v$layout, v$label)$summary)) %>% select(-ok),
  digits = 3, caption = "Sampling checks for both ways of telling the model about a seller's level.")
map_dfr(variants, function(v) tibble(
  Covariates = v$label, `Residual SD` = v$resid_sd, `Intercept SD (posterior mean)` = v$unit_sd,
  `Effect RMSE` = sqrt(mean((v$att$att[seq_len(S_observed)] - truth_att)^2)),
  `Interval width` = mean(v$att$intervals[2, seq_len(S_observed)] - v$att$intervals[1, seq_len(S_observed)]))) %>%
  kable(digits = 4, caption = sprintf("The true residual SD is %.2f and the seller-specific component the covariates never record has SD 0.35.", sd_eps))
```

Both recover the effect. What differs is the division of labour: with the lookback level in `x`, the forest explains part of each seller's level and the intercept shrinks to match. The recommendation for a panel like this one is to keep the intercept, which is a parameter per unit with a conjugate update, and to add a lookback summary only when it carries something the intercept cannot, such as a pre-period **trend**.

One rule is not optional: a lookback summary must be built from periods that are **not** part of the modelled panel. Weeks that appear both as a covariate and as an outcome make the prognostic forest partly a function of its own target.

### The error term is independent by assumption {#sec-longbet-serial}

$\epsilon_{it}$ is independent across units and periods. The cheapest check is on holdout sellers, whose outcomes the treatment never touched: take the prognostic surface without the intercept and ask whether the residuals still carry unit structure.

```{r residual-acf}
gamma_hat <- rowMeans(lb_fit$gamma_draws) * lb_fit$sdy
resid_h <- (y_train - (lb_pred_sum$muhats0.mean - gamma_hat))[is_hold, ]
lag1 <- mean(mapply(function(a, b) cor(resid_h[, a], resid_h[, b]),
                    1:(ncol(resid_h) - 1), 2:ncol(resid_h)))
se_iid <- sd(as.vector(resid_h)) / sqrt(length(resid_h))
se_block <- sd(rowMeans(resid_h)) / sqrt(nrow(resid_h))
cat("Mean lag-1 correlation of forest-only residuals among holdout sellers:", round(lag1, 3), "\n",
    "Design effect for a grand mean (block versus independent):", round((se_block / se_iid)^2, 2), "\n")
```

The generated innovations are independent, so what this measures is the seller-level component the covariates do not explain, which is exactly what the intercept absorbs. Run it on your own panel with the intercept off: a large number says the panel needs the intercept, which the main fit has.

## Covariates That Change Over the Panel {#sec-longbet-timevarying}

Everything in `x` is one number per seller. `x_tv` and `x_trt_tv` take `[N, T, P]` arrays instead, for the prognostic and treatment forests, so a split can put different weeks of the same seller in different leaves.

The mechanics are easy and the causal question is not. Before a moving variable goes into the treatment forest, ask when it was determined relative to the exposure being scored, and whether the treatment could have changed it. A variable the treatment moves is a consequence, and conditioning on it can block part of the effect you are trying to measure. The covariate here is a **promotional calendar set by the marketplace**: it is not something the optimizer can influence, which is what makes it a legitimate moderator.

Promotions lift GMV on their own and make the optimizer worth more in the same week, so the covariate belongs in both forests.

```{r promo-fit}
#| message: false
pm <- simulate_promo(ro)
promo_arr <- array(pm$promo, c(dim(pm$promo), 1))
promo_fit <- function(tag, label, tv) {
  force(tv); force(label)          # both are used only inside the builder, which runs in a worker
  lb_artifact(paste0("promo_", tag),
              lb_key("promo", tag, lb_budget(4), LB_MODEL, digest::digest(list(pm$y, pm$promo))),
              export = c("pm", "x", "z_train", "week_study"),
              builder = function() {
                f <- do.call(longbet::longbet,
                             c(list(y = pm$y, x = x, z = z_train, t = week_study,
                                    x_tv = tv, x_trt_tv = tv, random_seed = 515L),
                               LB_MODEL, lb_budget(4)))
                p <- predict(f, x = x, z = z_train, t = week_study, x_tv = tv, x_trt_tv = tv,
                             random_seed = 1)
                dm <- matrix(p$tauhats, nrow = length(z_train))[as.vector(pm$treated), ]
                pc <- pm$promo[pm$treated] - mean(pm$promo[pm$treated])
                slope_draws <- as.numeric(crossprod(pc, dm)) / sum(pc^2)
                bin_draws <- t(vapply(seq_len(10), function(b)
                  colMeans(dm[pm$bins == b, , drop = FALSE]), numeric(ncol(dm))))
                list(sd = mean(f$sigma0_draws) * f$sdy, catt = p$tauhats.mean,
                     slope = mean(slope_draws),
                     diagnostics = diagnose_draws(rbind(get_att(p)$att_full, slope_draws,
                                                        as.vector(f$sigma0_draws) * f$sdy, bin_draws),
                                                  f, label)$summary)
              })
}
promo_blind <- promo_fit("blind", "Promotion omitted", NULL)
promo_aware <- promo_fit("aware", "Promotion included", promo_arr)
kable(bind_rows(promo_blind$diagnostics, promo_aware$diagnostics) %>% select(-ok), digits = 3,
      caption = "Checks covering every average effect, the moderation slope, the residual SD and the ten plotted bin means.")
```

```{r fig-promo}
#| fig-cap: "Effect on the treated against the promotional intensity of the same week, as means within ten intensity bins."
#| fig-alt: "Three lines rising with promotional intensity. The truth and the model given the covariate rise together; the model without it is nearly flat."
#| fig-width: 8.5
#| fig-height: 4
tibble(promo = rep(pm$promo[pm$treated], 3), bin = rep(pm$bins, 3),
       catt = c(pm$tau_promo[pm$treated], promo_aware$catt[pm$treated], promo_blind$catt[pm$treated]),
       which = rep(c("Truth", "With the covariate", "Without it"), each = sum(pm$treated))) %>%
  group_by(which, bin) %>% summarise(promo = mean(promo), catt = mean(catt), .groups = "drop") %>%
  ggplot(aes(promo, catt, colour = which)) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.6) +
  scale_colour_manual(values = c(Truth = "#000000", `With the covariate` = PAL[["LongBet"]],
                                 `Without it` = PAL[["Group-time DiD"]])) +
  labs(x = "Promotional intensity that week", y = "Effect on log weekly GMV", colour = NULL) +
  theme(legend.position = "bottom")
```

```{r promo-table}
slope_of <- function(v) coef(lm(v[pm$treated] ~ pm$promo[pm$treated]))[2]
tibble(Model = c("Without the covariate", "With the covariate", "Truth"),
       `Residual SD` = c(promo_blind$sd, promo_aware$sd, sd_eps),
       `Slope of effect in promotion` = c(slope_of(promo_blind$catt), slope_of(promo_aware$catt),
                                          slope_of(pm$tau_promo))) %>%
  kable(digits = 3, caption = "A continuous covariate that varies within a seller over time, given to both forests or to neither.")
```

Given the covariate, the model recovers both channels: the residual scale drops to its true value because promotions no longer look like noise, and the effect's dependence on promotional intensity comes out close to the truth. Without it, the week-to-week moderation is invisible and the residual absorbs it.

Three practical points. Give the covariate to whichever forest needs it, prognostic for a cleaner baseline, treatment for moderation, both when both. It must be observed in every cell, including cells where the outcome is missing, because it is an input rather than an outcome. And it must be passed to `predict()` as well, since the fitted trees record which axis each split used.

## Conclusion

The trajectory extends past the study window because it is a smooth function of exposure rather than a set of unrelated period effects. Adoption that was chosen rather than randomized is fine when the drivers are recorded, and beyond any method's reach when they are not. A unit intercept is the cheapest way to give the model each unit's level, and time-varying covariates open the door to within-unit moderation as long as the treatment cannot move them.

The **[next chapter](longbet_ordinal.qmd)** changes the outcome from a continuous measure to an ordered rating; the **[last one](longbet_encouragement.qmd)** changes the design to one where adoption can only be encouraged.
