---
title: "LongBet: Ordinal Outcomes and Customer Experience"
description: "Dynamic effects on ordered ratings: heterogeneous category probabilities, rare-rating comparisons, and Bayesian rollout decisions."
format:
  html:
    css: longbet.css
execute:
  error: false
  warning: true
  cache: false
share:
  permalink: "https://book.martinez.fyi/longbet_ordinal.html"
  description: "Ordinal LongBet for customer-experience decisions"
  linkedin: true
  email: true
  mastodon: true
---

```{r ordinal-setup}
#| include: false
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
})
source("R/longbet-ordinal.R")
ordinal_example <- read_ordinal_example()
panels <- ordinal_example$panels
decision_draws <- ordinal_example$draws
provenance <- ordinal_example$provenance
theme_set(theme_minimal(base_size = 12))
```

## A rollout can improve more than the average rating

A SaaS company rolls out a new support workflow across customer accounts.
Each account reports a weekly satisfaction rating from one to five stars.
The product team wants to know whether the workflow creates more enthusiastic
customers, reduces very poor experiences, and helps different kinds of
accounts at different stages of adoption.

An average rating compresses those questions into one score. A five-star
indicator answers the enthusiasm question, but treats one and four stars
alike. **Ordinal LongBet models the whole ordered distribution**, then lets
the team choose the category effects that matter for its decision.

This chapter builds on [LongBet's panel model](longbet.qmd). The
[earlier ordinal BART chapter](phobart.qmd) introduces ordered outcomes in a
cross-sectional setting; here, calendar time and time since adoption are
separate inputs. The ordered-probit extension is implemented in
`longbet-jax` [@longbetpkg], building on the original LongBet framework
[@wang2024longbet].

## The model: a latent response, observed through ordered thresholds

Code the five ratings as $Y_{it}\in\{0,1,2,3,4\}$, with 0 representing one
star and 4 representing five stars. For account $i$ in week $t$, write

$$
\begin{aligned}
Y^*_{it} &= \eta_{it}+\epsilon_{it},
&\epsilon_{it}&\sim N(0,1),\\
\eta_{it} &= \mu(X_i,t)
 +b_{Z_{it}}\beta_{S_{it}}\nu(X_i,S_{it},t)+\gamma_i,\\
Y_{it}=k &\quad\Longleftrightarrow\quad
\theta_k<Y^*_{it}\leq\theta_{k+1}.
\end{aligned}
$$

Here $X_i$ contains account characteristics measured before treatment, $Z_{it}$
records adoption, and $S_{it}$ counts exposure weeks, starting at 1 after
adoption and remaining 0 before it. Treatment stays on once adopted.
The prognostic forest $\mu$ learns baseline patterns; the treatment forest
$\nu$ learns how responses vary with account characteristics and time.
The Gaussian process $\beta_S$ pools information across exposure lengths.
An optional $\gamma_i\sim N(0,\sigma_\gamma^2)$ captures persistent account
differences. The prognostic scaling parameter is fixed at 1 here.

The latent scale is a modeling device, not a satisfaction score with business
units. To identify it, the package fixes the residual variance to 1 and the
first finite threshold to 0:

$$
\theta_0=-\infty,\qquad
\theta_1=0<\theta_2<\cdots<\theta_{K-1},\qquad
\theta_K=\infty.
$$

Tree priors shrink the fitted surfaces, the GP prior regularizes the exposure
factor, and the remaining thresholds have a proper ordered-normal prior
(default scale 5). Thresholds and residual scale are shared across accounts,
weeks, and treatment. This avoids assigning equal distances to rating labels,
while imposing a common latent location-shift structure.

### Convert each posterior draw into probabilities

Compare adoption with never adopting at the **same account profile and calendar
week**. For a treated cell, the two latent locations are

$$
\begin{aligned}
\eta^1_{it}&=\mu(X_i,t)+b_1\beta_{S_{it}}\nu(X_i,S_{it},t)+\gamma_i,\\
\eta^0_{it}&=\mu(X_i,t)+b_0\beta_0\nu(X_i,0,t)+\gamma_i.
\end{aligned}
$$

The untreated contribution uses exposure zero in both the GP and the forest.
For posterior draw $d$, the category probabilities and their effects are

$$
\begin{aligned}
p^{a,(d)}_{itk}
&=\Phi\!\left(\theta^{(d)}_{k+1}-\eta^{a,(d)}_{it}\right)
 -\Phi\!\left(\theta^{(d)}_k-\eta^{a,(d)}_{it}\right),\\
\Delta^{(d)}_{itk}&=p^{1,(d)}_{itk}-p^{0,(d)}_{itk},
\qquad a\in\{0,1\}.
\end{aligned}
$$

$\Phi$ is the standard normal CDF. The five category effects sum to zero:
gains in some ratings must come from losses elsewhere. Calculate the CDF and
the contrast **inside each draw**, then summarize. Transforming an average
latent effect gives a different answer.

For exposure $s$, average these contrasts over the treated cells that reached
that exposure, denoted $\mathcal I_s$:

$$
\operatorname{ATT}^{(d)}_k(s)=
\frac{1}{|\mathcal I_s|}\sum_{(i,t)\in\mathcal I_s}\Delta^{(d)}_{itk}.
$$

The resulting draws describe uncertainty about the average probability effect.
Different exposure groups may contain different accounts. In these examples,
randomized adoption supplies identification; an observational application
still needs credible adjustment, overlap, and a defensible treatment-timing
assumption.

## Heterogeneity changes which customer outcome to target

There are two sources of heterogeneity. The treatment forest can learn
different latent responses for different account profiles. **Even the same
latent response produces different probability effects at different starting
levels of satisfaction.** Moving an account away from a one-star experience
is different from moving an already satisfied account into five stars.

For a numerical illustration, use thresholds 0, 0.7, 1.4, and 2.0. Compare
two profiles with untreated latent locations $-0.6$ and $0.5$, giving both
the same $+0.6$ latent treatment shift. These are specified model values,
not fitted estimates or evidence about comparative accuracy.

```{r}
#| echo: false
#| label: fig-ordinal-profiles
#| fig-width: 6
#| fig-height: 3.6
#| fig-alt: "For an equal latent improvement, the low-satisfaction profile gains more in avoiding one-star ratings; the high-satisfaction profile gains more in five-star ratings."
#| fig-cap: "The same latent improvement can imply different customer-experience priorities. Values are calculated directly from the ordered-probit model."
profiles <- tibble::tibble(
  profile = c("Lower starting\nsatisfaction", "Higher starting\nsatisfaction"),
  eta0 = c(-.6, .5), eta1 = eta0 + .6,
  `Fewer one-star ratings` = pnorm(-eta0) - pnorm(-eta1),
  `More five-star ratings` = pnorm(eta1 - 2) - pnorm(eta0 - 2)) %>%
  pivot_longer(c(`Fewer one-star ratings`, `More five-star ratings`),
               names_to = "goal", values_to = "gain")
profiles$profile <- factor(profiles$profile,
  levels = c("Lower starting\nsatisfaction", "Higher starting\nsatisfaction"))
ggplot(profiles, aes(profile, 100*gain, fill = profile)) +
  geom_col(width = .6, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f pp", 100*gain)), vjust = -.4, size = 3.5) +
  facet_wrap(~goal) + scale_fill_manual(values = c("#0072B2", "#D55E00")) +
  scale_y_continuous(limits = c(0, 26), expand = expansion(mult = c(0, .02))) +
  labs(x = NULL, y = "Probability improvement (pp)")
```

A customer-success team trying to prevent poor experiences and a growth team
trying to create advocates could prioritize different profiles. A latent-effect
ranking alone misses that distinction. LongBet supplies conditional effects
at account profiles; it does not identify whether a particular person's
unobserved individual rating would improve.

## When keeping the lower ratings improves estimation

Consider a support workflow where five-star ratings are uncommon. We compare
methods on **20 independently simulated rollouts**, each with 240 training
accounts, eight weeks, and five rating levels. Adoption is randomized across
weeks 3, 4, 5, and never. Predictions are evaluated on another 240 independent
accounts within the observed calendar and exposure range. The highest rating
appears in only **`r sprintf("%.1f%%", 100*provenance$mean_training_top_share)`**
of training observations on average.

The generating model has a smooth baseline, a saturating exposure response,
a covariate-dependent treatment effect, and a shared residual scale. There
are no persistent account random effects in this comparison. Every method
uses the same inputs; held-out accounts and known simulation effects never
enter fitting or tuning.

The main question is whether modeling all five ratings improves the estimate
of the **five-star probability ATT**, relative to binary LongBet fitted only
to the five-star indicator. The table also reports error in the conditional
effects for all categories, to check the richer output used for heterogeneous
customer-experience analysis.

```{r ordinal-comparison}
#| echo: false
method_names <- c(ordinal_longbet = "Ordinal LongBet",
                  binary_longbet = "Binary LongBet",
                  ordinal_spline = "Ordinal spline regression",
                  boosted_tree = "Tuned boosted trees")
comparison <- panels %>%
  filter(scenario == "rare_top", method %in% names(method_names)) %>%
  group_by(method) %>%
  summarize(top_rmse = 100*sqrt(mean(top_att_mse)),
            category_rmse = 100*sqrt(mean(category_cate_mse)),
            seconds = mean(fit_seconds + prediction_seconds), .groups = "drop") %>%
  arrange(match(method, names(method_names)))
kable(comparison %>% transmute(
  Model = unname(method_names[method]),
  `Five-star ATT RMSE (pp)` = sprintf("%.2f", top_rmse),
  `Category effect RMSE (pp)` = ifelse(is.na(category_rmse), "—", sprintf("%.2f", category_rmse))),
  caption = "Error against known probability effects; lower is better. Each panel has equal weight. Binary LongBet estimates only the five-star effect.")
dispersion_rmse <- panels %>% filter(scenario == "dispersion") %>%
  group_by(method) %>% summarize(rmse = 100*sqrt(mean(top_att_mse)), .groups = "drop")
dispersion_error <- setNames(dispersion_rmse$rmse, dispersion_rmse$method)
dispersion_failures <- sum(panels$scenario == "dispersion" &
  panels$method == "ordinal_longbet" & panels$diagnostic_failures > 0, na.rm = TRUE)
```

Ordinal LongBet had **`r sprintf("%.1f%%", 100*provenance$primary$reduction)`
lower mean squared error** than binary LongBet for the five-star ATT and won
on **`r provenance$primary$wins` of 20 panels**. Lower ratings contain
information about the shared latent response that dichotomization discards.
This is a useful advantage even when the final business metric is binary.

The prespecified paired comparison's 97.5% interval for binary minus ordinal
MSE was **`r sprintf("[%.2f, %.2f]", 10000*provenance$primary$interval[1], 10000*provenance$primary$interval[2])`**
squared percentage points, favoring ordinal LongBet. It uses resampling of
the 20 panels; its tails are approximate. All 40 ordinal/binary fits pass
the monitored ATT and threshold diagnostics after the specified extensions.

The other comparisons are descriptive. The spline baseline includes nonlinear
covariate terms, week and exposure effects, and treatment interactions;
the boosted classifier is tuned by cross-validation grouped by training
account. Ordinal LongBet also had lower conditional category-effect error
here, but **this does not establish a general nonlinear modeling advantage**.
A harder nonlinear pilot did not mix adequately and provides no such evidence.

Simple ordinal regression is a strong practical alternative: its five-star
error was close, and its fit and prediction took under a second per panel,
versus about six minutes for ordinal LongBet on this CPU run. The information
gain from using all categories is available to both. The trade-off is
LongBet's flexible panel structure and posterior decision analysis against
additional computation and sampling checks.

## From rating effects to a rollout probability

There are two distinct probabilities in the analysis:

- **Outcome probability:** how likely is a five-star rating under a given
  treatment history and account profile?
- **Posterior decision probability:** given the data and model, how probable
  is it that the treatment effect meets the team's target?

Frequentist ordinal regression and flexible classifiers can estimate outcome
probabilities and heterogeneous effects too. The Bayesian advantage here is
a posterior over those effects: the same paired draws support intervals,
business thresholds, and joint criteria while preserving their dependence.
A confidence level from a frequentist interval is not that posterior
probability. [Ordinal regression](https://cran.r-project.org/web/packages/ordinal/vignettes/clm_article.pdf)
remains a useful baseline rather than a method that cannot handle probabilities.

For a worked decision, take the **first prespecified panel**, seed 91000,
at exposure week 4. Suppose the team requires at least a 3-percentage-point
increase in five-star probability and a 20-point reduction in one-star
probability. These thresholds are illustrative business choices, not targets
selected or validated by the simulation. The question is

$$
\Pr\!\left\{
\operatorname{ATT}_4(4)\geq0.03\ \text{and}\
\operatorname{ATT}_0(4)\leq-0.20
\mid\mathcal D
\right\}.
$$

Here the subscript indexes the rating code; the argument is the exposure week.
The calculation is short:

```{r ordinal-decision}
# These columns contain paired category ATT draws for exposure 4.
passes <- with(decision_draws, top_att >= 0.03 & lowest_att <= -0.20)
probability_meets_targets <- mean(passes)
probability_meets_targets
```

```{r ordinal-decision-check}
#| include: false
event_chains <- matrix(as.numeric(passes), nrow = 5000, ncol = 4)
event_rhat <- posterior::rhat(event_chains)
# Diagnose the indicator's mean, the posterior probability being reported.
# Quantile-tail ESS is undefined when a Bernoulli tail threshold is constant.
event_ess <- posterior::ess_basic(event_chains)
event_mcse <- posterior::mcse_mean(event_chains)
stopifnot(is.finite(event_rhat), event_rhat <= 1.01,
          is.finite(event_ess), event_ess >= 400,
          all(apply(event_chains, 2, function(x) length(unique(x))) == 2))
```

The posterior probability of meeting **both** targets is
**`r sprintf("%.1f%%", 100*probability_meets_targets)`**. This concerns the
average probability effects for the specified account population and exposure,
not the chance that every customer improves. Check both conditions inside
the same draw; multiplying their marginal probabilities would assume away
their posterior dependence. The event itself passes the sampling checks
($\widehat R=$ `r sprintf("%.3f", event_rhat)`, effective sample size
`r format(round(event_ess), big.mark = ",")`); its Monte Carlo standard error
is about `r sprintf("%.1f", 100*event_mcse)` percentage points.

For a customer segment, average its account-level probability contrasts
within each draw first, then apply the same business rule. Segment definitions,
exposure support, rollout costs, and the consequences of a wrong decision
still matter. A high posterior probability under the fitted model does not
by itself make deployment worthwhile.

### A minimal fitting workflow

Use a revision of `longbet-jax` that includes the ordinal R API. In R, `y`
is an account-by-week matrix coded 0–4, `z` is the matching adoption
matrix, and `x` contains baseline account features. This recipe uses the
comparison's settings; the chapter's numerical results above come from the
saved, checked experiments.

```r
library(longbet)
fit <- longbet(
  y = y, x = x, z = z, t = seq_len(ncol(y)),
  outcome = "ordinal", num_categories = 5L,
  num_chains = 4L, num_burnin = 10000L,
  num_sweeps = 5000L, n_skip = 4L,
  num_trees_pr = 20L, num_trees_trt = 20L,
  max_depth_pr = 5L, max_depth_trt = 5L,
  b_scaling = FALSE, split_time_trt = FALSE,
  random_intercept = FALSE, device = "cpu", random_seed = 391000L)
pred <- predict(fit, x = x, z = z, summary_only = TRUE)
category_att <- att_probabilities(pred)
# [exposure, category, draw]; category position 5 is the five-star rating.
top_att_draws <- category_att$att_full[, 5, ]
```

Each chain retains 5,000 draws after 20,000 sampling iterations, following
10,000 burn-in iterations. Longer sampling may be needed: diagnose the
effects and business events you report. Fixed coding sets $b_0=b_1=1$;
`split_time_trt = FALSE` removes exposure splits while retaining calendar
time in the treatment forest. These choices change the prior. The comparison
disables account intercepts because its simulated errors have no persistent
account component; that choice needs reconsideration for real panels.

## When to use a different model

An ordinal location shift cannot describe every change in customer experience.
If treatment makes both one-star and five-star ratings more likely **within
the same account profile**, a shared-scale shift is inadequate. In a separate
20-panel dispersion simulation, five-star ATT RMSE was
**`r sprintf("%.2f", dispersion_error[["ordinal_longbet"]])` points** for
ordinal LongBet, **`r sprintf("%.2f", dispersion_error[["ordinal_scale"]])`**
for ordinal regression with an exposure-dependent scale, and
**`r sprintf("%.2f", dispersion_error[["boosted_tree"]])`** for boosted trees.
`r dispersion_failures` LongBet fits retained sampling
failures; these results are a warning about both model fit and computation.

Use ordinal LongBet when the whole rating distribution matters, shared
thresholds and scale are credible, and the panel's heterogeneous trajectories
justify a flexible model. Include ordinal regression as a fast baseline.
If the goal is explicitly the mean of a chosen numeric score, continuous
regression can be appropriate. If only a common binary threshold matters,
a binary model can also be sufficient. The rare-rating comparison establishes
a specific gain from retaining information, not a universal ranking of methods.

<details><summary>Data and computation</summary>

The chapter rebuilds its numerical summaries from
[panel-level comparison results](data/longbet-ordinal/panel-results.csv) and
[paired decision draws](data/longbet-ordinal/decision-draws.csv.gz), checking
their recorded hashes. The [method notes](data/longbet-ordinal/README.md)
give the generating mechanisms, model settings, diagnostic rules, and
reproduction instructions; [provenance](data/longbet-ordinal/provenance.json)
records the source and software versions. These are simulated business
examples, not results from a production rollout.

</details>
