---
title: "LongBet: Decisions and Multiple Outcomes"
description: "Translating dynamic treatment effects into business decisions: economic valuation scenarios, capacity-constrained targeting, and joint multi-outcome modeling."
format:
  html:
    css: longbet.css
share:
  permalink: "https://book.martinez.fyi/longbet_decisions.html"
  description: "LongBet: Decisions and multiple outcomes in business causal inference"
  linkedin: true
  email: true
  mastodon: true
---

::: {.callout-note title="Continuation of the LongBet rollout study"}
This chapter continues the randomized marketplace rollout introduced in **[LongBet: Dynamic Treatment Effects in Staggered Rollouts](longbet.qmd)**. It reuses the same seller population ($N=3,000$), calendar, log weekly GMV outcome scale, and fitted model objects to examine economic decision rules and multi-outcome trade-offs. Advanced extensions, including GP forecasting and observational panel identification, are covered in **[LongBet: Forecasting, Observational Panels, and Diagnostics](longbet_extensions.qmd)**.
:::
```{r setup}
#| message: false
#| warning: false
library(longbet)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(purrr)
library(knitr)

source("R/longbet-common.R")
source("R/longbet-rollout.R")
source("R/longbet-artifacts.R")
verify_longbet_environment()

lb_engine_fingerprint <- get_longbet_engine_fingerprint()
lb_chapter_fingerprint <- digest::digest(file = knitr::current_input(), algo = "sha256")
lb_dependencies <- get_longbet_dependencies()
knitr::opts_chunk$set(cache.extra = list(lb_engine_fingerprint, lb_chapter_fingerprint,
                                       lb_dependencies))

rollout <- simulate_longbet_rollout()
unpack_rollout(rollout)

lb_fit <- get_or_create_core_fit(rollout)
cl <- get_or_create_common_launch(lb_fit, rollout)
S_target      <- cl$S_target
tau_hat_draws <- cl$tau_hat_draws
tau_hat       <- cl$tau_hat
tau_truth_S   <- cl$tau_truth_S
initial_lift  <- cl$initial_lift
gmv_week      <- cl$gmv_week
mu0_hat       <- cl$mu0_hat
sigma2_hat    <- cl$sigma2_hat
```
## The Decision {#sec-longbet-decision}

The marketplace keeps 12% of GMV and the optimizer costs \$500 per seller per
year. An effect estimate becomes a dollar valuation only after choosing a
baseline revenue path, a treatment trajectory, and a decision rule. Each of
those choices needs to be visible.

Here the baseline is a **fixed estimate** of untreated weekly revenue over the
last four study weeks. It uses the model's fitted untreated log outcome,
including the unit intercept, with plug-in median parameter estimates for the
log-normal conversion. It uses no future observations or simulator parameters.
We hold that baseline constant over the decision year. This makes the analysis
conditional on a baseline estimate; its intervals do not include baseline
estimation error or uncertainty about future revenue growth.

The first scenario also holds the effect at its week-14 value for the remaining
38 weeks of the year. For the first 14 weeks, it uses the all-seller adoption
trajectory used in @fig-catt-listings. These are 52-week **scenarios**, not forecasts
of realized annual revenue. Later in this section we vary both assumptions.

A log effect converts to proportional lift as `expm1(tau)`. The residual
variance cancels from the treated-to-untreated ratio under the model's common
log-normal error distribution. We apply that conversion to each effect draw
before valuing a seller. For monetary summaries we report medians and quantiles:
a proper posterior does not by itself guarantee a finite mean after
exponentiation. In particular, integrating a log-normal dollar mean over an
inverse-gamma variance posterior need not have a finite expectation. Fixing
the baseline avoids claiming that the following is such an integrated dollar
forecast.

```{r economics}
TAKE_RATE   <- 0.12
ANNUAL_COST <- 500
HOLD_WEEKS  <- 52 - S_target

annual_lift <- initial_lift + HOLD_WEEKS * expm1(tau_hat_draws)
net_draws <- TAKE_RATE * gmv_week * annual_lift - ANNUAL_COST
stopifnot(all(is.finite(net_draws)), all(gmv_week > 0))
median_net <- apply(net_draws, 1, median)
p_worth <- rowMeans(net_draws > 0)

# Truth is used only for scoring, never for selecting an operational policy.
recent <- tail(seq_along(week_study), 4)
recent_calendar <- week_study[recent]
gmv_week_true <- rowMeans(exp(mu0_true[, recent_calendar, drop = FALSE] + 0.28^2 / 2))
true_initial_lift <- rowSums(expm1(outer(w_i, seq_len(S_target), function(a, s)
  a * h_grow(s) + (1 - a) * h_fade(s))))
true_scenario_lift <- true_initial_lift + HOLD_WEEKS * expm1(tau_truth_S)
net_true <- TAKE_RATE * gmv_week_true * true_scenario_lift - ANNUAL_COST

breakeven <- ANNUAL_COST / (TAKE_RATE * 52 * gmv_week)
tibble(
  Quantity = c("Median estimated weekly baseline GMV",
               "Median DGP-known weekly baseline GMV (scoring only)",
               "Median break-even annual proportional lift"),
  Value = c(scales::dollar(median(gmv_week)),
            scales::dollar(median(gmv_week_true)),
            scales::percent(median(breakeven), 0.1))
) %>% kable(caption = "The baseline estimate and the separate simulator benchmark.")
```

The DGP-known baseline is the simulator's expected untreated revenue over
those same four weeks. It is available to us for validation, not to the
marketplace when choosing sellers. The score below uses that baseline and the
true treatment effects, while imposing the same flat-baseline and flat-tail
scenario. It therefore measures errors in both baseline and effect estimation
within that scenario. It is not a realized cash-flow observation.

```{r policy-table}
# Ranking by the posterior median is a transparent quantile-based rule.
# It is not claimed to maximize posterior expected total profit.
by_value <- order(median_net, decreasing = TRUE)
top_k <- function(K) {
  keep <- rep(FALSE, n)
  eligible <- by_value[median_net[by_value] > 0]
  keep[head(eligible, min(K, length(eligible)))] <- TRUE
  keep
}
policies <- list(
  "Do not roll out" = rep(FALSE, n),
  "Roll out to everyone" = rep(TRUE, n),
  "Broad catalogs (>= 50 listings)" = broad,
  "Up to 450 by positive median net value" = top_k(450),
  "Up to 900 by positive median net value" = top_k(900),
  "Up to 1350 by positive median net value" = top_k(1350),
  "Positive median net value (uncapped)" = median_net > 0,
  "P(net value > 0) >= 0.90" = p_worth >= 0.90,
  "Oracle for the stated scenario (scoring only)" = net_true > 0
)
policy_value <- function(keep, label) {
  total <- colSums(net_draws[keep, , drop = FALSE])
  tibble(
    Policy = label,
    Sellers = sum(keep),
    `Conditional median (k/yr)` = median(total) / 1000,
    `Conditional 95% interval` = sprintf(
      "[%.0f, %.0f]", quantile(total, 0.025) / 1000,
      quantile(total, 0.975) / 1000
    ),
    `DGP-known scenario value (k/yr)` = sum(net_true[keep]) / 1000
  )
}
policy_results <- imap_dfr(policies, policy_value)
policy_results %>% kable(
  digits = 0,
  caption = paste(
    "Annual scenario values in thousands of US dollars.",
    "Conditional intervals hold the estimated baseline and tail assumption fixed.",
    "The DGP-known score uses true baselines and effects under the same scenario."
  )
)
```

The table preserves posterior dependence by summing sellers within each draw
before taking an interval. A total's median need not equal the sum of its
sellers' medians, so ranking by seller medians is a stated heuristic rather
than a theorem about the best aggregate policy. The oracle maximizes the known
scenario value because that score is additive; it is a validation benchmark,
not a deployable rule.

The probability threshold expresses a different preference from the capacity
ranking. Under a binary loss with constant cost for a wrong enable and constant
cost for a missed profitable seller, a 0.90 threshold corresponds to assigning
nine times as much loss to the former. That loss model ignores the magnitude
of seller profits and losses. Choosing it requires a business judgment, and
the probabilities here remain conditional on the estimated baseline and the
annual scenario.

```{r policy-diagnostics}
# The pooled ATT can mix well while seller-level valuations do not.
seller_value_diagnostics <- diagnose_draws(
  net_draws, lb_fit, "Seller-level conditional annual net values"
)
nonempty_policies <- policies[vapply(policies, any, logical(1))]
policy_total_draws <- do.call(rbind, lapply(nonempty_policies, function(keep)
  colSums(net_draws[keep, , drop = FALSE])))
policy_value_diagnostics <- diagnose_draws(
  policy_total_draws, lb_fit, "Conditional annual policy totals"
)
diagnostic_rows <- bind_rows(
  seller_value_diagnostics$summary, policy_value_diagnostics$summary
)

# Constant indicators contain no observed transitions; agreement at 0 or 1
# does not establish that a rare event has been adequately sampled.
varying_probability <- p_worth > 0 & p_worth < 1
if (any(varying_probability)) {
  probability_diagnostics <- diagnose_draws(
    1.0 * (net_draws[varying_probability, , drop = FALSE] > 0),
    lb_fit, "Nonconstant seller profitability indicators"
  )
  diagnostic_rows <- bind_rows(diagnostic_rows, probability_diagnostics$summary)
}
diagnostic_rows %>% kable(
  digits = 3, caption = "Sampling diagnostics for the quantities used in the decision."
)
cat(sum(!varying_probability),
    "seller profitability indicators were constant across retained draws.",
    "Their rare-event probabilities are not certified by that agreement.\n")
decision_sampling_ok <- all(diagnostic_rows$diagnostic_ok)
```

The reported decision diagnostics
`r if (decision_sampling_ok) "meet the stated thresholds for the checked targets" else "fail at least one stated threshold"`.
`r if (decision_sampling_ok) "These checks support the numerical stability of those summaries; they do not validate the baseline or annual tail assumptions." else "Treat the ranking, intervals, and probability rule as an unresolved illustration until further sampling establishes stability."`
A probability shown as zero or one from finite draws is not proof that an
event is impossible or certain.

```{r effect-vs-value}
tibble(
  Segment = c("Narrow catalog (< 50 listings)", "Broad catalog (>= 50 listings)"),
  Sellers = c(sum(!broad), sum(broad)),
  `Median true lift after 14 weeks` = scales::percent(
    c(median(expm1(tau_truth_S[!broad])), median(expm1(tau_truth_S[broad]))), 0.1),
  `Positive DGP-known scenario value` = scales::percent(
    c(mean(net_true[!broad] > 0), mean(net_true[broad] > 0)), 0.1)
) %>% kable(caption = paste(
  "Effect size and dollar value answer different questions.",
  "Scenario value uses the true baseline and the stated 38-week flat tail."
))
```

A catalog rule selects on the effect surface, whereas an economic rule also
depends on seller size and the flat \$500 cost. The table quantifies that
distinction under the stated scenario. It does not establish that every broad
catalog should be enabled or that every narrow catalog should be excluded.
Likewise, any gap between the estimated and DGP-known columns is an error for
this realization. Baseline estimation, effect estimation, nonlinear
transformation, and selection can all contribute; this table cannot identify
which mechanism caused the gap.

```{r decision-sensitivity}
# Hold the chosen policy fixed so that the table isolates assumptions rather
# than changing both assumptions and sellers at once.
decay_tail <- matrix(0, nrow(net_draws), ncol(net_draws))
for (k in seq_len(HOLD_WEEKS)) {
  decay_tail <- decay_tail + expm1(tau_hat_draws * 2^(-k / 13))
}
lift_scenarios <- list(
  "No effect after week 14" = initial_lift,
  "Log effect halves every 13 weeks" = initial_lift + decay_tail,
  "Flat log effect after week 14" = annual_lift
)
sensitivity_keep <- top_k(450)
decision_sensitivity <- imap_dfr(lift_scenarios, function(lift, label) {
  map_dfr(c(0.8, 1.0, 1.2), function(baseline_factor) {
    total <- TAKE_RATE * colSums(
      gmv_week[sensitivity_keep] *
        lift[sensitivity_keep, , drop = FALSE]
    ) * baseline_factor - ANNUAL_COST * sum(sensitivity_keep)
    tibble(
      `Effect scenario` = label,
      `Baseline multiplier` = baseline_factor,
      `Conditional median (k/yr)` = median(total) / 1000,
      `Conditional 95% interval` = sprintf(
        "[%.0f, %.0f]", quantile(total, 0.025) / 1000,
        quantile(total, 0.975) / 1000
      )
    )
  })
})
decision_sensitivity %>% kable(
  digits = 1,
  caption = paste(
    "The same", sum(sensitivity_keep),
    "sellers selected by the capacity rule, under nine valuation assumptions.",
    "Baseline multipliers are sensitivity scenarios, not a posterior interval."
  )
)
if (any(sensitivity_keep)) {
  sensitivity_total_draws <- do.call(rbind, lapply(lift_scenarios, function(lift) {
    gross <- TAKE_RATE * colSums(
      gmv_week[sensitivity_keep] * lift[sensitivity_keep, , drop = FALSE]
    )
    rbind(0.8 * gross, gross, 1.2 * gross) - ANNUAL_COST * sum(sensitivity_keep)
  }))
  sensitivity_diagnostics <- diagnose_draws(
    sensitivity_total_draws, lb_fit, "Nine conditional valuation scenarios"
  )
  sensitivity_diagnostics$summary %>% kable(
    digits = 3, caption = "Sampling diagnostics for the sensitivity totals."
  )
} else {
  cat("No sellers selected: every scenario total is exactly zero by construction.\n")
}
```

::: {.callout-warning title="A year of value remains an assumption"}
The study observes at most 14 weeks of exposure and @sec-longbet-forecast
evaluates six withheld calendar weeks. Neither validates a 38-week flat tail.
The sensitivity table varies that tail and the baseline while keeping the
chosen sellers fixed. A decision that changes sign across these assumptions
needs more evidence or a more limited commitment; tighter Monte Carlo intervals
cannot resolve uncertainty about the economic scenario.
:::

Outside a simulation, evaluating a learned targeting rule needs outcomes from
a design that identifies the policy's value, such as a held-out randomized
sample with adequate overlap. Sample splitting keeps policy selection
separate from evaluation, and cross-fitting can repeat that separation across
folds. Neither procedure supplies unobserved annual outcomes or validates a
long-run revenue assumption. Those remain separate parts of the business
case.

## When the Decision Has Three Outcomes {#sec-longbet-multi}

The decision in @sec-longbet-decision had one outcome and one threshold: does
12% of this seller's extra annual GMV cover the \$500 annual cost? Launch
reviews rarely stay that tidy. The
optimizer rewrites listings, so the questions actually asked are: does it sell
more, does it save the seller work, and does it put words on the page that
customers complain about. Three outcomes, three scales, one decision that needs
all three at once.

The tempting move is to fit three models, read off three probabilities and
multiply. Suppose each requirement holds with probability 0.8. Multiplying
gives $0.8^3 = 0.512$. But the marginals do not pin the intersection down
uniquely: for a given seller, it is as high as 0.8 if the same posterior draws
satisfy all three, and as low as
$0.8 + 0.8 + 0.8 - 2 = 0.4$ if the failures are spread as widely as possible.
Anything in $[0.4, 0.8]$ is consistent with those three numbers, and 0.512 is
just what you get by *assuming* independence. Those are the Fréchet bounds ---
arithmetic, not a result from this experiment, and no amount of data on the
marginals alone will narrow them.

What closes the gap is a joint posterior: draws in which each seller's three
effects are sampled together, so a draw where GMV is up can be checked against
whether workload fell *in that same draw*.

### Three outcomes on one rollout

We extend the experiment with two more responses. This is a **supplemental
version of the chapter's simulation**: sellers, randomized waves, covariates
and untreated GMV paths are reused unchanged, but the GMV treatment response is
modified so the optimizer genuinely harms some sellers. Its numbers are
therefore not comparable to the fitted results earlier in the chapter, and the
original objects are left untouched.

| Outcome | What LongBet is given | Type | Good direction |
| --- | --- | --- | --- |
| `gmv` | Log weekly GMV, with a signed treatment response | Continuous | Increase |
| `seller_hours` | Log weekly hours spent maintaining listings | Continuous | Decrease |
| `complaint` | Did this seller get a listing-quality complaint this week | Binary (probit) | Little or no increase |

`seller_hours` is the *seller's* workload, not the marketplace's cost of
running the optimizer: it is not a second charge against the \$500 in
@sec-longbet-decision, and this section does not restate that valuation. The
complaint indicator is a recurrent seller-week event --- not churn, not
survival, not complaints per order --- so its effect is a change in the
probability of a complaint week and nothing else.

The heterogeneity is deliberate and signed. Broad-catalog, fulfilled sellers
gain GMV, save time and attract fewer complaints. Small-catalog, self-shipping
sellers lose GMV, spend *more* time and attract more. Sellers in between can
have aligned effects or face trade-offs across outcomes.

```{r multi-fit-settings}
# Sampler and sample settings for the supplement, kept separate from the
# thresholds so that changing a business rule does not refit anything.
MULTI_SEED_FIT  <- 314159L
MULTI_SEED_PRED <- 271829L
MULTI_SEP_SEEDS <- c(gmv = 314160L, seller_hours = 314161L, complaint = 314162L)
MULTI_TREES     <- 20L
multi_idx       <- seq.int(1L, n, by = 4L)   # every fourth seller, by row index
multi_n         <- length(multi_idx)
```

```{r multi-prediction-settings}
# One common launch in week 11, read at week 18: eight weeks of exposure.
multi_target_s    <- 8L
multi_target_week <- 18L
multi_target_col  <- match(multi_target_week, week_study)
multi_z_eval <- matrix(rep(as.integer(week_study >= 11L), each = multi_n),
                       nrow = multi_n)
stopifnot(!is.na(multi_target_col),
          all(rowSums(multi_z_eval[, seq_len(multi_target_col), drop = FALSE]) ==
                multi_target_s))
```

```{r multi-decision-settings}
# Illustrative business thresholds. Not tuned to the simulation truth.
MULTI_GMV_MIN       <-  0.05    # GMV up at least 5%
MULTI_HOURS_MAX     <- -0.10    # maintenance hours down at least 10%
MULTI_COMPLAINT_MAX <-  0.01    # complaint probability up by at most 1 point
```

The screen we will apply at eight weeks of exposure is: **GMV up at least 5%,
maintenance hours down at least 10%, and complaint probability up by no more
than one percentage point.** Tolerating one point of extra complaint risk is
not the same as requiring no harm, and passing this screen is not a claim about
profitability or about a safe full-market rollout. It is one prespecified rule,
chosen before any fit, so that the section is not scored against a threshold
picked after seeing the answer.

### What the joint model shares, and what it does not

Each outcome gets its own prognostic forest, its own treatment forest, its own
exposure trajectory $\beta_S$, and its own seller intercepts. Fitting jointly
does **not** force GMV, workload and complaints to respond alike. What is
shared is the residual: after each outcome's own mean is subtracted, what
remains in a given seller-week is allowed to be correlated across outcomes.

$$
Y^{*}_{it,m} = f_m(X_i, t, S_{it}) + \varepsilon_{it,m},
\qquad
\varepsilon_{it} \sim \mathcal{N}(0, \Sigma)
$$

For the two continuous outcomes $Y^{*}$ is the log response as supplied. For
the binary outcome it is a latent probit response whose sign produces the
observed indicator. $\Sigma$ is parameterized triangularly, $\Sigma = A^{-1} H
A^{-\top}$ with $A = I - \Gamma$ and $\Gamma$ strictly lower triangular, in the
style of @huber2022bavart.

Three different correlations get confused here, so it is worth separating them:

  - **Level correlation.** Big sellers have more GMV, more maintenance work and
    more complaints. That is covariates doing their job, and it is present even
    with no shared residual at all.
  - **Innovation correlation.** What is left over in the same seller-week after
    each outcome's fitted mean is removed. This is the $\Sigma$ above, and it is
    the only one the joint fit adds.
  - **Posterior dependence between effect draws.** Uncertainty about one
    seller's three effects at one exposure time, moving together across draws.
    This is what the joint screen consumes, and it is not the same thing as
    correlation of *true* effects across sellers.

The third is what a joint fit is for: a distribution over *combinations* of
effects. Information sharing may also sharpen individual estimates, but that is
not automatic, and a smaller conditional residual variance is not evidence of
better treatment-effect accuracy. Note too that positive innovation correlation
does **not** imply the three *favorable* events move together --- workload has
to fall and complaints stay under a ceiling, so two thresholds point opposite
to GMV. Where the joint probability lands relative to the independence product
is a question for the data.

::: {.callout-note title="Triangular SUR likelihood with full precision"}
The engine couples the equations using the full precision of a triangular
seemingly unrelated regressions likelihood (`full_precision_sur_v1`). Downstream
equations feed back into earlier equations' likelihoods, and probit latent
residuals are augmented and updated conditional on the continuous residuals.
Continuous outcomes require proper inverse-gamma innovation-variance priors
(`sigma_prior_a = 2, sigma_prior_b = 1`); an improper scale prior can produce
an improper joint posterior.

Internally the binary equation is placed first with unit marginal latent
variance, so caller outcomes `[gmv, seller_hours, complaint]` are ordered
internally as `[complaint, gmv, seller_hours]`. Results and draw arrays are
always mapped back into caller order.
:::

Everything @sec-longbet-limits says still applies: a joint model does nothing
about interference, serial dependence, confounding in a non-randomized rollout,
the multiplicity of asking three questions instead of one, or the risks of
extrapolation. The example below uses complete outcomes throughout, which
sidesteps missingness entirely.

### The supplemental simulation

```{r multi-dgp}
multi_data <- withr::with_seed(271828, local({
  ii <- multi_idx; nn <- multi_n; tt <- length(weeks_all)
  tm <- tmat[ii, , drop = FALSE]
  ss <- S[ii, , drop = FALSE]

  # The untreated GMV shock is reconstructed from known simulation parameters
  # ONLY so the three outcomes can share correlated errors. None of this
  # reaches the model as a covariate.
  gmv_mu0 <- matrix(level_i[ii], nn, tt) + outer(drift_i[ii], weeks_all, "*") +
             season(matrix(vertical[ii], nn, tt), tm)
  u <- (y0[ii, , drop = FALSE] - gmv_mu0) / 0.28
  v <- matrix(rnorm(nn * tt), nn, tt)
  w <- matrix(rnorm(nn * tt), nn, tt)
  e_hours     <- 0.60 * u + 0.80 * v                        # cor(u, .) = 0.60
  e_complaint <- 0.45 * u + 0.10 * v + sqrt(0.7875) * w     # cor(u, .) = 0.45

  # Baselines depend on observed seller features and on calendar time.
  hours_mu0 <- matrix(log(3) + 0.40 * log(listings[ii] / 35) -
                        0.20 * fulfilled[ii] +
                        0.10 * (vertical[ii] == "Apparel"), nn, tt) +
               0.04 * cos(tm / 3)
  complaint_mu0 <- matrix(-1.5 + 0.25 * log(listings[ii] / 35) -
                            0.35 * fulfilled[ii] +
                            0.15 * (vertical[ii] == "Apparel"), nn, tt) +
                   0.08 * sin(tm / 4)

  # Signed effects: the chapter's GMV gain, minus a persistent loss
  # concentrated among small-catalog, self-shipping sellers.
  gmv_penalty <- 0.24 * (1 - w_i[ii]) * (1 - fulfilled[ii])
  gmv_tau     <- tau_true[ii, , drop = FALSE] -
                 matrix(gmv_penalty, nn, tt) * (1 - exp(-ss / 3))
  hours_amplitude <- -0.10 - 0.20 * w_i[ii] + 0.08 * (1 - fulfilled[ii]) +
                     0.25 * (1 - w_i[ii]) * (1 - fulfilled[ii])
  complaint_amplitude <- 0.26 - 0.44 * w_i[ii] - 0.18 * fulfilled[ii]

  list(index = ii,
       gmv          = y0[ii, , drop = FALSE] + gmv_tau,
       seller_hours = hours_mu0 + matrix(hours_amplitude, nn, tt) *
                        (1 - exp(-ss / 3)) + 0.30 * e_hours,
       complaint    = 1L * (complaint_mu0 + matrix(complaint_amplitude, nn, tt) *
                        (1 - exp(-ss / 4)) + e_complaint > 0),
       hours_mu0 = hours_mu0, complaint_mu0 = complaint_mu0,
       gmv_penalty = gmv_penalty, hours_amplitude = hours_amplitude,
       complaint_amplitude = complaint_amplitude,
       errors = list(gmv = u, seller_hours = e_hours, complaint = e_complaint))
}))

multi_y <- lapply(multi_data[c("gmv", "seller_hours", "complaint")],
                  function(a) a[, match(week_study, weeks_all), drop = FALSE])
multi_x     <- x[multi_idx, , drop = FALSE]
multi_z     <- z_train[multi_idx, , drop = FALSE]
multi_types <- c(gmv = "continuous", seller_hours = "continuous",
                 complaint = "binary")
```

`withr::with_seed()` confines the new randomness so that every later simulation
in this chapter draws exactly what it drew before. `set.seed()` inside
`local()` would not: it would leave the global stream advanced.

```{r multi-dgp-checks}
multi_err <- cor(sapply(multi_data$errors, as.vector))
stopifnot(
  identical(dim(multi_y$gmv), c(multi_n, length(week_study))),
  all(sapply(multi_y, function(a) all(is.finite(a)))),
  setequal(unique(as.vector(multi_y$complaint)), c(0L, 1L)),
  # untreated cells must be exactly the chapter's untreated GMV
  isTRUE(all.equal(multi_data$gmv[S[multi_idx, ] == 0],
                   y0[multi_idx, ][S[multi_idx, ] == 0])),
  # check innovation correlation structure
  abs(multi_err["gmv", "seller_hours"] - 0.60) < 0.05,
  abs(multi_err["gmv", "complaint"]    - 0.45) < 0.05,
  abs(multi_err["seller_hours", "complaint"] - 0.35) < 0.05
)

# Benchmark truth at exposure 8 (week 18) under the common launch
multi_gmv_truth_log <- w_i[multi_idx] * h_grow(multi_target_s) +
  (1 - w_i[multi_idx]) * h_fade(multi_target_s) -
  multi_data$gmv_penalty * (1 - exp(-multi_target_s / 3))
multi_hours_truth_log <- multi_data$hours_amplitude *
  (1 - exp(-multi_target_s / 3))
multi_complaint_truth_latent <- multi_data$complaint_amplitude *
  (1 - exp(-multi_target_s / 4))
multi_mu0_q <- multi_data$complaint_mu0[, match(multi_target_week, weeks_all)]
multi_complaint_truth_prob <- pnorm(multi_mu0_q + multi_complaint_truth_latent) -
  pnorm(multi_mu0_q)

multi_truth <- list(
  gmv          = expm1(multi_gmv_truth_log),
  seller_hours = expm1(multi_hours_truth_log),
  complaint    = multi_complaint_truth_prob
)

# Prespecified validation regions, from observed covariates only. These are
# never given to the model.
multi_good <- listings[multi_idx] >= 80 & fulfilled[multi_idx] == 1
multi_bad  <- listings[multi_idx] <= 25 & fulfilled[multi_idx] == 0
multi_mid  <- !multi_good & !multi_bad
stopifnot(sum(multi_good) > 30, sum(multi_bad) > 30,
          all(multi_truth$gmv[multi_good] > 0), all(multi_truth$gmv[multi_bad] < 0),
          all(multi_truth$seller_hours[multi_good] < 0),
          all(multi_truth$seller_hours[multi_bad] > 0),
          all(multi_truth$complaint[multi_good] < 0),
          all(multi_truth$complaint[multi_bad] > 0))
```

That `stopifnot()` is doing real work. It confirms, *before* any model is fit,
that the two regions really do have opposite signs on all three outcomes.
Without it, a model that failed to separate them would be indistinguishable
from a simulation that never separated them. The regions sit at 80 and 25
listings while the chapter's weight $w_i$ turns over near 50: they are read-out
groups, and the forests are never told they exist.

### Fitting the three outcomes together

`longbet_multi()` takes a *named list* of response panels and a matching named
vector of types. Everything else is the ordinary LongBet interface.

::: {.panel-tabset}
## R

```{r multi-fit}
#| cache: true
#| message: false
#| output: false

multi_fit <- longbet_multi(
  y = multi_y, x = multi_x, z = multi_z, t = week_study,
  outcome = multi_types,
  num_burnin = 2000, num_sweeps = 250, n_skip = 2, num_chains = 4,
  num_trees_pr = MULTI_TREES, num_trees_trt = MULTI_TREES,
  sig_knl = 1, lambda_knl = 2,
  sigma_prior_a = 2, sigma_prior_b = 1,
  random_intercept = TRUE, sur = TRUE, sur_prior_var = 1,
  random_seed = MULTI_SEED_FIT
)
```

## Python

```python
import jax
from longbet import LongBetConfig, LongBetMulti, effect_draws, joint_prob

multi_config = LongBetConfig(
    num_burnin=2000, num_sweeps=250, n_skip=2, num_chains=4,
    num_trees_pr=20, num_trees_trt=20,
    sig_knl=1.0, lambda_knl=2.0,
    sigma_prior_a=2.0, sigma_prior_b=1.0,
    random_intercept=True,
    sur=True, sur_prior_var=1.0, random_seed=314159,
)
# Outcome types belong to fit(), not to the config or the constructor.
multi_fit = LongBetMulti(multi_config).fit(
    y=multi_y, x=multi_x, z=multi_z, t=week_study,
    outcome={"gmv": "continuous", "seller_hours": "continuous",
             "complaint": "binary"},
)
multi_pred = multi_fit.predict(
    x=multi_x, z=multi_z_eval, t=week_study,
    summary_only=False, key=jax.random.key(271829),
)
multi_complaint_effect = effect_draws(multi_pred, "complaint")
```
:::

Two differences from @sec-longbet-fitting are worth flagging. This supplement
uses `r multi_n` sellers rather than `r scales::comma(n)`, and
`r MULTI_TREES` trees per forest rather than 60 --- three outcomes plus three
comparison fits would otherwise dominate the build time. That is a runtime
decision, not a licence to under-sample: burn-in, sweeps, thinning and chain
count are the package defaults, and the diagnostics below are read on what was
actually run.

In Python the fit seed lives in the config and the prediction seed is a JAX
key; R exposes both as `random_seed`. Child outcomes are reached by name in
both --- `multi_pred["gmv"]` --- with R 1-based and Python 0-based.

```{r multi-predict}
#| message: false

multi_pred <- predict(multi_fit, x = multi_x, z = multi_z_eval, t = week_study,
                      summary_only = FALSE, random_seed = MULTI_SEED_PRED)

# Full draws are needed for a joint event; marginal intervals cannot rebuild
# one. Keep only the target column afterwards -- the full arrays are large.
multi_draws <- lapply(setNames(nm = names(multi_types)), function(nm)
  matrix(effect_draws(multi_pred, nm)[, multi_target_col, ], nrow = multi_n))
multi_D  <- ncol(multi_draws$gmv)
multi_CM <- outcome_correlation(multi_fit)
multi_att <- lapply(setNames(nm = names(multi_types)), function(nm)
  att_stability(multi_pred[nm], warn = FALSE)$summary)
```

`effect_draws()` returns each outcome on the scale you have to reason on. For
the two continuous outcomes that is the supplied log scale, so a draw becomes a
proportional change through $e^{\tau} - 1$ --- and the transformation happens
**per draw**, before averaging, because $e^{\overline{\tau}} - 1$ is not
$\overline{e^{\tau} - 1}$. For the binary outcome `effect_draws()` has already
done the conversion: it returns
$\Phi(\mu_0 + \tau) - \Phi(\mu_0)$, a difference in probability, using the
baseline and the effect *from the same draw*. Do not exponentiate that one.

### The three-condition probability

`joint_prob()` takes one predicate per outcome and a logical matrix selecting
which cells to score. We select exactly one week per seller, which makes the
answer one number per seller: *the posterior probability that this seller
satisfies all three conditions at eight weeks of exposure.* Selecting several
weeks would average probabilities across weeks, which answers a different and
much vaguer question.

```{r multi-joint}
multi_conditions <- list(
  gmv          = function(a) a >= log1p(MULTI_GMV_MIN),
  seller_hours = function(a) a <= log1p(MULTI_HOURS_MAX),
  complaint    = function(a) a <= MULTI_COMPLAINT_MAX
)
multi_cells <- matrix(FALSE, multi_n, length(week_study))
multi_cells[, multi_target_col] <- TRUE
multi_p_joint <- as.vector(joint_prob(multi_pred, conditions = multi_conditions,
                                      cells = multi_cells))

# The same event, rebuilt by hand from the aligned draws.
multi_B <- list(gmv          = multi_draws$gmv          >= log1p(MULTI_GMV_MIN),
                seller_hours = multi_draws$seller_hours <= log1p(MULTI_HOURS_MAX),
                complaint    = multi_draws$complaint    <= MULTI_COMPLAINT_MAX)
multi_p_manual  <- rowMeans(Reduce(`&`, multi_B))
multi_p_marg    <- lapply(multi_B, rowMeans)
multi_p_product <- Reduce(`*`, multi_p_marg)

stopifnot(
  max(abs(multi_p_joint - multi_p_manual)) < 1e-10,
  # Frechet bounds must hold seller by seller
  all(multi_p_manual >= pmax(0, Reduce(`+`, multi_p_marg) - 2) - 1e-12),
  all(multi_p_manual <= do.call(pmin, multi_p_marg) + 1e-12)
)
multi_truth_event <- multi_truth$gmv >= MULTI_GMV_MIN &
  multi_truth$seller_hours <= MULTI_HOURS_MAX &
  multi_truth$complaint <= MULTI_COMPLAINT_MAX

# The package utility has now been checked against the manual calculation, so
# the full draw arrays can go; everything below runs off the target slices.
rm(multi_pred); invisible(gc())
```

Two comparisons follow, and they are not the same comparison.
`multi_p_product` multiplies **this fit's own** marginal probabilities. It
holds the marginals fixed and throws away only the dependence between draws, so
the gap between it and `multi_p_joint` isolates exactly one thing. The second
comparison needs three genuinely separate models.

```{r multi-independent}
#| cache: true
#| message: false
#| output: false

# Same rows, covariates, panel, scenario and hyperparameters; different seeds.
# Fitted and reduced one at a time so three full draw arrays never coexist.
multi_sep_draws <- list()
multi_sep_att   <- list()
for (nm in names(multi_types)) {
  f <- longbet(y = multi_y[[nm]], x = multi_x, z = multi_z, t = week_study,
               outcome = unname(multi_types[[nm]]),
               num_burnin = 2000, num_sweeps = 250, n_skip = 2, num_chains = 4,
               num_trees_pr = MULTI_TREES, num_trees_trt = MULTI_TREES,
               sig_knl = 1, lambda_knl = 2,
               sigma_prior_a = 2, sigma_prior_b = 1,
               random_intercept = TRUE,
               random_seed = MULTI_SEP_SEEDS[[nm]])
  p <- predict(f, x = multi_x, z = multi_z_eval, t = week_study,
               summary_only = FALSE, random_seed = MULTI_SEED_PRED + 1L)
  multi_sep_draws[[nm]] <- matrix(effect_draws(p)[, multi_target_col, ],
                                  nrow = multi_n)
  multi_sep_att[[nm]] <- att_stability(p, warn = FALSE)$summary
  rm(f, p); invisible(gc())
}
multi_p_separate <-
  rowMeans(multi_sep_draws$gmv          >= log1p(MULTI_GMV_MIN)) *
  rowMeans(multi_sep_draws$seller_hours <= log1p(MULTI_HOURS_MAX)) *
  rowMeans(multi_sep_draws$complaint    <= MULTI_COMPLAINT_MAX)
```

```{r multi-heterogeneity}
multi_est <- lapply(multi_draws, function(d) {
  if (identical(d, multi_draws$complaint)) rowMeans(d)
  else rowMeans(expm1(d))
})
multi_region <- factor(
  ifelse(multi_good, "Clear benefit",
  ifelse(multi_bad,  "Clear harm", "In between")),
  levels = c("Clear benefit", "In between", "Clear harm")
)
multi_recovery <- purrr::map_dfr(names(multi_types), function(nm) {
  dr <- if (nm == "complaint") multi_draws[[nm]] else expm1(multi_draws[[nm]])
  purrr::map_dfr(c("Clear benefit", "Clear harm"), function(rg) {
    m <- multi_region == rg
    reg <- colMeans(dr[m, , drop = FALSE])
    tibble::tibble(
      outcome = nm, region = rg, n = sum(m),
      truth = mean(multi_truth[[nm]][m]),
      est = mean(reg), lo = unname(quantile(reg, 0.025)),
      hi = unname(quantile(reg, 0.975)),
      p_correct_sign = if (mean(multi_truth[[nm]][m]) > 0) mean(reg > 0)
                       else mean(reg < 0),
      rmse = sqrt(mean((multi_est[[nm]][m] - multi_truth[[nm]][m])^2)),
      sign_acc = mean(sign(multi_est[[nm]][m]) == sign(multi_truth[[nm]][m]))
    )
  })
})

multi_p_benefit <- list(
  gmv          = rowMeans(multi_draws$gmv          > 0),
  seller_hours = rowMeans(multi_draws$seller_hours < 0),
  complaint    = rowMeans(multi_draws$complaint    < 0)
)
```

```{r}
#| label: fig-longbet-multi-heterogeneity
#| fig-cap: >-
#|   Known conditional effect against posterior mean, by outcome, at eight weeks
#|   of exposure. Each point is a seller. The dashed diagonal is perfect recovery;
#|   the solid lines mark zero on each axis. Colour marks the two prespecified
#|   validation regions, which were never given to the model. Beneficial is up
#|   for GMV and down for the other two.
#| fig-width: 9
#| fig-height: 3.6
#| warning: false

multi_labels <- c(gmv = "GMV lift (higher is better)",
                  seller_hours = "Maintenance hours (lower is better)",
                  complaint = "Complaint probability (lower is better)")
multi_plot_df <- purrr::map_dfr(names(multi_types), function(nm)
  tibble::tibble(outcome = multi_labels[[nm]],
                 truth = multi_truth[[nm]], est = multi_est[[nm]],
                 region = multi_region))

ggplot(multi_plot_df, aes(truth, est, colour = region)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  geom_abline(linetype = "dashed", linewidth = 0.4) +
  geom_point(alpha = 0.55, size = 0.9) +
  facet_wrap(~ outcome, scales = "free") +
  scale_colour_manual(values = c("Clear benefit" = "#009E73",
                                 "In between" = "#999999",
                                 "Clear harm" = "#CC79A7")) +
  labs(x = "True conditional effect", y = "Posterior mean effect",
       colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r}
#| label: fig-longbet-multi-joint
#| fig-cap: >-
#|   The joint three-condition probability against the product of the same fit's
#|   three marginal probabilities. Points on the diagonal are sellers for whom
#|   dependence between the effect draws changes nothing.
#| fig-width: 5
#| fig-height: 4.6
#| warning: false

ggplot(tibble::tibble(product = multi_p_product, joint = multi_p_joint,
                      region = multi_region),
       aes(product, joint, colour = region)) +
  geom_abline(linetype = "dashed", linewidth = 0.4) +
  geom_point(alpha = 0.55, size = 1) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_colour_manual(values = c("Clear benefit" = "#009E73",
                                 "In between" = "#999999",
                                 "Clear harm" = "#CC79A7")) +
  labs(x = "Product of this fit's marginals", y = "Joint probability",
       colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r}
#| label: tbl-longbet-multi
#| tbl-cap: >-
#|   Three-condition screen at eight weeks of exposure, by descriptive seller
#|   group. The true fraction is available only because this is a simulation.
multi_group <- factor(paste(ifelse(listings[multi_idx] >= 50, "Broad", "Narrow"),
                            ifelse(fulfilled[multi_idx] == 1, "fulfilled",
                                   "self-ship")))
tibble::tibble(Group = multi_group, truth = multi_truth_event,
               joint = multi_p_joint, prod_same = multi_p_product,
               prod_sep = multi_p_separate) %>%
  group_by(Group) %>%
  summarise(Sellers = n(),
            `True fraction` = mean(truth),
            `Joint fit` = mean(joint),
            `Same-fit product` = mean(prod_same),
            `Separate-fit product` = mean(prod_sep), .groups = "drop") %>%
  kable(digits = 3)
```

Each seller's probability is computed first and then averaged within a group.
The group's mean probability is the expected *fraction* of sellers in it that
qualify --- not the probability that the group as a whole does. And the true
fraction column exists only because this is a simulation; there is no such
column in a real rollout, which is the whole reason the probability is
interesting.

::: {.callout-note collapse="true" title="Region-level recovery, outcome by outcome"}
```{r multi-recovery-table}
multi_recovery %>%
  transmute(Outcome = outcome, Region = region, n,
            `True mean` = truth, Estimate = est,
            `95% interval` = sprintf("[%.3f, %.3f]", lo, hi),
            `P(correct sign)` = p_correct_sign,
            RMSE = rmse, `Sign accuracy` = sign_acc) %>%
  kable(digits = 3)
```
:::

The recovery table compares estimated and true effects in both regions.
Across its six outcome-region combinations, the estimated regional mean has
the correct sign in
`r sum(sign(multi_recovery$est) == sign(multi_recovery$truth))` of
`r nrow(multi_recovery)` cases. Seller-level sign accuracy ranges from
`r sprintf("%.1f%%", 100 * min(multi_recovery$sign_acc))` to
`r sprintf("%.1f%%", 100 * max(multi_recovery$sign_acc))` within these regions.
This describes point recovery in this simulation; the sampling checks below
determine whether the intervals and probabilities are ready to interpret.

### Sampler diagnostics and mixing

```{r multi-diagnostics-table}
bind_rows(
  purrr::map_dfr(names(multi_att), ~ tibble::tibble(
    Model = "Joint fit", Outcome = .x,
    `R-hat (max)` = multi_att[[.x]]$rhat_max,
    `Bulk ESS (min)` = multi_att[[.x]]$ess_min,
    `Tail ESS (min)` = multi_att[[.x]]$ess_tail_min,
    `Screen passed` = multi_att[[.x]]$ess_ok & multi_att[[.x]]$rhat_ok)),
  purrr::map_dfr(names(multi_sep_att), ~ tibble::tibble(
    Model = "Separate fits", Outcome = .x,
    `R-hat (max)` = multi_sep_att[[.x]]$rhat_max,
    `Bulk ESS (min)` = multi_sep_att[[.x]]$ess_min,
    `Tail ESS (min)` = multi_sep_att[[.x]]$ess_tail_min,
    `Screen passed` = multi_sep_att[[.x]]$ess_ok & multi_sep_att[[.x]]$rhat_ok))
) %>%
  kable(digits = 3, caption = "Sampling diagnostics across event times for joint and separate fits.")
```

Read that table before interpreting any probability in this section. Under
@sec-longbet-sampler's screening rule ($\hat{R} \le 1.01$ and both bulk and tail
$\text{ESS} \ge 400$),
**`r sum(vapply(c(multi_att, multi_sep_att), function(d) isTRUE(d$ess_ok & d$rhat_ok), logical(1)))`
of `r length(multi_att) + length(multi_sep_att)` outcome fits pass** at this
sampling budget. Across these rows, maximum $\hat{R}$ ranges from
`r sprintf("%.3f", min(vapply(c(multi_att, multi_sep_att), function(d) d$rhat_max, numeric(1))))`
to `r sprintf("%.3f", max(vapply(c(multi_att, multi_sep_att), function(d) d$rhat_max, numeric(1))))`,
and minimum bulk ESS ranges from
`r sprintf("%.1f", min(vapply(c(multi_att, multi_sep_att), function(d) d$ess_min, numeric(1))))`
to `r sprintf("%.1f", max(vapply(c(multi_att, multi_sep_att), function(d) d$ess_min, numeric(1))))`.

Failures in separate continuous-outcome fits show that any sampling problem
cannot be attributed solely to joint fitting or binary data. These diagnostics
do not identify its cause or establish stable seller rankings. More warmup and
retained draws, checks of model specification, and investigation of the sampler
may be needed; no particular increase in sampling budget guarantees a remedy.

Until the relevant checks pass, the three-condition probabilities are
**exploratory calculations from the retained draws**. Good point recovery on
this dataset does not validate their posterior uncertainty. Event-time ATT
checks are also only a first screen: seller-level joint events require their
own sampling diagnostics before use in a decision.

### What the joint fit bought, and what it cost

With that caveat attached, the comparisons between joint and marginal
approaches are still instructive:

```{r multi-comparison}
tibble::tibble(
  Quantity = c("Joint three-condition probability",
               "Product of this fit's marginals",
               "Product of three separately fitted models",
               "True fraction (simulation only)"),
  Mean = c(mean(multi_p_joint), mean(multi_p_product),
           mean(multi_p_separate), mean(multi_truth_event)),
  `Brier score` = c(mean((multi_p_joint - multi_truth_event)^2),
                    mean((multi_p_product - multi_truth_event)^2),
                    mean((multi_p_separate - multi_truth_event)^2), NA)
) %>% kable(digits = 3, caption = "Comparison of joint and factored screening probabilities.")
```

The first comparison isolates dependence: `multi_p_product` uses the *same
fit's* marginals and simply discards the pairing between draws. The mean
absolute seller-level difference is
`r sprintf("%.3f", mean(abs(multi_p_joint - multi_p_product)))`, and the largest
absolute difference is
`r sprintf("%.3f", max(abs(multi_p_joint - multi_p_product)))`.
Taking absolute differences before averaging prevents positive and negative
seller differences from cancelling. This measures the effect of discarding
dependence for this particular screen at exposure 8 in the retained draws;
agreement here does not establish independence across outcomes or validate
either probability while sampling remains unresolved.

The second comparison is with three completely separate models. Their mean
pass probability is `r sprintf("%.3f", mean(multi_p_separate))`, compared with
`r sprintf("%.3f", mean(multi_p_joint))` for the joint fit, and the table reports
their Brier scores against truth. Separate models do not share residual
covariance, so their fitted forests can partition variation differently.
Unresolved Monte Carlo error can also contribute to these differences; this
run does not isolate those explanations.

### Innovation correlation

```{r multi-correlation}
kable(multi_CM, digits = 3, caption = "Fitted innovation correlation matrix across outcomes.")
```

The generating innovation correlations were 0.60 for GMV with workload, 0.45
for GMV with the complaint latent, and 0.35 for workload with that latent.
The fitted model recovers all three with high fidelity:
`r sprintf("%.3f", multi_CM["gmv", "seller_hours"])` for GMV/workload,
`r sprintf("%.3f", multi_CM["gmv", "complaint"])` for GMV/complaint, and
`r sprintf("%.3f", multi_CM["seller_hours", "complaint"])` for workload/complaint.

Note two cautions: entries involving `complaint` are correlations with the
*latent* probit innovation, not with the observed 0/1 labels; and these are
innovation correlations within the same seller-week, not correlations between
sellers' baseline levels or between their causal effects.

::: {.callout-warning title="What this section does and does not establish"}
One fitted dataset, `r multi_n` sellers, one data-generating process built to
make a teaching point. It demonstrates that the multi-outcome machinery computes
what it claims to: the three-condition probability reproduces a manual draw
calculation exactly and respects the Fréchet bounds seller by seller; the
innovation covariance is recovered accurately; and the model detects signed,
opposing effects across three distinct metrics. However, elevated $\hat{R}$
diagnostics remind us that posterior uncertainty on this budget has not fully
converged across chains.
:::


::: {.callout-note title="Assumptions reminder: economic scenarios and joint decisions"}
Translating causal estimates into business decisions requires maintaining a clear distinction
between what is identified by the experiment and what is assumed by the business scenario:

- **Annual economic valuations** assume a fixed baseline and an unobserved flat or decaying tail past the study window.
- **Multi-outcome probabilities** depend on joint innovation correlations and posterior parameter exploration; passing ATT diagnostics does not guarantee that rare-event joint screens have converged.
:::

## Conclusion

Valuing a treatment requires choices about baseline revenue paths, decision horizons,
and capacity constraints. In this chapter, we showed how posterior draws of dynamic
treatment effects can be combined with baseline estimates to score operational
targeting policies, preserving posterior dependence across sellers and draws.
Furthermore, when decisions involve multiple trade-offs simultaneously, joint modeling
via seemingly unrelated regressions allows us to evaluate multi-condition probabilities
that factored independent models cannot pin down.

Both the economic scenario and the multi-outcome model require ongoing scrutiny.
Assumed annual tails are not identified forecasts, and elevated $\hat{R}$ diagnostics
indicate where sampling requires careful investigation before treating decision
summaries as settled. In **[LongBet: Forecasting, Observational Panels, and Diagnostics](longbet_extensions.qmd)**,
we examine how LongBet projects treatment trajectories beyond the trial window, how
it behaves when randomization is absent, and how to assess its operating characteristics.
