---
title: "Ordinal Regression with BART and the Cloglog Link"
share:
  permalink: "https://book.martinez.fyi/phobart.html"
  description: >-
    Ordinal regression with Bayesian Additive Regression Trees (BART) using
    the complementary log-log link to model non-linearities, interactions,
    and avoid cutpoint constraints.
  linkedin: true
  email: true
  mastodon: true
author:
  - name: Entejar Alam 
  - name: Ignacio Martinez  
---

## Ordinal Outcomes in Business

Many of the outcomes we care about most in industry are ordered, but not
numeric. Customer satisfaction is measured on a five-point Likert scale (one
to five stars), Net Promoter Score (NPS) responses fall into detractor,
passive, and promoter buckets, credit files are graded, and support tickets are
triaged into severity tiers. While practitioners often treat these ratings as
continuous numbers by averaging them, doing so imposes an unverifiable cardinal
scale that assumes equal distances between categories and can produce severe
distortions (@liddell2018analyzing). Rather than modeling raw scores with linear
regression, standard Bayesian models for ordinal data rely on ordered probit
and ordered logit (cumulative-link) models. These are a good starting point,
but they carry two assumptions that are easy to overlook and frequently wrong:

1.  **Linearity and additivity.** Predictors are assumed to impact a latent
    scale linearly, with no interactions unless we hand-specify them. For
    example, if a product change delights power users but frustrates newcomers,
    a linear model averages these out, reporting a misleading effect that
    describes neither group.
2.  **Ordered cutpoints in Gibbs sampling.** Cumulative-link models rely on
    estimating thresholds ($c_1 < \dots < c_{K-1}$) that must remain strictly
    ordered. While Hamiltonian Monte Carlo (e.g., in Stan) handles ordered
    cutpoints in parametric models via monotonic transformations, Gibbs
    samplers—the standard engine for tree ensembles like BART—struggle with
    strong posterior dependencies among constrained cutpoints, which
    significantly degrades MCMC mixing and slows down convergence (@cowles1996).

In previous chapters, we saw how BART (@chipman2010bart) relaxes the first
assumption for continuous outcomes, discovering non-linearities and
interactions on its own. This chapter extends that flexibility to ordinal
outcomes, utilizing the framework from @alam2025unified to eliminate the second
assumption as well.


## The Cloglog Link and PHOBART

The fix starts with the link function. Instead of probit or logit, we use the
complementary log-log (cloglog) link,
$$
g(p) = \log\{-\log(1 - p)\}.
$$
It is the only link under which the two standard formulations of ordinal
regression, cumulative-link and continuation-ratio, coincide
(@laara1985equivalence). For an outcome with $K$ ordered categories
$Y_i \in \{1, \dots, K\}$, @alam2025unified write the model in cumulative form:
$$
\text{cloglog}(\Pr(Y_i \le k \mid X_i = x)) = c_k + \lambda(x),
$$
or equivalently in discrete proportional-hazards form,
$$
\Pr(Y_i > k \mid X_i = x) = \{S_0(k)\}^{e^{-\lambda(x)}},
$$
where $\lambda(x)$ is assigned a BART sum-of-trees prior. This yields a
**proportional hazards ordinal BART** model (**PHOBART**).

Unlike the symmetric probit and logit links, the cloglog link is asymmetric:
the cumulative probability approaches 0 and 1 at different rates on the latent
scale. This asymmetry is advantageous when extreme rating categories behave
asymmetrically in probability space, while BART's tree ensemble is responsible
for discovering non-linear threshold effects and interactions across covariates
$X$.

PHOBART also elegantly resolves the cutpoint constraint issue. The strictly
ordered cutpoints $c_k$ are reparameterized through unconstrained parameters
$\gamma_j \in \mathbb{R}$ via:
$$
c_k = \log \sum_{j \le k} e^{\gamma_j}, \qquad k = 1, \dots, K-1.
$$
Because the exponential is strictly positive, the ordering
$c_1 < c_2 < \dots < c_{K-1}$ holds by construction. Furthermore, because the
inverse cloglog link is the CDF of an exponential distribution, augmenting each
observation with a truncated-exponential latent variable allows the sampler to
update the unconstrained parameters and the tree ensemble using standard
conjugate and Gibbs updates. While the sampler works on the unconstrained
scale, the package returns the transformed cutpoints $c_k$.

Under suitable Hölder smoothness and sparsity conditions on the latent
regression function $\lambda(x)$, @alam2025unified prove that PHOBART achieves
minimax-optimal posterior contraction rates (up to logarithmic factors),
guaranteeing that the posterior concentrates around the true data-generating
process at the optimal statistical rate.

The same latent-exponential machinery unifies ordinal, survival, and density
regression under one framework. In the practical example below, our code targets
`stochtree` version `0.5.x`. 


## A Business Example: A Feature Change and Customer Satisfaction

Suppose we roll out a product change and measure its effect on a one-to-five
satisfaction rating. Our concern is heterogeneity: we suspect the change helps
some customers and hurts others depending on how engaged they were to begin
with. If we are right, a linear ordered probit will average those opposing
effects into something misleading, and we could ship (or kill) the feature for
the wrong reason.

To see this cleanly, we simulate data where we know the truth. Let `x1` be a
centered engagement score and `t` the feature flag. The latent satisfaction `z`
is defined as
$$
z = 1.5 \sin\!\left(\frac{\pi\, x_1}{2}\right) + \tau(x_1)\, t + \epsilon,
\qquad \epsilon \sim \mathcal{N}(0, 0.8^2),
$$
where the feature's effect $\tau(x_1)$ is $+1.5$ for engaged customers
($x_1 > 0$) and $-1.0$ for less-engaged ones ($x_1 \le 0$). This construction
induces both a non-linear baseline effect through the sine function and
heterogeneous treatment effects whose sign depends on $x_1$. 

The observed ordinal outcome is generated by discretizing $z$ into a
five-point Likert scale using fixed thresholds. Because treatment $t$ is
assigned completely at random ($t \sim \text{Bernoulli}(0.5)$) independent of
$x_1$, causal identification (unconfoundedness and overlap) holds by design.
This isolates the comparison entirely to the functional form: whether each
model can accurately discover the non-linear response surface and subgroup
heterogeneity. The latent noise is Gaussian, which favors the probit benchmark
in terms of link specification.

```{r simulate, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(stochtree)
library(brms)
library(coda)
library(knitr)

set.seed(1982)
N <- 2500

fake_data <- tibble(
  x1 = runif(N, -2, 2),
  t  = sample(c(0, 1), N, replace = TRUE)
) %>%
  mutate(
    f_x    = 1.5 * sin(pi * x1 / 2),
    tau_x  = ifelse(x1 > 0, 1.5, -1.0),
    z_true = f_x + tau_x * t + rnorm(N, 0, 0.8),
    y = case_when(
      z_true <  -1.5 ~ 1,
      z_true <  -0.5 ~ 2,
      z_true <   0.5 ~ 3,
      z_true <   1.5 ~ 4,
      TRUE           ~ 5
    )
  )

# Verify all five rating categories are well-populated
table(fake_data$y)
```

We then compute the *true* population effect of the feature on each rating
level by integrating over the counterfactuals (feature off vs. feature on) for
every customer.

```{r truth, message=FALSE, warning=FALSE}
truth_ate <- fake_data %>%
  mutate(z0 = f_x, z1 = f_x + tau_x) %>%
  mutate(
    p0_1 = pnorm(-1.5, z0, 0.8),
    p0_2 = pnorm(-0.5, z0, 0.8) - pnorm(-1.5, z0, 0.8),
    p0_3 = pnorm( 0.5, z0, 0.8) - pnorm(-0.5, z0, 0.8),
    p0_4 = pnorm( 1.5, z0, 0.8) - pnorm( 0.5, z0, 0.8),
    p0_5 = 1 - pnorm(1.5, z0, 0.8),
    p1_1 = pnorm(-1.5, z1, 0.8),
    p1_2 = pnorm(-0.5, z1, 0.8) - pnorm(-1.5, z1, 0.8),
    p1_3 = pnorm( 0.5, z1, 0.8) - pnorm(-0.5, z1, 0.8),
    p1_4 = pnorm( 1.5, z1, 0.8) - pnorm( 0.5, z1, 0.8),
    p1_5 = 1 - pnorm(1.5, z1, 0.8)
  ) %>%
  summarise(across(starts_with("p0_") | starts_with("p1_"), mean)) %>%
  pivot_longer(everything(), names_to = c("group", "y"), names_sep = "_") %>%
  pivot_wider(names_from = group, values_from = value) %>%
  mutate(y = as.integer(y), tau = p1 - p0)

kable(
  truth_ate,
  digits = 3,
  col.names = c("Rating (y)", "P(Y=y | t=0)", "P(Y=y | t=1)", "True ATE (tau)")
)
```

### BART Model Fitting (PHOBART)

We fit the PHOBART model using `stochtree::bart()` with an `OutcomeModel`
specifying `outcome = "ordinal"` and `link = "cloglog"` inside
`general_params`. We also configure key hyperparameters:

- `sample_sigma2_global = FALSE`: Ordinal outcomes do not have a free Gaussian
  residual variance $\sigma^2$ because scale is fixed by the link function.
- `random_seed = 2025`: In `stochtree`, MCMC sampling randomness is controlled
  via `random_seed` inside `general_params` rather than base R's `set.seed()`.
- `mean_forest_params = list(num_trees = 50, sample_sigma2_leaf = FALSE)`:
  Configures 50 trees with leaf variance sampling disabled for stronger
  regularization.
- `num_gfr = 0`: Runs standard MCMC sampling from initialization.

We run a single MCMC chain here for brevity and monitor ESS and trace plots; in
production settings, one can set `num_chains` in `stochtree::bart()` to assess
convergence diagnostics such as $\hat{R}$ across multiple independent chains
alongside ESS.

```{r fit-phobart, message=FALSE, warning=FALSE}
X_train <- fake_data %>% select(x1, t) %>% as.matrix()
y_train <- fake_data$y

X0 <- X_train; X0[, "t"] <- 0
X1 <- X_train; X1[, "t"] <- 1

phobart <- bart(
  X_train = X_train,
  y_train = y_train,
  X_test  = rbind(X0, X1),
  num_gfr = 0,
  num_burnin = 1000,
  num_mcmc   = 1000,
  general_params = list(
    sample_sigma2_global = FALSE,
    random_seed = 2025,
    outcome_model = OutcomeModel(outcome = "ordinal", link = "cloglog")
  ),
  mean_forest_params = list(num_trees = 50, sample_sigma2_leaf = FALSE)
)
```

### Checking convergence

Before trusting the estimates, we check that the MCMC sampler has mixed well.
Cutpoint draws are extracted via
`extractParameter(phobart, "cloglog_cutpoints")`, returning a matrix of shape
$(K-1, S)$ containing draws of $c_k$.

Because the tree ensemble $\lambda(x)$ can absorb an arbitrary global shift
(an implicit intercept), raw cutpoint draws $c_k$ and the forest predictions
share a non-identifiable additive constant. If unadjusted, raw cutpoints can
drift even in a well-converged chain, yielding a misleadingly low Effective
Sample Size (ESS). To evaluate convergence on a strictly identified quantity,
we shift the cutpoints by the per-sample mean of the training predictions,
$\bar{\lambda}_{\text{train}} = \frac{1}{N}\sum_{i=1}^N \lambda(X_i)$, tracing
$c_2 + \bar{\lambda}_{\text{train}}$. (Note that this shift automatically
cancels when computing counterfactual contrast probabilities.)

```{r traceplot, message=FALSE, warning=FALSE}
cutpoint_draws <- extractParameter(phobart, "cloglog_cutpoints")
lambda_train   <- predict(
  phobart, X_train, scale = "linear", terms = "y_hat", type = "posterior"
)

# Shift cutpoints by the mean training latent prediction
adj_c2 <- cutpoint_draws[2, ] + colMeans(lambda_train)

tibble(
  iteration = seq_along(adj_c2),
  c2_adj    = adj_c2
) %>%
  ggplot(aes(iteration, c2_adj)) +
  geom_line(color = "blue", alpha = 0.6) +
  labs(
    title = "Trace plot for the second cutpoint (intercept-adjusted)",
    x = "Iteration",
    y = expression(c[2] + bar(lambda)[train])
  ) +
  theme_bw()
```

```{r ess, message=FALSE, warning=FALSE}
cat("ESS for adjusted c_2:", round(effectiveSize(adj_c2), 1), "\n")
```

The trace looks stationary with no drift or sticking, and the ESS is healthy,
confirming adequate posterior exploration. In practice, you would inspect the
remaining thresholds and tree ensemble predictions similarly.

### Turning the posterior into category probabilities

Instead of manually evaluating the link algebra, `stochtree` provides a direct
accessor:
`predict(..., scale = "probability", terms = "y_hat", type = "posterior")`
returns an array of dimension
$(\text{observations}, \text{categories}, \text{samples})$.
Averaging over observations yields the population-level category probabilities
for each posterior draw.

```{r phobart-probs, message=FALSE, warning=FALSE}
# Predict category probabilities: returns (observations, categories, samples)
probs0 <- predict(
  phobart, X0, scale = "probability", terms = "y_hat", type = "posterior"
)
probs1 <- predict(
  phobart, X1, scale = "probability", terms = "y_hat", type = "posterior"
)

# Average over observations to get population category probabilities per sample
pop0_phobart <- t(apply(probs0, c(2, 3), mean)) # samples x categories
pop1_phobart <- t(apply(probs1, c(2, 3), mean))

tau_phobart <- as_tibble(
  pop1_phobart - pop0_phobart,
  .name_repair = ~ paste0("V", seq_along(.x))
) %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "y", values_to = "tau") %>%
  mutate(y = as.integer(sub("V", "", y)))
```

### Benchmark: linear ordered probit

To benchmark our approach, we fit a Bayesian linear ordered probit model with
`brms`, to see how it performs when the true relationship is non-linear and
includes interactions.

```{r probit, message=FALSE, warning=FALSE, results="hide"}
dir.create("cache", showWarnings = FALSE)

probit_lm <- brm(
  y ~ x1 + t,
  data    = fake_data %>% mutate(y = ordered(y)),
  family  = cumulative("probit"),
  chains  = 4, cores = 4, iter = 2000, refresh = 0,
  seed    = 2025,
  file    = "cache/probit_lm"
)
```

```{r probit-probs, message=FALSE, warning=FALSE}
# posterior_epred returns draws x observations x categories for ordinal families
ep0 <- posterior_epred(
  probit_lm,
  newdata = fake_data %>% mutate(t = 0, y = ordered(y))
)
ep1 <- posterior_epred(
  probit_lm,
  newdata = fake_data %>% mutate(t = 1, y = ordered(y))
)

pop0_probit <- apply(ep0, c(1, 3), mean)   # draws x categories
pop1_probit <- apply(ep1, c(1, 3), mean)

tau_probit <- as_tibble(
  pop1_probit - pop0_probit,
  .name_repair = ~ paste0("V", seq_along(.x))
) %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "y", values_to = "tau") %>%
  mutate(y = as.integer(sub("V", "", y)))
```

### Comparison of Average Treatment Effects

```{r}
#| label: fig-ate-comparison
#| fig-cap: >-
#|   Comparison of average treatment effect estimates across models against
#|   the ground truth.
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

summ <- function(df, model) {
  df %>% group_by(y) %>%
    summarise(val  = median(tau),
              low  = quantile(tau, 0.025),
              high = quantile(tau, 0.975), .groups = "drop") %>%
    mutate(Model = model)
}

bind_rows(summ(tau_phobart, "PHOBART"), summ(tau_probit, "Linear Probit")) %>%
  ggplot(aes(factor(y), val, color = Model)) +
  geom_linerange(aes(ymin = low, ymax = high), linewidth = 1,
                 position = position_dodge(0.5)) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_point(data = truth_ate, aes(factor(y), tau, color = "Truth"),
             size = 4, shape = 4, stroke = 1.5, inherit.aes = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(
    values = c(
      "PHOBART"       = "#0072B2",
      "Linear Probit" = "#D55E00",
      "Truth"         = "#000000"
    )
  ) +
  labs(
    title = "Effect of the feature on each rating level",
    subtitle = paste0(
      "The linear model averages away the sign-flipping heterogeneity"
    ),
    x = "Rating (1–5)",
    y = "Change in probability"
  ) +
  theme_bw() + theme(legend.position = "bottom")
```

- **Rating 1 (1 Star):** The linear probit estimates a small decrease in 1-star
  ratings. The true latent shift averages to $0.5(1.5) + 0.5(-1.0) = 0.25$, so
  the linear model estimates a slightly positive treatment coefficient
  ($\hat{\beta}_t \approx 0.24$), forcing the modeled probability of the lowest
  category to fall. In reality, the harmed subgroup ($x_1 \le 0$) is pushed
  below the $-1.5$ threshold, causing 1-star reviews to *increase*. The linear
  model commits a dangerous **sign error** on the company's most dissatisfied
  customers.
- **Ratings 2, 3, and 4:** Under the true data-generating process, probability
  mass drains out of the middle ratings as customers polarize toward the
  extremes. The linear probit predicts negligible changes across these tiers,
  missing the hollowed-out middle.
- **Rating 5 (5 Stars):** Both models detect an increase in 5-star ratings, but
  PHOBART accurately quantifies the large positive surge driven by delighted
  power users ($x_1 > 0$).
- **Overall:** PHOBART flexibly recovers the full U-shaped polarization across
  all five categories, tracking the ground truth closely.

## Why These Results Matter

In this data-generating process, the treatment effect is heterogeneous: it is
positive for individuals with $x_1 > 0$ and negative for those with
$x_1 \le 0$. A linear ordinal probit model is forced to represent this
structure using a single treatment coefficient, which leads to biased
population-level estimates when effects cancel out across subgroups.

The linear model's failure here is not subtle: it tells leadership that angry
reviews are dropping when they are actually spiking. The PHOBART model correctly
identifies both the non-linear baseline relationship between $x_1$ and the
outcome and the interaction between $x_1$ and treatment. By combining the
flexibility of BART with the computational advantages of the cloglog link, it
accurately recovers the true average treatment effects across all ordinal
categories. This is the practical payoff: when an intervention helps some
customers and hurts others, only the flexible model tells you so.

### Decision-Making Under Polarization: What to Do Monday Morning

The core business dynamic demonstrated here is **customer polarization**: the
feature simultaneously manufactures both promoters (4–5 stars) and detractors
(1 star). In customer experience management and Net Promoter Score (NPS)
frameworks, a release that polarizes the user base can leave headline summary
metrics—such as mean star ratings or net scores—looking modestly positive or
unmoved, while masking severe customer churn among vulnerable cohorts.

Because Bayesian models produce posterior draws across the full probability
distribution, we can directly query the posterior for decision-relevant
probabilities:

```{r decision-probabilities, message=FALSE, warning=FALSE}
# Compute promoter and detractor comparisons
prob_prom_probit <- mean(
  (pop1_probit[, 4] + pop1_probit[, 5]) > (pop0_probit[, 4] + pop0_probit[, 5])
)
prob_detr_probit <- mean(pop1_probit[, 1] > pop0_probit[, 1])
prob_pol_probit  <- mean(
  ((pop1_probit[, 4] + pop1_probit[, 5]) >
   (pop0_probit[, 4] + pop0_probit[, 5])) &
  (pop1_probit[, 1] > pop0_probit[, 1])
)

prob_prom_phobart <- mean(
  (pop1_phobart[, 4] + pop1_phobart[, 5]) >
  (pop0_phobart[, 4] + pop0_phobart[, 5])
)
prob_detr_phobart <- mean(pop1_phobart[, 1] > pop0_phobart[, 1])
prob_pol_phobart  <- mean(
  ((pop1_phobart[, 4] + pop1_phobart[, 5]) >
   (pop0_phobart[, 4] + pop0_phobart[, 5])) &
  (pop1_phobart[, 1] > pop0_phobart[, 1])
)

decision_summary <- tibble(
  `Business Question` = c(
    "Pr(Share of 4–5 star ratings [Promoters] increased)",
    "Pr(Share of 1-star ratings [Detractors] increased)",
    "Pr(Polarization: Both 1-star and 4–5 star ratings increased)"
  ),
  `Linear Probit` = sprintf(
    "%.1f%%",
    100 * c(prob_prom_probit, prob_detr_probit, prob_pol_probit)
  ),
  `PHOBART` = sprintf(
    "%.1f%%",
    100 * c(prob_prom_phobart, prob_detr_phobart, prob_pol_phobart)
  )
)

kable(
  decision_summary,
  col.names = c("Decision-Relevant Question", "Linear Probit", "PHOBART")
)
```

Looking at these results on Monday morning completely alters the executive
decision:

- **The Linear Model's Blind Spot:** The linear probit reports an 85.2%
  posterior probability that promoters increased, but only a 14.8% probability
  that detractors increased, and **0.0% probability** of customer polarization
  (both extremes increasing simultaneously). Because standard cumulative-link
  models impose monotonic global shifts from a single parameter, they are
  mathematically incapable of detecting simultaneous increases at both
  extremes. Relying on this model, leadership would greenlight a full 100%
  rollout, unaware that they are actively alienating newer users.
- **The Nonparametric Prescription:** PHOBART reveals near certainty that
  top ratings increased *and* near certainty that 1-star ratings spiked. Rather
  than a simplistic "ship or kill" verdict, this enables targeted action:
  1. **Cohort-Gated Rollout:** Ship the feature immediately to power users and
     highly engaged accounts ($x_1 > 0$) who experience clear value.
  2. **Product Redesign for Onboarding:** Investigate why less-engaged
     customers ($x_1 \le 0$) struggle with the update, iterating on UI
     guidance and onboarding workflows before widening exposure.

## Conclusion

This example illustrates the practical value of a unified Bayesian
nonparametric framework: it is easier to implement because it avoids cutpoint
constraints, and more accurate because it adapts automatically to complex
structure in the data.

The usual cautions apply. The causal reading rests on the standard
assumptions, unconfoundedness and adequate overlap between treated and control
units, which no model supplies on its own; when treatment is not randomized,
including a propensity score as an additional covariate can mitigate
regularization-induced confounding (@hahn2020bayesian), though causal
identification ultimately depends on unconfoundedness and adequate overlap.
The proportional-hazards version assumes the covariate effect does not interact
with the rating level; the non-proportional variant (NPHOBART) relaxes this at
some computational cost. @alam2025unified develop PHOBART and NPHOBART for
ordinal regression and note that sharing forests, in the spirit of
@hahn2020bayesian, is a natural route to richer causal estimation.

::: {.callout-tip}
## Learn more
For the theoretical framework, minimax posterior contraction proofs, and
extensions to non-proportional hazards ordinal regression (NPHOBART), see
@alam2025unified, *A unified Bayesian nonparametric framework for ordinal,
survival, and density regression using the complementary log-log link*
([arXiv:2502.00606](https://arxiv.org/abs/2502.00606)).
:::

## Appendix: Session Information

```{r session-info}
sessionInfo()
```
